---
title: "Lake_areal_calcs"
author: "Lauren Knose"
date: "2024-02-29"
output: html_document
editor_options: 
  chunk_output_type: console
---

The purpose of this program is to calculate the lake surface area and area of stratified and unstratified zones. 

# Step 1. Load dependent packages and data:

a) loading dependent packages...

```{r}
#install.packages('utils') #run if first time package used
library(utils) #needed to download and unzip files
#install.packages('sf') #run if first time package used
library(sf) #needed for working with geospatial vectors
#install.packages('tigris') #run if first time package used
library(tigris) #needed to download TIGER/line shapefiles
#install.packages('dplyr') #run if first time package used
library(dplyr) #needed for reformatting data
#install.packages('ggplot2') #run if first time package used
library(ggplot2) #needed for graphing data
#install.packages('leaflet') #run if first time package used
library(leaflet) #needed to make interactive maps
```

Packages loaded.

b) loading dependent data...

```{r}
url<- "https://gis.ohiodnr.gov/geodata/Clermont/eastfork_lake.zip"
download.file(url, file=) #link to download zip file with bathymetry data
unzip(temp) #unzip files
Lake_bath<- st_read("eastfork_10ftcontours.shp") #read
st_crs(Lake_bath) #prints out the coordinate system
unlink(temp) #remove temp file
```

# Step 2. Map the bathymetry data:

a) if bathymetry data is raster (e.g., GeoTIFF) file...



b) if bathymetry data is shapefile...

```{r}


```

# Step 3. Set the epilimnion lower bound and calculate total area of stratified and unstratified zones. 

a) setting the epilimnion lower bound...

```{r}
m_ft<- 1/0.305 #conversion factor from meters to feet
epi_lower<- 2*m_ft #epilimnion lower bound in feet
```

Epilimnion lower bound set.

b) separating lake area by above and below epi lower bound...

```{r}


```

c) 