---
title: "Mendota_SRPexternal_inflow.Rmd"
author: "Lauren Knose"
date: "2023-07-26"
output: html_document
---

The purpose of this program is to calculate the external soluble reactive P (SRP)
loading to Lake Mendota, from the three main inflows of Pheasant Branch (PB), 
Yahara River (YR), and Six-Mile Creek (SMC). 
SRP_inflow = SRP concentration * flow rate

# Step 1. Load the data and packages needed:

a) Loading the packages needed...

```{r}
library(dplyr) #needed for transforming data
library(lubridate) #needed for date conversion functions
library(ggplot2) #needed for plotting
```

Packages loaded.


b) Loading the data needed...

```{r}
PB_SRP_flow<- read.csv(file="Cleaned_data/Inflow_PheasantBranch_SRP.csv")
YR_SRP_flow<- read.csv(file="Cleaned_data/YaharaRiver_DissP_combined.csv")
SMC_SRP_flow<- read.csv(file="Cleaned_data/Inflow_SixMileCreek_SRP.csv")
```

Soluble Reactive P (SRP) data loaded.

# Step 2. Transform and merge the data:

a) transforming the data...

```{r}
lb_kg= 0.453592/1 #conversion factor from lb to kg (1lb = 0.453592 kg)
cf_L= 1/0.0353147 #conversion factor from ft^3 to L (1 ft^2 = 28.3168 L)
sec_day= 86400/1 #conversion factor from sec to day (86400 sec = 1 day)
mg_kg= 1/1000000 #conversion factor from mg to kg (1,000,000 mg = 1 kg)
```

Conversion coefficients defined.

```{r}
PB_SRP_flow<- PB_SRP_flow %>%
  mutate(sampledate=as.Date(sampledate, origin="1899-12-30", format="%m/%d/%Y"),
         Q_cfs=Qout_cfs, #rename column for consistency across data frames
         Qout_cfs=NULL, #remove excess column
         SRP_kgday=(Q_cfs*SRP_mgL*cf_L*sec_day*mg_kg)
         ) %>%
  filter(Q_cfs>=0) #remove any records where there is a negative flow rate
YR_SRP_flow<- YR_SRP_flow %>%
  mutate(sampledate=as.Date(sampledate, origin="1899-12-30", format="%m/%d/%Y"),
         SRP_kgday=(Q_cfs*SRP_mgL*cf_L*sec_day*mg_kg)
         )%>%
  filter(Q_cfs>=0) #remove any records where there is a negative flow rate
SMC_SRP_flow<- SMC_SRP_flow %>%
  mutate(sampledate=as.Date(sampledate, origin="1899-12-30", format="%m/%d/%Y"),
         SRP_kgday=(Q_cfs*SRP_mgL*cf_L*sec_day*mg_kg)
         )%>%
  filter(Q_cfs>=0) #remove any records where there is a negative flow rate
```

SRP data transformed and filtered.

```{r}
PB_SRP_flow <- PB_SRP_flow %>%
  mutate(Inflow="Pheasant Branch") #add column with name of inflow
YR_SRP_flow <- YR_SRP_flow %>%
  mutate(Inflow="Yahara River") #add column with name of inflow
SMC_SRP_flow <- SMC_SRP_flow %>%
  mutate(Inflow="Six-mile Creek") #add column with name of inflow

Inflows_SRP<- merge(PB_SRP_flow, YR_SRP_flow, #merge two data frames
                all=TRUE) #keep all records for both data frames
Inflows_SRP<- merge(Inflows_SRP, SMC_SRP_flow, #merge two data frames
                all=TRUE) #keep all records for both data frames                
Inflows_SRP<- Inflows_SRP %>%
  mutate(Site_ID=NULL, #remove excess column
         Source_ID=NULL, #remove excess column
         daynum=as.integer(format(sampledate, format="%j")), #add column with 
         #daynum for filtering
         Yr=as.integer(format(sampledate, format="%Y")))
```

SRP data files merged into one file, "Inflows_SRP."

```{r}
layout(matrix(c(1,2,3),3,1)) # optional 3 graphs/page
plot(PB_SRP_flow$sampledate, PB_SRP_flow$SRP_kgday, 
     xlab = "Sample Date", ylab = "SRP Loading (kg/day)", 
     main = "Pheasant Branch", cex.lab=1.5, cex.main=2, cex.axis=2)
plot(YR_SRP_flow$sampledate, YR_SRP_flow$SRP_kgday, 
     xlab = "Sample Date", ylab = "SRP Loading (kg/day)", 
     main = "Yahara River", cex.lab=1.5, cex.main=2, cex.axis=2)
plot(SMC_SRP_flow$sampledate, SMC_SRP_flow$SRP_kgday, 
     xlab = "Sample Date", ylab = "SRP Loading (kg/day)", 
     main = "Six-Mile Creek", cex.lab=1.5, cex.main=2, cex.axis=2)
par(mfrow=c(1,1))#returns graphing viewer to 1 plot per page

```

SRP data plotted. SRP available for 1995 to 2010 in PB, 2004 to 2010 in YR, and
2013 to 2022 in SMC. Data does not overlap for all three inflows.

```{r}
ggplot(data=Inflows_SRP, 
       aes(x=daynum, 
           y=SRP_kgday)) +
  geom_line()+
  facet_wrap(~Inflow) +
  theme_classic()+
  labs(y="Daily Soluble Reactive P loading (kg/day)",
       x="Day of Year",
      caption="White highlight indicates CyanoHAB season, 
      mid-July to mid-Sept.") +
  ylim(c(0, max(Inflows$TotP_lbday))) +
  annotate('rect', fill='gray60', alpha=.2, #creates a shadow box 
           xmin=-Inf, xmax=195, ymin=0, ymax=Inf) +#shadow box highlights 
#CyanoHAB season
  annotate('rect', fill='gray60', alpha=.2, #creates a shadow box 
           xmin=255, xmax=Inf, ymin=0, ymax=Inf)  #shadow box highlights 
#CyanoHAB season
```

Graph plotted of SRP for day of year across the three inflows, for all years.
