---
title: "Mendota_Pexternal_inflow"
author: "Lauren A. Knose, ORISE-EPA"
date: '2023-03-06'
output:
  html_document: default
  pdf_document: default
---

The purpose of this program is to process and calculate the external P loading
to Lake Mendota, from the three main inflows, Pheasant Branch (1), Yahara River 
(2), and Six-Mile Creek (3). 


# Step 1. Load the data and packages needed:

a) Loading the packages needed...

```{r}
library(dplyr) #needed for transforming data
library(lubridate) #needed for date conversion functions
library(ggplot2) #needed for plotting
```

Packages loaded.


b) Loading the data needed...

```{r}
PB_TotP_flow<- read.csv(file="Cleaned_data/Inflow_PheasantBranch_TotP.csv")
PB_TotP_flow<- PB_TotP_flow %>%
  mutate(sampledate=as.Date(sampledate, origin="1899-12-30", format="%m/%d/%Y"),
         Q_cfs=Qout_cfs, #make sure all column names same across data frames
         Qout_cfs=NULL)
YR_TotP_flow<- read.csv(file="Cleaned_data/YaharaRiver_TotP_combined.csv")
YR_TotP_flow<- YR_TotP_flow %>%
  mutate(sampledate=as.Date(sampledate, origin="1899-12-30", format="%m/%d/%Y"))
SMC_TotP_flow<- read.csv(file="Cleaned_data/Inflow_SixMileCreek_TotP.csv")
SMC_TotP_flow<- SMC_TotP_flow %>%
  mutate(sampledate=as.Date(sampledate, origin="1899-12-30", format="%m/%d/%Y"))

PB_DisP_flow<- read.csv(file="Cleaned_data/Inflow_PheasantBranch_SRP.csv")
PB_DisP_flow<- PB_DisP_flow %>%
  mutate(sampledate=as.Date(sampledate, origin="1899-12-30", format="%m/%d/%Y"))
YR_DisP_flow<- read.csv(file="Cleaned_data/YaharaRiver_DissP_combined.csv")
YR_DisP_flow<- YR_DisP_flow %>%
  mutate(sampledate=as.Date(sampledate, origin="1899-12-30", format="%m/%d/%Y"))
SMC_DisP_flow<- read.csv(file="Cleaned_data/Inflow_SixMileCreek_SRP.csv")
SMC_DisP_flow<- SMC_DisP_flow %>%
  mutate(sampledate=as.Date(sampledate, origin="1899-12-30", format="%m/%d/%Y"))
```

Data loaded.

The site ID and source for Pheasant Branch is `r unique(PB_TotP_flow$Site_ID)`.
The site ID and source for Yahara River is `r unique(YR_TotP_flow$Site_ID)`.
The site ID and source for Six-mile Creek is `r unique(SMC_TotP_flow$Site_ID)`.
Note, there are two sites, one with historic data and one with current data, for
Six-mile Creek.

c) Merge the three data frames:

```{r}
PB_TotP_flow <- PB_TotP_flow %>%
  mutate(Inflow="Pheasant Branch") #add column with name of inflow
YR_TotP_flow <- YR_TotP_flow %>%
  mutate(Inflow="Yahara River") #add column with name of inflow
SMC_TotP_flow <- SMC_TotP_flow %>%
  mutate(Inflow="Six-mile Creek") #add column with name of inflow

Inflows<- merge(PB_TotP_flow, YR_TotP_flow, #merge two data frames
                all=TRUE) #keep all records for both data frames
Inflows<- merge(Inflows, SMC_TotP_flow, #merge two data frames
                all=TRUE) #keep all records for both data frames                
```

The three data frames are merged.

# Step 2. View and summarize the data:

Graphing total P external loading for each inflow...

```{r}
layout(matrix(c(1,2,3),3,1)) # optional 3 graphs/page
plot(PB_TotP_flow$sampledate, PB_TotP_flow$TotP_lbday, 
     xlab = "Sample Date", ylab = "Total P Loading (lb/day)", main = "Pheasant Branch")
plot(YR_TotP_flow$sampledate, YR_TotP_flow$TotP_lbday, 
     xlab = "Sample Date", ylab = "Total P Loading (lb/day)", main = "Yahara River")
plot(SMC_TotP_flow$sampledate, SMC_TotP_flow$TotP_lbday, 
     xlab = "Sample Date", ylab = "Total P Loading (lb/day)", main = "Six-Mile Creek")
par(mfrow=c(1,1))#returns graphing viewer to 1 plot per page
```

Graphing average total P external loading by daynum for each inflow between 2010 and 2020...
```{r}
lb_kg = 1/2.20462 #conversion factor from pound to kg
Inflows_totP <- Inflows %>%
  mutate(Yr=as.integer(format(sampledate, "%Y")), #add column with year
         daynum=as.integer(format(sampledate, "%j"))) %>% #add column with day number
  filter(Yr >= 2012 & Yr <=2022) %>%  #filter data to include only the past decade
  group_by(daynum, Inflow) %>% #for each inflow on each day
  summarize(TotP_kgday_avg=mean(TotP_lbday*lb_kg, #calculate mean total P (in kg/day)
                                na.rm=TRUE)) #average the total P (lb/day) across the years

ggplot(data=Inflows_totP, 
       aes(x=daynum, 
           y=TotP_kgday_avg)) +
  geom_line()+
  facet_wrap(~Inflow) +
  theme_classic()+
  labs(y="Average Daily Total P loading from streams (kg/day), 2012 to 2022",
       x="Day of Year",
      caption="Grey shading indicates CyanoHAB season, mid-July to mid-Sept.") +
  ylim(c(0, max(Inflows$TotP_lbday))) +
  annotate('rect', fill='gray60', alpha=.2, #creates a shadow box 
           xmin=-Inf, xmax=195, ymin=0, ymax=Inf) +#shadow box highlights CyanoHAB season
  annotate('rect', fill='gray60', alpha=.2, #creates a shadow box 
           xmin=255, xmax=Inf, ymin=0, ymax=Inf)  #shadow box highlights CyanoHAB season
```

Daily average total P loading from each of the inflows plotted.

```{r}
Inflows_totP %>%
  group_by(Inflow) %>%
  summarise(mean=mean(TotP_kgday_avg, na.rm=TRUE),
            min=min(TotP_kgday_avg, na.rm=TRUE),
            max=max(TotP_kgday_avg, na.rm=TRUE))
```

# Step 3. Combine the data into total external load from all inflows:

```{r}
Inflows_sumP<- Inflows %>%
  group_by(sampledate) %>% #for each sample date...
  summarize(TotP_kgday_sum=sum(TotP_lbday*lb_kg)) %>% #sum the total P loads for all three inflows
  filter(!is.na(TotP_kgday_sum)) %>% #remove NAs 
  filter(TotP_kgday_sum>=0) #remove any negative values
```

# Step 4. Save new data file.

```{R}
write.csv(Inflows_sumP, file="Cleaned_data/Mendota_P_inflows.csv") #save as new data file
```

Plot total P load by year...

```{r}
Inflows_sumP<- read.csv(file="Cleaned_data/Mendota_P_inflows.csv")
Inflows_sumP <- Inflows_sumP %>%
  mutate(Yr=year(sampledate),#create column with year
         sampledate=NULL) %>% #remove extra column
  filter(Yr >= 2012 & Yr <=2022) %>% #filter data to include only the past decade
  mutate(X=NULL) #remove index column

Inflows_sumP_yr<- Inflows_sumP %>%
  group_by(Yr) %>% #group all records by year
  summarise(sum_TP_kg=sum(TotP_kgday_sum)) #calculate the total P load 

ggplot(Inflows_sumP_yr, aes(x=as.factor(Yr), y=sum_TP_kg)) +
  geom_col() +
  theme_classic() +
  labs(x="Year", y="Total P loading from three inflows (kg)") 
```

Total P loading (kg) from the three inflows  plotted.

The average annual total P load, from 2012 to 2022, was 
`r mean(Inflows_sumP_yr$sum_TP_kg)` +- `r sd(Inflows_sumP_yr$sum_TP_kg)/sqrt(length(Inflows_sumP_yr$Yr))` kg/yr. 


