---
title: "Mendota_Algae"
author: "Lauren A. Knose, ORISE-EPA"
date: '2023-02-15'
output: html_document
---

The following text and code documents how algal data was processed and analyzed.
Data was also converted to time series.

# Step 1. Load dependent packages and data:

a) loading the dependent packages...
```{r}
library(ggplot2) #needed for plots
library(ggbreak) #needed for plot scale breaks
library(dplyr) #needed for reshaping/reformaing data
library(lubridate) #needed for reformatting data and using "year" function
library(DiagrammeR)
library(plotly) #needed for making 3D plots
```

Packages loaded.

b) loading the chlorophyll-a data...
```{r}
Chl<- read.csv(file="Original_data/ntl38_v5_chl.csv", header=TRUE) #has chl data
```

Chlorophyll-a data loaded. 

c) loading the phytoplankton data...
```{r}
phyto<-read.csv(file="Original_data/ntl88_v13_phytoplankton.csv", header=TRUE) #has the phytoplankton community dataset
```

Phytoplankton community data loaded. 


# Step 2. Filter and transform data:

a) filter and transform data for total algal biomass (as Chl-a) by sample date...

```{r}
Me_Chl<- Chl %>%
  mutate(sampledate=as.Date(sampledate, origin="1899-12-30"), #tells R to format field as Date
         Chla_ugL= tri_chl_spec, #renames field to common name with units
         Yr=year4) %>% #renames field to common name
  filter(lakeid=="ME", #filter for Lake Mendota data
         depth_range_m=="0-8", #filter for data from integrated samples 0 to 8m
         rep=="1") %>% #remove QA analytic replicates
  filter(!is.na(Chla_ugL)) %>% #removes NAs in chl-a data
  select(sampledate, Yr, Chla_ugL) #selects only fields of interest
```

Chl-a data filtered and transformed. 

Plotting Chl-a data available...

```{r}
plot(Me_Chl$sampledate, Me_Chl$Chla_ugL,
     xlab="Date",
     ylab="Total Chl-a (ugL)") #plots the new time series
```

Chl-a data plotted. 

b) filter and transform data for cyanobacteria relative biovolume (CRB) by sample date...

```{r}
Me_cyanoRB<- phyto %>%
  mutate(sampledate=as.Date(sampledate, origin="1899-12-30"), #tells R to format field as Date
         Yr=year4) %>% #renames field to coomon name
  filter(lakeid=="ME", #filter for Lake Mendota data
         depth_range=="0-8m", #filter for data from integrated samples 0 to 8m
         division=="Cyanophyta") %>% #filter for cyanobacteria data
  group_by(sampledate) %>% #group all observations by sample date
  summarize(relbiovol=sum(relative_total_biovolume), #add all observations of cyanobacteria biovolumes together for each sample date
            cellsmL=sum(cells_per_ml)) %>% #adds all the cyanobacteria cells/mL together for each sample date
  filter(relbiovol<=100) #remove any invalid records (>100%  not valid entry)          
```

CRB data filtered and transformed.

Plotting CRB data available...

```{r}
par(mfrow=c(1,2)) #returns graphing viewer to 2 plots per page
plot(Me_cyanoRB$sampledate, Me_cyanoRB$relbiovol, #point plot x, y
     xlab="Date",
     ylab="Cyanobacteria Relative Biovolume (as % total algal biovolume)") 
plot(Me_cyanoRB$sampledate, Me_cyanoRB$cellsmL, #point plot x, y
     xlab="Date",
     ylab="Cyanobacteria Density (cells/mL)") #plots the new time series
par(mfrow=c(1,1)) #returns graphing viewer to 1 plot per page
```

CRB data plotted. 

# Step 3. Interpolate missing data values for equal time intervals between observations:

a) interpolating Chl-a data...

```{r}
date_range<- seq( #creates a vector of sequence data
  min(Me_Chl$sampledate), #starting at the first sample date
  max(Me_Chl$sampledate), #ending at the last sample date
  by="1 day") #observations every day  
interpolated<- approx( #creates a vector of data that linearly interpolates 
  Me_Chl$sampledate, #across date
  Me_Chl$Chla_ugL, #the predicted values of Chl-a
  xout=date_range) #for every 1 day
Chla_interp <- data.frame( #creates a new data frame that has 
  sampledate = as.Date(interpolated$x), #renames the new vector as common field 
  Chla_ugL=interpolated$y) #renames the new vector as common field with units
```

Chl-a data interpolated.

Plotting new Chl-a time series...

```{r}
plot(Chla_interp$sampledate, Chla_interp$Chla_ugL,
     xlab="Date",
     ylab="Total Chl-a (ugL)") #plots the new time series
```

New Chl-a time series plotted. 

b) interpolating CRB data...

```{r}
date_range<- seq( #creates a vector of sequence data
  min(Me_cyanoRB$sampledate), #starting at the first sample date
  max(Me_cyanoRB$sampledate), #ending at the last sample date
  by="1 day") #observations every day  
interpolated<- approx( #creates a vector of data that linearly interpolates 
  Me_cyanoRB$sampledate, #across date
  Me_cyanoRB$relbiovol, #the predicted values of CRB
  xout=date_range) #for every 1 day
CRB_interp <- data.frame( #creates a new data frame that has 
  sampledate = as.Date(interpolated$x), #renames the new vector as common field 
  Cyanorelbiovol=interpolated$y) #renames the new vector as common field with units
```

CRB interpolated.

Plotting new CRB time series...

```{r}
plot(CRB_interp$sampledate, CRB_interp$Cyanorelbiovol,
     xlab="Date",
     ylab="Cyanobacteria Relative Biovolume (as % total algal biovolume)") #plots the new time series
```

New CRB time series plotted. 

c) interpolating Cyanobacteria Density (CD) data...

```{r}
date_range<- seq( #creates a vector of sequence data
  min(Me_cyanoRB$sampledate), #starting at the first sample date
  max(Me_cyanoRB$sampledate), #ending at the last sample date
  by="1 day") #observations every day  
interpolated<- approx( #creates a vector of data that linearly interpolates 
  Me_cyanoRB$sampledate, #across date
  Me_cyanoRB$cellsmL, #the predicted values of CD
  xout=date_range) #for every 1 day
CD_interp <- data.frame( #creates a new data frame that has 
  sampledate = as.Date(interpolated$x), #renames the new vector as common field 
  CyanoDens_cellsmL=interpolated$y) #renames the new vector as common field with units
```

CyanoDens_cellsmL interpolated.

Plotting new CD time series...

```{r}
plot(CD_interp$sampledate, CD_interp$CyanoDens_cellsmL,
     xlab="Date",
     ylab="Cyanobacteria Density (cells/mL)") #plots the new time series
```

New CD time series plotted. 

# Step 4.  Merge the three time series:
Note, you can only merge 2 time series at a time.

```{r}
Me_Algae<- merge(CD_interp, Chla_interp, by="sampledate") #merge CD and Chl-a data time series first
Me_Algae<- merge(Me_Algae, CRB_interp, by = "sampledate") #merge combined time series with CRB data second
```

Time series merged.

# Step 5. Summarize and view the data:

a) adding fields needed for plotting...

```{r}
Me_Algae<- Me_Algae %>%
  mutate(Yr=as.integer(format(sampledate, format="%Y")), #adds field with year
         daynum=as.integer(format(sampledate, format="%j"))) #adds field with Julian day number
```

b) Plot Chl-a by day of year...
```{r}
ggplot(data=subset(Me_Algae, Yr<2002 | Yr>2008  & !Yr==2017), #remove the data gap from 2002 to 2008 and in 2017
       aes(x=daynum,y=Chla_ugL)) + 
  geom_point(aes(col=Yr))+ #colors points by year 
  labs(title="Lake Mendota, WI", #add title
       x="Day of Year", #add x axis label
       y="Extracted Chl-a (ug/L)", #add y axis label
       caption="White shading represents CyanoHAB season, mid-Jul to mid-Sep. n/
       Black solid line represents smoothed loess curve") + #add caption for notation
  ylim(0, max(Me_Algae$Chla_ugL)) + #limit y-axis
  scale_x_continuous(breaks = seq(0, max(Me_Algae$daynum), #places axis tick marks
                                  by=30)) + #every 30 days (1 month)
  theme_classic()+
  scale_color_continuous(name="Year")+ #adds color scale for year
  geom_smooth(colour="black") + #adds loess curve (moving average regression model)
  annotate('rect', fill='gray60', alpha=.2, #creates a shadow box 
           xmin=-Inf, xmax=195, ymin=0, ymax=Inf) +#shadow box days outside of CyanoHAB season
  annotate('rect', fill='gray60', alpha=.2, #creates a shadow box 
           xmin=255, xmax=Inf, ymin=0, ymax=Inf)  #shadow box days outside of CyanoHAB season
```

c) Plot Chl-a each year, with coding for CyanoHAB severity:

```{r}
ggplot(data=subset(Me_Algae, Yr<2002 | Yr>2008  & !Yr==2017), #remove the data gap from 2002 to 2008 and in 2017
       aes(x=daynum, y=Chla_ugL)) + 
  geom_point() + #point plot
  geom_line() + #add line between data points
  scale_y_continuous(trans="log10") + # log-transforms y axis
  facet_wrap(~Yr) + #separates out plots by year
  labs(title="Lake Mendota, WI", #adds title
       x="Day of Year",  #adds x-axis label
       y="Total Algal Biomass as Chl-a (ug/L)", #adds y-axis label
       caption="White shading indicates CyanoHAB season, mid-July to mid-Sept.") + #adds caption for notation
  theme_classic()+
  annotate('rect', fill='gray60', alpha=.2, #creates a shadow box 
           xmin=-Inf, xmax=195, ymin=0, ymax=Inf) +#shadow box highlights CyanoHAB season
  annotate('rect', fill='gray60', alpha=.2, #creates a shadow box 
           xmin=255, xmax=Inf, ymin=0, ymax=Inf) +#shadow box highlights CyanoHAB season
  annotate('rect', xmin=0, xmax=Inf, ymin=0, ymax=10, alpha=.2, fill='green') + #low/no CyanoHAB
  annotate('rect', xmin=0, xmax=Inf, ymin=10, ymax=20, alpha=.2, fill='orange')+ #moderate CyanoHAB
  annotate('rect', xmin=0, xmax=Inf, ymin=20, ymax=Inf, alpha=.2, fill='red') #severe CyanoHAB
```

d) Plot CRB by day of year...

```{r}
ggplot(data=Me_Algae, aes(x=daynum, y=Cyanorelbiovol)) + 
  geom_point(aes(col=Yr))+ #colors points by year
  labs(title = "Lake Mendota, WI", #adds title
       x="Day of Year", #adds x-axis label
       y="Cyanobacteria relative biovolume (% of total algal biovolume)", #adds y-xis label
       caption="White shading represents CyanoHAB season, mid-Jul to mid-Sep. n/
       Black solid line represents smoothed loess curve") + #add caption for notation
  scale_x_continuous(breaks = seq(0, max(Me_Algae$daynum), #places axis tick marks
                                  by=30)) + #every 30 days (1 month)
  theme_classic()+
  scale_color_continuous(name="Year")+ #adds color scale for year
  theme_classic()+
  geom_hline(yintercept = 50, lty="dashed", col="red") + #adds horizontal line to indicate 50% of total biovolume
  geom_smooth(colour="black") + #adds loess curve (moving average regression model)
  annotate('rect', fill='gray60', alpha=.2, #creates a shadow box 
           xmin=-Inf, xmax=195, ymin=0, ymax=Inf) +#shadow box highlights CyanoHAB season
  annotate('rect', fill='gray60', alpha=.2, #creates a shadow box 
           xmin=255, xmax=Inf, ymin=0, ymax=Inf) #shadow box highlights CyanoHAB season
```

e) Plot CD by day of year...

```{r}
ggplot(data=Me_Algae, aes(x=daynum, y=CyanoDens_cellsmL)) + 
  geom_point(aes(col=Yr))+ #colors points by year
  labs(title = "Lake Mendota, WI", #adds title
       x="Day of Year", #adds x-axis label
       y="Cyanobacteria Density (cells/mL)", #adds y-xis label
       caption="White shading represents CyanoHAB season, mid-Jul to mid-Sep. n/
       Black solid line represents smoothed loess curve") + #add caption for notation
  scale_x_continuous(breaks = seq(0, max(Me_Algae$daynum), #places axis tick marks
                                  by=30)) + #every 30 days (1 month)
  theme_classic()+
  scale_color_continuous(name="Year")+ #adds color scale for year
  theme_classic()+
  geom_hline(yintercept = 50, lty="dashed", col="red") + #adds horizontal line to indicate 50% of total biovolume
  geom_smooth(colour="black") + #adds loess curve (moving average regression model)
  annotate('rect', fill='gray60', alpha=.2, #creates a shadow box 
           xmin=-Inf, xmax=195, ymin=0, ymax=Inf) +#shadow box highlights CyanoHAB season
  annotate('rect', fill='gray60', alpha=.2, #creates a shadow box 
           xmin=255, xmax=Inf, ymin=0, ymax=Inf) #shadow box highlights CyanoHAB season
```

# Step 6. Save as new data file:

```{r}
write.csv(Me_Algae, file="Cleaned_data/Mendota_Algae.csv")
```