---
title: "0Dependent_packages"
author: "Lauren Knose"
date: "2024-11-26"
output: html_document
---

The purpose of this program is to install dependent packages used in the sequential programs to the user's R version. The beginning of each 

The list of dependent packages are as follows: 
- dplyr 
- reshape2
- data.table
- tidyverse
- zoo
- car
- ggplot2 
- ggpubr
- ggpattern
- plotly
- remotes 
- rloadest
- rLakeAnalyzer
- forecast

The following code will install the dependent packages to user's machine from the CRAN.

```{r}
install.packages(c("dplyr", "reshape2", "data.table", "tidyverse", "zoo","car", "ggplot2", "ggpubr", "ggpattern", "plotly", "rLakeAnalyzer", "forecast"))
```

Unless errors reported, all packages are installed. 

The following code will install the dependent packages to user's machine from the remote repository. 

```{r}
install.packages("remotes")
remotes::install_github("USGS-R/smwrData", force=TRUE) #use if 1st time using rloadest
remotes::install_github("USGS-R/smwrBase", force=TRUE) #use if 1st time using rloadest
remotes::install_github("USGS-R/smwrGraphs", force=TRUE) #use if 1st time using rloadest
remotes::install_github("USGS-R/smwrStats", force=TRUE) #use if 1st time using rloadest
remotes::install_github("USGS-R/smwrQW", force=TRUE) #use if 1st time using rloadest
remotes::install_github("USGS-R/rloadest", force=TRUE) #use if 1st time using rloadest
```

Unless errors reported, all dependent packages are installed.