---
title: "Impact_model.Rmd"
author: "Lauren A. Knose, ORISE-EPA"
date: '2024-06-10'
output: html_document
editor_options: 
  chunk_output_type: console
---

The purpose of this program is to perform an impact analysis comparing the internal vs external P loading on changes in Chl-a and CyanoDensity.


# Step 1. Load dependent packages and data needed:

a) Loading dependent packages...
```{r}
library(dplyr) #needed for reshaping/reformating data
library(reshape2) #needed for reshaping data frames
library(ggplot2) #needed for plotting and saving plots
library(car) #needed to get model VIFs
library(ggpattern) #needed to add patterns to barplots
```

Packages loaded.

b) Loading data...

```{r}
### loading the data ###
weekly<- read.csv(file="Cleaned_data/Impact_model_weekly.csv")
```

Data loaded.

# Step 2. Define predictors (transformations and lags):

a) checking for normality...

```{r}
par(mfrow=c(2,5))
hist(weekly$Chla_ugL, breaks=18,#histograph of Chl-a
     main="non-transformed",
     xlab="Chl-a") 
hist(weekly$CD_cellsmL, breaks=18,#histograph of CD
     main="non-transformed",
     xlab="CD") 
hist(weekly$Pext, breaks=18,#histogrpah of P_ext
     main="non-transformed",
     xlab="P_ext") 
hist(weekly$Pint, breaks=18,#histograph of P_int
     main="non-transformed",
     xlab="P_int") 
hist(weekly$avgTotP_HEratio, breaks=18,#histograph of P_hypo_epi
     main="non-transformed",
     xlab="P_hypo:P_epi") 
hist(log10(weekly$Chla_ugL), breaks=18,#histograph of log10(Chl-a)
     main="log-transformed",
     xlab="Chl-a") 
hist(log10(weekly$CD_cellsmL), breaks=18,#histograph of log10(CD)
     main="log-transformed",
     xlab="CD") 
hist(log10(weekly$Pext), breaks=18,#histograph of log10(P_ext)
     main="log-transformed",
     xlab="P_ext") 
hist(log10(weekly$Pint), breaks=18,#histograph of log10(P_int)
     main="log-transformed",
     xlab="P_int") 
hist(log10(weekly$avgTotP_HEratio), breaks=18,#histograph of log10(P_hypo:epi)
     main="log-transformed",
     xlab="P_hypo:P_epi") 
par(mfrow=c(1,1)) #return to default
#save manually
```

Comparison of non-transformed and transformed data graphed.

b) checking lags...

```{r}
### vary lags ###
test_k<- glm(formula=CD_cellsmL~log10(Pint_k28), data=weekly) #data frame
summary(lm(test_k))
```

Predictor lags defined.

# Step 3. Run regression:

a) running model: CD_t ~ CD_t-k + log10(P_ext,t-k) + log10(P_int, t-k) + error

```{r}
### remove outliers as needed ###
#inputs<- weekly[-c(22, 26),]

### define the model ###
regka_CD<- glm(formula=CD_cellsmL~log10(Pint_k21)+log10(Pext_k28), data=weekly)

### print model AIC ###
summary(regka_CD)

### print model R^2 ###
summary(lm(regka_CD))
```

Model defined.

b) running model diagnostics...

```{r}
### plot expected vs observed ###
plot(x=predict(regka_CD), y=weekly$CD_cellsmL, 
    xlab="Predicted Values", ylab="Observed Values")
abline(a=0,b=1) #add 1:1 line to plot

### print model VIFs ###
vif(regka_CD) #prints the VIFs 

### print model diagnostics ###
png(file="Products/SI_figures/Figure14_color.png", width=750, height=500) #open png file
layout(matrix(c(1,2,3,4),2,2)) #layout 4 graphs/page
SIfig14<-plot(regka_CD) #creates plot
SIfig14
dev.off() #closes the file
par(mfrow=c(1,1))
```

Model diagnostics complete.

# Step 4. Calculate the partial sum squares for each year:

a) calculating partial squares for each year...

```{r}
### define the model ###
YRregk_CD_2013<- aov(formula=CD_cellsmL~log10(Pint_k21)+log10(Pext_k28), data=subset(weekly, Yr==2013)) #subset data by year
            
### print the Sum of Squares as table ###
SS_2013<- summary(YRregk_CD_2013)
impact2013<- data.frame(SS_2013[[1]])[,2,drop=FALSE] #pulls out the SS as table

### define the model ###
YRregk_CD_2014<- aov(formula=CD_cellsmL~log10(Pint_k21)+log10(Pext_k28), data=subset(weekly, Yr==2014)) #subset data by year

### print the Sum of Squares as table ###
SS_2014<- summary(YRregk_CD_2014)#prints the sum of squares for the regression 
impact2014<- data.frame(SS_2014[[1]])[,2,drop=FALSE] #pulls out the SS as table

### define the model ###
YRregk_CD_2015<- aov(formula=CD_cellsmL~log10(Pint_k21)+log10(Pext_k28), data=subset(weekly, Yr==2015)) #subset data by year

### print the Sum of Squares as table ###
SS_2015<- summary(YRregk_CD_2015)#prints the sum of squares for the regression 
impact2015<- data.frame(SS_2015[[1]])[,2,drop=FALSE] #pulls out the SS as table

### define the model ###
YRregk_CD_2016<- aov(formula=CD_cellsmL~log10(Pint_k21)+log10(Pext_k28), data=subset(weekly, Yr==2016)) #subset data by year

### print the Sum of Squares as table ###
SS_2016<- summary(YRregk_CD_2016)#prints the sum of squares for the regression 
impact2016<- data.frame(SS_2016[[1]])[,2,drop=FALSE] #pulls out the SS as table

### define the model ###
YRregk_CD_2017<- aov(formula=CD_cellsmL~log10(Pint_k21)+log10(Pext_k28),  data=subset(weekly, Yr==2017)) #subset data by year

### print the Sum of Squares as table ###
SS_2017<- summary(YRregk_CD_2017)#prints the sum of squares for the regression 
impact2017<- data.frame(SS_2017[[1]])[,2,drop=FALSE] #pulls out the SS as table

### define the model ###
YRregk_CD_2018<- aov(formula=CD_cellsmL~log10(Pint_k21)+log10(Pext_k28), data=subset(weekly, Yr==2018)) #subset data by year

### print the Sum of Squares as table ###
SS_2018<- summary(YRregk_CD_2018)#prints the sum of squares for the regression 
impact2018<- data.frame(SS_2018[[1]])[,2,drop=FALSE] #pulls out the SS as table
```

Partial Sum Squares calculated for each year.

b) merging the Sum Squares tables into one data file...

```{r}
### merging the Sum Squares tables ###
Model_output_CD<- cbind(impact2013, impact2014, impact2015, impact2016, impact2017, impact2018) #combines all tables together
Model_output_CD$Sum_Sq_part<- c("Internal load","External load",  "Residual") #add column with predictor name
colnames(Model_output_CD)<- c("2013", "2014", "2015", "2016", "2017", "2018", "Sum_Sq_part") # renames columns with year for sum squares

### save SS table as new data file ###
write.csv(Model_output_CD, file="Products/Figure3_data.csv")

### print SS table as bar plot ###
CD_table<- melt(Model_output_CD, id.vars="Sum_Sq_part", variable.name = "Year", value.name = "PSS") %>%
  mutate(Sum_Sq_part=factor(Sum_Sq_part, levels=c("Residual", "External load",  "Internal load"), ordered=TRUE))
fig3a<- ggplot(data=CD_table, aes(y=PSS, x=Year, pattern=Sum_Sq_part, fill=Sum_Sq_part)) +
  geom_bar_pattern(position="fill", stat="identity") +
  labs(y="Sum of Squares for Regression", pattern="Predictor",title="a) Baseline") +
  guides(fill="none") +
  theme(legend.position="bottom")+
  scale_pattern_manual(values=c("stripe","wave","circle"))
fig3a
ggsave(fig3a, file="Products/Figure3a_color.png")
```

Yearly source-attributed impact outputted.