---
title: "cyanoHAB_season"
author: "Lauren Knose"
date: "2024-05-23"
output: html_document
editor_options: 
  chunk_output_type: console
---

The purpose of this program is to define the cyanoHAB season of the lake. Two variables are needed for this program, Secchi disk depth and phytoplankton cell counts. 

# Step 1. Load dependent packages and data:

a) loading the dependent packages...

```{r}
library(dplyr) #needed for reshaping/reformating data
library(ggplot2) #needed for plots
library(ggpubr) #needed to plot mupltiple plots together
```

Packages loaded.

b) loading the data...

```{r}
### load the Secchi disk depth data ###
Secchi<- read.csv(file="Original_data/ntl31_v9_secchi.csv")

### load the phytoplankton cell count data ###
phyto<-read.csv(file="Original_data/ntl88_v13_phytoplankton.csv", header=TRUE)
```

Data loaded.

# Step 2. Reformat and filter data:

a) reformatting and filtering data for cyanobacteria relative biovolume (CRB) and cyanobacteria density (CD)...

```{r}
Me_cyanoRB<- phyto %>%
  mutate(sampledate=as.Date(sampledate, origin="1899-12-30")) %>% #tells R to format field as Datey
  filter(lakeid=="ME", #filter for Lake Mendota data
         depth_range=="0-8m", #filter for data from integrated samples 0 to 8m
         division=="Cyanophyta") %>% #filter for cyanobacteria data
  group_by(sampledate) %>% #group all observations by sample date
  summarize(cyano_relbiovol=sum(relative_total_biovolume), #add all observations of cyanobacteria biovolumes together for each sample date
            CD_cellsmL=sum(cells_per_ml)) %>% #adds all the cyanobacteria cells/mL together for each sample date
  filter(cyano_relbiovol<=100) #remove any invalid records (>100%  not valid entry)
```

Cyanobacteria data defined and filtered.

b) reformatting and filtering the Secchi depth data...

```{r}
Me_Secchi<- Secchi %>%
  mutate(sampledate=as.Date(sampledate, origin="1899-12-30"),#tells R to format field as Date
         SD_m=secnview) %>% #adds units to field
  filter(lakeid=="ME") %>% #filter for Lake Mendota data
  select(sampledate, SD_m) #select the fields needed
```

Secchi depth data defined and filtered.

c) merging the data...
```{r}
### merging the two data files ###
SIfig4_data<- merge(Me_cyanoRB, Me_Secchi, by=c("sampledate"),
                       all=TRUE)  #merges data frames and keeps all records 
### reformatting dates ###
SIfig4_data<- SIfig4_data %>%
  mutate(Yr=as.numeric(format(sampledate, "%Y")), #renames field
         daynum=as.numeric(format(sampledate, "%j"))) #adds field with day
```

Data merged.

# Step 3. View the data available:

a)  plotting Secchi depth ...

```{r}
### plot Secchi depth by day of year ###
SIfig4a<- ggplot(data=SIfig4_data, aes(x=daynum, y=SD_m)) + #plots SD across day
  geom_point() + #plots observations as points
  theme_classic(base_size=12) +
  geom_smooth() + #adds regression curve
  labs(x="Day of year", y="Secchi depth (m)", title="a)") +
  annotate('rect', fill='gray60', alpha=.2, #creates a shadow box 
           xmin=-Inf, xmax=196, ymin=0, ymax=Inf) +#shadow box highlights CyanoHAB season
  annotate('rect', fill='gray60', alpha=.2, #creates a shadow box 
           xmin=252, xmax=Inf, ymin=0, ymax=Inf)  #shadow box highlights CyanoHAB season
```

Plot generated.

b) plotting cyanobacteria relative biovolume...

```{r}
### plot cyanobacteria relative biovolume by day of year ###
SIfig4b <- ggplot(data=SIfig4_data, aes(x=daynum, y=cyano_relbiovol)) + #plots cyano RB across day
  geom_point() + #plots observations as points
  theme_classic(base_size=12) +
  geom_smooth() + #adds regression curve
  labs(x="Day of year", 
       y="Cyanobacteria biovolume (% total)", title="b)") +
  annotate('rect', fill='gray60', alpha=.2, #creates a shadow box 
           xmin=-Inf, xmax=196, ymin=0, ymax=Inf) +#shadow box highlights CyanoHAB season
  annotate('rect', fill='gray60', alpha=.2, #creates a shadow box 
           xmin=252, xmax=Inf, ymin=0, ymax=Inf)  #shadow box highlights CyanoHAB season
SIfig4<- ggarrange(SIfig4a, SIfig4b, nrow=2, ncol=1) #plot figures together
SIfig4 #print figure
ggsave(SIfig4, file="Products/SI_figures/Figure4_color.png") #saves plot to Products folder
write.csv(SIfig4_data, file="Products/SI_figures/Figure4_data.csv") #save data used to make figure
```

Graphs displayed to get cyanoHAB season.
