---
title: "cyanoHAB_severity.Rmd"
author: "Lauren A. Knose, ORISE-EPA"
date: '2024-06-10'
output: html_document
editor_options: 
  chunk_output_type: console
---

The following text and code documents how algal data was processed and analyzed.
Data was also converted to time series.

# Step 1. Load dependent packages and data:

a) loading the dependent packages...

```{r}
library(ggplot2) #needed for plots
library(dplyr) #needed for reshaping/reformating data
library(ggpubr) #needed for graphing multiple plots
library(zoo) #needed for na aggregate
library(plotly) #needed for 3D plots
```

Packages loaded.

b) loading the data...

```{r}
### load phytoplankton cell count data ###
phyto<-read.csv(file="Original_data/ntl88_v13_phytoplankton.csv", header=TRUE) #has the phytoplankton community dataset

### load chlorophyll-a data ###
Chl<- read.csv(file="Original_data/ntl38_v5_chl.csv", header=TRUE) #has chl data
```

Data loaded. 

# Step 2. Calculate cyanobacteria cell density (CD):

a) reformatting and filtering the data...

```{r}
### reformatting and filtering phyto data ###
Me_cyano<- phyto %>%
  mutate(sampledate=as.Date(sampledate, origin="1899-12-30"), #tells R to format field as Date
         Yr=as.numeric(format(sampledate, "%Y")), #renames field
         daynum=as.numeric(format(sampledate, "%j"))) %>%  #adds field with day
  rename(CD_cellsmL=cells_per_ml)%>% #renames field 
  filter(lakeid=="ME", #filter for Lake Mendota data
         depth_range=="0-8m", #filter for data from integrated samples 0 to 8m
         division=="Cyanophyta", #filter for cyanobacteria
         Yr>=2013 & Yr<=2018) %>% #filter for study period
  select(sampledate, CD_cellsmL)

### reformatting and filtering chl-a data ###
Me_Chl<- Chl %>%
  mutate(sampledate=as.Date(sampledate, origin="1899-12-30"), #tells R to format field as Date
         Yr=as.numeric(format(sampledate, "%Y")), #renames field
         daynum=as.numeric(format(sampledate, "%j"))) %>% #renames field to common name
  rename(Chla_ugL= tri_chl_spec) %>% #renames field to common name with units)
  filter(lakeid=="ME", #filter for Lake Mendota data
         depth_range_m=="0-8", #filter for data from integrated samples 0 to 8m
         rep=="1",#remove QA analytic replicates
         Yr>=2013 & Yr<=2018, #filter for study period
         Yr!=2017) %>% #filter the error data out
  filter(!is.na(Chla_ugL)) %>% #removes NAs in chl-a data
  select(sampledate, Yr, Chla_ugL, daynum) #selects only fields of interest
```

Data reformatted and filtered.

b) calculating cyano cell density (CD)...

```{r}
Me_cyano<- Me_cyano %>%  
  group_by(sampledate) %>% #group all observations by sample date
  summarize(CD_cellsmL=sum(CD_cellsmL)) %>% #adds all the cyanobacteria cells/mL together for each sample date
  mutate(Yr=as.numeric(format(sampledate, "%Y")), #renames field
         daynum=as.numeric(format(sampledate, "%j")))  #adds field with day
```

CD calculated.

# Step 3. View the data available:

a) plotting Chl-a data available...

```{r}
### filter data to cyanoHAB season ###
fig3a_data<- Me_Chl %>%
  filter(daynum>= 189 & daynum<= 259) #day num +- 7 days

### plot cyanoHAB severity as chl-a over time ###
SI_fig3a<- ggplot(data=fig3a_data, aes(x=daynum, y=Chla_ugL)) + #plots Chl-a across date
  geom_point() + #adds points at observations
  geom_line() + #adds lines between 
  facet_wrap(~Yr) +
  theme_classic(base_size=12) +
  labs(x="Day of Year", y="Total Chl-a (ug/L)", title="a)") +
  annotate('rect', xmin=min(fig3a_data$daynum), xmax=max(fig3a_data$daynum),
           ymin=0, ymax=12, alpha=.2, fill='green') + #low/no CyanoHAB
  annotate('rect', xmin=min(fig3a_data$daynum), xmax=max(fig3a_data$daynum),
           ymin=12, ymax=50, alpha=.2, fill='orange')+ #moderate CyanoHAB
  annotate('rect', xmin=min(fig3a_data$daynum), xmax=max(fig3a_data$daynum),
           ymin=50, ymax=Inf, alpha=.2, fill='red') #severe CyanoHAB
SI_fig3a
```

Chl-a data plotted. 

b) Plotting CD data available...

```{r}
### filter data to cyanoHAB season ###
fig3b_data<- Me_cyano %>%
  filter(daynum>= 189 & daynum<= 259) #day num +- 7 days

### plot cyanoHAB severity as CD over time ###
SI_fig3b<- ggplot(data=fig3b_data, aes(x=daynum, y=CD_cellsmL))+
  geom_point() +  #point plot 
  geom_line() + #lines between points
  facet_wrap(~Yr) + #by year
  theme_classic(base_size=12)+
  annotate('rect', xmin=min(fig3b_data$daynum), xmax=max(fig3b_data$daynum),
           ymin=0, ymax=20000, alpha=.2, fill='green') + #low/no CyanoHAB
  annotate('rect', xmin=min(fig3b_data$daynum), xmax=max(fig3b_data$daynum),
           ymin=20000, ymax=100000, alpha=.2, fill='orange')+ #moderate CyanoHAB
  annotate('rect', xmin=min(fig3b_data$daynum), xmax=max(fig3b_data$daynum),
           ymin=100000, ymax=Inf, alpha=.2, fill='red') +#severe CyanoHAB
  labs(x="Day of Year", y="Cyanobacteria Density (cells/mL)", title="b)")
SI_fig3b #print image
```


c) plotting chla and CD together...

```{r}
SI_fig3<- ggarrange(SI_fig3a, SI_fig3b, ncol=2, nrow=1, widths=c(0.5, 0.5))
SI_fig3
ggsave("Products/SI_figures/Figure3_color.png") #saves plot to Products folder
```

Chla and CD data plotted. 

d) merging the data files...
```{r}
### merge data files ###
Chla_CD<- merge(Me_Chl, Me_cyano, by=c("sampledate", "Yr", "daynum"), all=TRUE) 
```

Data files merged.

# Step 4. Bin observations into expected frequencies:

a) defining expected observation frequencies...

```{r}
### define the number of observation years ###
Yr<- seq(from=2013, #define year start of study period
             to=2018, #define year end of study period
             by=1) #creates a sequence every 1 year
Yr #prints the expected years of data

### define the observation frequency ###
daynum_bins<- seq(from=196, #define date start of CyanoHAB season
                  to=252, #define date end of CyanoHAB season
                  by=14) #creates a sequence every 14 days 
daynum_bins #prints the expected observation dates of data

### create data frame with expected observations ###
cyanoHAB_bins<- data.frame(Yr=rep(Yr, each=5), #repeat Yr for every daynum 
                           daynum_bin=rep(daynum_bins, times=6)) #repeat daynum for every year
```

Defined expected observation frequency.

b) binning observed dates into expected dates...

```{r}
Chla_CD <- Chla_CD  %>%
  filter(daynum>= 189 & daynum<= 259) %>% #filter for cyanoHAB season
  mutate(daynum_bin=
         ifelse(daynum<(daynum_bins[2]-7), #if observation is <= bin 2 min
                daynum_bins[1], #yes returns bin 1
                ifelse(daynum>=(daynum_bins[2]-7)&daynum<(daynum_bins[3]-7),
              #if observation is > bin 2 min and <= bin 3 min
              daynum_bins[2], #yes returns bin 2
              ifelse(daynum>=(daynum_bins[3]-7)&daynum<(daynum_bins[4]-7),
                     #if observation is > bin 3 min and <= bin 4 min
                     daynum_bins[3], #yes returns bin 3
                     ifelse(daynum>=(daynum_bins[4]-7)&daynum<(daynum_bins[5]-7),
                            #if observation is > bin 4 min and <= bin 5 min 
                            daynum_bins[4], #yes returns bin 4
                            daynum_bins[5])))))#no returns bin 5
                     
Chla_CD<- Chla_CD %>% 
  filter(!duplicated(cbind(Yr, daynum_bin), fromLast = TRUE))  #remove first duplicate
```

Observed dates assigned to closest expected dates.

c) merging the data frames with expected bins...
Note, R only merges two data frames at one time.

```{r}
cyanoHAB<- merge(cyanoHAB_bins, #x data frame
                 Chla_CD, #y data frame
                 by=c("Yr","daynum_bin"),
                 all.x=TRUE) #define the key fields
```

Data merged.

# Step 5. Check for missing values and address:

a) checking for NAs...

```{r}
### checking for NAs ###
summary(cyanoHAB)
```

3 NAs identified.

b) if NAs randomly missing, replacing with annual mean...

```{r}
### calculate annual means ###
yr_mean<- cyanoHAB %>%
  group_by(Yr) %>%
  summarize(avg_Chla=mean(Chla_ugL, na.rm=TRUE),
            se_Chla=(sd(Chla_ugL, na.rm=TRUE)/sqrt(length(Chla_ugL))), 
            avg_CD=mean(CD_cellsmL, na.rm=TRUE),
            se_CD=(sd(CD_cellsmL, na.rm=TRUE)/sqrt(length(CD_cellsmL))))
yr_mean

### fill missing values with the annual mean ###
cyanoHAB<- cyanoHAB %>%
  group_by(Yr) %>%
  mutate(across(Chla_ugL, ~na.aggregate(., FUN=mean, na.rm=FALSE)),
         across(CD_cellsmL, ~na.aggregate(., FUN=mean, na.rm=TRUE))) %>%
  ungroup()

### check that missing values replaced with means ###
summary(cyanoHAB$Chla_ugL)
summary(cyanoHAB$CD_cellsmL)
```

NAs replaced.

# Step 6. Interpolate weekly response values:

a) define expected frequencies...

```{r}
### determine expected observation frequencies ###
week_bins<- seq(from=196, #define date start of CyanoHAB season
                  to=252, #define date end of CyanoHAB season
                  by=7) #creates a sequence every week 
week_bins #prints the expected observation dates of data

### create data frame with the frequencies wanted ###
week_bins<- data.frame(Yr=rep(Yr, each=9),week_bin=rep(week_bins, times=6))
```

Expected weekly frequencies defined.

b) merging data to expected frequencies... 

```{r}
### merge observations with data frame of frequencies ###
cyanoHAB2<- merge(week_bins, cyanoHAB, by.x=c("Yr","week_bin"), 
                      by.y=c("Yr", "daynum_bin"), all=TRUE) 
```

Weekly frequencies added to observations.

c) interpolating missing daily values...

```{r}
### linearly interpolate the missing values ###
cyanoHAB_approx<- cyanoHAB2 %>%  
  group_by(Yr) %>%
  mutate(Chla_ugL=na.approx(Chla_ugL, na.rm=F, rule=2),
         CD_cellsmL=na.approx(CD_cellsmL, na.rm=F, rule=2)) %>% #interpolates where NA
  ungroup() 
```

Daily observations interpolated.

# Step 7. Add fields with lagged values:

```{r}
cyanoHAB_approx<- cyanoHAB_approx %>%
  mutate(Chla_ugL_k7=lag(Chla_ugL, n=1), 
         CD_cellsmL_k7=lag(CD_cellsmL, n=1),
         Chla_ugL_k14=lag(Chla_ugL, n=2), 
         CD_cellsmL_k14=lag(CD_cellsmL, n=2),
         Chla_ugL_k21=lag(Chla_ugL, n=3), 
         CD_cellsmL_k21=lag(CD_cellsmL, n=3),
         Chla_ugL_k28=lag(Chla_ugL, n=4), 
         CD_cellsmL_k28=lag(CD_cellsmL, n=4)) 
```

Field with lagged values added.

# Step 8. Save as new data files:

```{r}
### save the new Chla data file ###
write.csv(cyanoHAB_approx, file="Cleaned_data/cyanoHAB_severity.csv")
```

# Step 9: Make 3D graphs of response variables for entire year:

a) 3D plot of Chl-a across day number and year...

```{r}
Me_Chl<- Chl %>%
  mutate(sampledate=as.Date(sampledate, origin="1899-12-30"), #tells R to format field as Date
         Chla_ugL= tri_chl_spec, #renames field to common name with units
         ) %>% 
  filter(lakeid=="ME", #filter for Lake Mendota data
         depth_range_m=="0-8", #filter for data from integrated samples 0 to 8m
         rep=="1") %>% #remove QA analytic replicates
  filter(!is.na(Chla_ugL)) %>% #removes NAs in chl-a data
  select(sampledate, Chla_ugL) #selects only fields of interest

date_range<- seq( #creates a vector of sequence data
  min(Me_Chl$sampledate), #starting at the first sample date
  max(Me_Chl$sampledate), #ending at the last sample date
  by="1 day") #observations every day  
interpolated<- approx( #creates a vector of data that linearly interpolates 
  Me_Chl$sampledate, #across date
  Me_Chl$Chla_ugL, #the predicted values of Chl-a
  xout=date_range) #for every 1 day
Chla_interp <- data.frame( #creates a new data frame that has 
  sampledate = as.Date(interpolated$x), #renames the new vector as common field 
  Chla_ugL=interpolated$y) #renames the new vector as common field with units

Chla_interp<- Chla_interp %>%
  mutate(Yr=as.numeric(format(sampledate, "%Y")), #adds field with year
         daynum=as.numeric(format(sampledate, "%j")) #adds field with day of yr
         ) %>%
  filter(Yr>=2013 & Yr<=2018) #filter for study period

SI_fig4a<- plot_ly(x=Chla_interp$Yr, 
  y=Chla_interp$daynum, 
  z=Chla_interp$Chla_ugL, 
  type="scatter3d", mode="markers",
  color=Chla_interp$Chla_ugL) %>%
  layout(scene=list(xaxis=list(title="Year"),
         yaxis=list(title="Day of Year"),
         zaxis=list(title="Chlorophyll-a (ug/L)")))
SI_fig4a #prints the plot
#save plot mannually
```

b) 3D plot of CD across day number and year...

```{r}
Me_cyanoRB<- phyto %>%
  mutate(sampledate=as.Date(sampledate, origin="1899-12-30")) %>% #tells R to format field as Datey
  filter(lakeid=="ME", #filter for Lake Mendota data
         depth_range=="0-8m", #filter for data from integrated samples 0 to 8m
         division=="Cyanophyta") %>% #filter for cyanobacteria data
  group_by(sampledate) %>% #group all observations by sample date
  summarize(relbiovol=sum(relative_total_biovolume), #add all observations of cyanobacteria biovolumes together for each sample date
            CD_cellsmL=sum(cells_per_ml)) %>% #adds all the cyanobacteria cells/mL together for each sample date
  filter(relbiovol<=100) %>% #remove any invalid records (>100%  not valid entry)   
  mutate(Yr=as.numeric(format(sampledate, "%Y")), #renames field
         daynum=as.numeric(format(sampledate, "%j"))) %>% #adds field with day
  filter(Yr>=2013 & Yr<=2018) #filter for study years

date_range<- seq( #creates a vector of sequence data
  min(Me_cyanoRB$sampledate), #starting at the first sample date
  max(Me_cyanoRB$sampledate), #ending at the last sample date
  by="1 day") #observations every day  
interpolated<- approx( #creates a vector of data that linearly interpolates 
  Me_cyanoRB$sampledate, #across date
  Me_cyanoRB$CD_cellsmL, #the predicted values of Cyanos
  xout=date_range) #for every 1 day
CD_interp <- data.frame( #creates a new data frame that has 
  sampledate = as.Date(interpolated$x), #renames the new vector as common field 
  CD_cellsmL=interpolated$y) #renames the new vector as common field with units

CD_interp<- CD_interp %>%
  mutate(Yr=as.numeric(format(sampledate, "%Y")), #adds field with year
         daynum=as.numeric(format(sampledate, "%j")) #adds field with day of yr
         ) %>%
  filter(Yr>=2013 & Yr<=2018) #filter for study period

SI_fig4b<- plot_ly(x=CD_interp$Yr, 
  y=CD_interp$daynum, 
  z=CD_interp$CD_cellsmL, 
  type="scatter3d", mode="markers",
  color=CD_interp$CD_cellsmL)%>%
  layout(scene=list(xaxis=list(title="Year"),
         yaxis=list(title="Day of Year"),
         zaxis=list(title="Cyanobacteria Density (cells/mL)")))
SI_fig4b #prints the figure
#save manually
```

3D plots graphed.
