---
title: "Stream_P_datamerge"
author: "Lauren Knose"
date: "2024-04-25"
output: html_document
editor_options: 
  chunk_output_type: console
---

The purpose of this package is to take original stream data and merge into a single data file

# Step 1. Load the data and packages needed:

a) Loading the packages needed...

```{r}
library(dplyr) #needed for transforming data
library(data.table) #needed for matching values from 2 data frames
```

Packages loaded.


b) Loading the data needed...

```{r}
### load original data for stream A ###
PB_1<- read.csv(file="Original_data/USGS_PheasantBranch_P.csv") #load USGS data, Pheasant Branch
PB_2<- read.csv(file="Original_data/WIDNR_PheasantBRanch_P.csv") #load WIDNR data, Pheasant Branch

### load original data for stream B ###
SMC_1<- read.csv(file="Original_data/USGS_SixMileCreek_P.csv") #load USGS data, Six Mile Creek
SMC_2<- read.csv(file="Original_data/WIDNR_SixMileCreek_P.csv") #load WIDNR data, Six Mile Creek

### load original data for stream C ###
YR_1<- read.csv(file="Original_data/USGS_YaharaRiver_P.csv") #load USGS data, Yahara River
YR_2<- read.csv(file="Original_data/WIDNR_YaharaRiver_P.csv") #load WIDNR data, Yahara River
```

# Step 2. Clean and reformat the data:

a) reformatting the data of stream A...

```{r}
### rename fields of data frame 1 ###
colnames(PB_1) #print column names
colnames(PB_1)[colnames(PB_1) =='USGS_MeanSRP_mgL']<- 'SRP_mgL' #rename column
colnames(PB_1)[colnames(PB_1) =='USGS_MeanSRP_lbday']<- 'SRP_lbday' #rename column
colnames(PB_1)[colnames(PB_1) =='USGS_MeanTotP_lbday']<- 'TotP_lbday' #rename column
colnames(PB_1)[colnames(PB_1) =='USGS_MeanTotP_mgL']<- 'TotP_mgL' #rename column
colnames(PB_1)[colnames(PB_1) =='datetime']<- 'sampledate' #rename column
colnames(PB_1)[colnames(PB_1) =='Site_Name']<- 'Site' #rename column
colnames(PB_1) #print column names to check they were changed

### rename fields of data frame 2 to match 1 ###
colnames(PB_2)
colnames(PB_2)[colnames(PB_2) =='PO4_mgL']<- 'SRP_mgL' #rename column
PB_2$X=NULL #removed empty field
PB_2$X.1=NULL #removed empty field
PB_2$X.2=NULL #removed empty field
colnames(PB_2) #print column names to check they were changed

### format data in fields to match ###
PB_1<- PB_1 %>%
  mutate(sampledate=as.Date(sampledate, origin="1899-12-30", format="%m/%d/%Y"))

PB_2<- PB_2 %>%
  mutate(sampledate=as.Date(sampledate, origin="1899-12-30", format="%m/%d/%Y"))
```

Stream A data formatted.

b) reformatting the data of stream B...

```{r}
### rename fields of data frame 1 ###
colnames(SMC_1) #print column names
colnames(SMC_1)[colnames(SMC_1) =='USGS_MeanSRP_mgL']<- 'SRP_mgL' #rename column
colnames(SMC_1)[colnames(SMC_1) =='USGS_MeanSRP_lbday']<- 'SRP_lbday' #rename column
colnames(SMC_1)[colnames(SMC_1) =='USGS_MeanTotP_lbday']<- 'TotP_lbday' #rename column
colnames(SMC_1)[colnames(SMC_1) =='USGS_MeanTotP_mgL']<- 'TotP_mgL' #rename column
colnames(SMC_1)[colnames(SMC_1) =='datetime']<- 'sampledate' #rename column
colnames(SMC_1) #print column names to check they were changed

### rename fields of data frame 2 to match 1 ###
colnames(SMC_2)
colnames(SMC_2)[colnames(SMC_2) =='DissP_PO4_mgL']<- 'SRP_mgL' #rename column
SMC_2$X=NULL #removed empty field
SMC_2$X.1=NULL #removed empty field
SMC_2$X.2=NULL #removed empty field
SMC_2$X.3=NULL #removed empty field
colnames(SMC_2) #reprint 

### format data in fields to match ###
SMC_1<- SMC_1 %>%
  mutate(sampledate=as.Date(sampledate, origin="1899-12-30", format="%m/%d/%Y"))

SMC_2<- SMC_2 %>%
  mutate(sampledate=as.Date(sampledate, origin="1899-12-30", format="%m/%d/%Y"))
```

Stream B data formatted.

c) reformating the data of stream C...

```{r}
### rename fields of data frame 1 ###
colnames(YR_1) #print column names
colnames(YR_1)[colnames(YR_1) =='USGS_MeanSRP_mgL']<- 'SRP_mgL' #rename column
colnames(YR_1)[colnames(YR_1) =='USGS_MeanSRP_lbday']<- 'SRP_lbday' #rename column
colnames(YR_1)[colnames(YR_1) =='USGS_MeanTotP_lbday']<- 'TotP_lbday' #rename column
colnames(YR_1)[colnames(YR_1) =='USGS_MeanTotP_mgL']<- 'TotP_mgL' #rename column
colnames(YR_1)[colnames(YR_1) =='datetime']<- 'sampledate' #rename column
YR_1$Site<- as.character('Yahara River') #add column with name
colnames(YR_1) #print column names to check they were changed

### rename fields of data frame 2 to match 1 ###
colnames(YR_2)
YR_2$X=NULL #removed empty field
colnames(YR_2) #print column names to check they were changed

### format data in fields to match ###
YR_1<- YR_1 %>%
  mutate(sampledate=as.Date(sampledate, origin="1899-12-30", format="%m/%d/%Y"))

YR_2<- YR_2 %>%
  mutate(sampledate=as.Date(sampledate, origin="1899-12-30", format="%Y-%m-%d"))
```

Stream C data formatted.

# Step 3. Combine data for each source:

a) combine data for stream A...

```{r}
### save shared column names as vector ###
nm2 <- intersect(names(PB_1)[-1], names(PB_2)[-1])
### make a copy of data frame 1 ###
PB <- copy(PB_1)
### add data from data frame 2 if conditions match in data frame 1 ###
setDT(PB)[PB_2, (nm2) := Map(fcoalesce, mget(nm2),
       mget(paste0('i.', nm2))), on = .(sampledate)]
### count the NAs removed by adding data ###
sum(is.na(PB_1$TotP_mgL)) - sum(is.na(PB$TotP_mgL))
```

Stream 1 data combined.

b) combining data for stream 2...

```{r}
### save shared column names as vector ###
nm2 <- intersect(names(SMC_1)[-1], names(SMC_2)[-1])
### make a copy of data frame 1 ###
SMC <- copy(SMC_1)
### add data from data frame 2 if conditions match in data frame 1 ###
setDT(SMC)[SMC_2, (nm2) := Map(fcoalesce, mget(nm2),
       mget(paste0('i.', nm2))), on = .(sampledate)]
### count the NAs removed by adding data ###
sum(is.na(SMC_1$TotP_mgL)) - sum(is.na(SMC$TotP_mgL))
```
Stream 2 data combined.

c) combining data for stream C…

```{r}
### save shared column names as own vector ###
nm2 <- intersect(names(YR_1)[-1], names(YR_2)[-1])
### make a copy of data frame 1 ###
YR <- copy(YR_1)
### add data from data frame 2 if conditions match in data frame 1 ###
setDT(YR)[YR_2, (nm2) := Map(fcoalesce, mget(nm2),
       mget(paste0('i.', nm2))), on = .(sampledate)]
### count the NAs removed by adding data ###
sum(is.na(YR_1$TotP_mgL)) - sum(is.na(YR$TotP_mgL))
```

# Step 4. Save data as new data file:

```{r}
write.csv(PB, file="Cleaned_data/PheasantBranch_P.csv")
write.csv(SMC, file="Cleaned_data/SixMileCreek_P.csv")
write.csv(YR, file="Cleaned_data/YaharaRiver_P.csv")
```