---
title: "TotalP_external_inflows.Rmd"
author: "Lauren A. Knose, ORISE-EPA"
date: '2024-05-14'
output:
  html_document: default
  pdf_document: default
editor_options: 
  chunk_output_type: console
---

The purpose of this program is to calculate the external total P loading to Lake Mendota, from the three main inflows of Pheasant Branch (PB), Yahara River (YR), and Six-Mile Creek (SMC). 
P_load (kg/day) = P concentration * flow rate


# Step 1. Load the data and packages needed:

a) Loading the packages needed...

```{r}
library(dplyr) #needed for transforming data
library(ggplot2) #needed for plotting
library(ggpubr) #needed to plot several graphs together
library(rloadest) #needed to estimate daily concentrations from flow
```

Packages loaded.

b) Loading the data needed...

```{r}
### load data for stream A ###
PB_P_flow<- read.csv(file="Cleaned_data/PheasantBranch_P.csv")

### load data for stream B ###
SMC_P_flow<- read.csv(file="Cleaned_data/SixMileCreek_P.csv")

### load data for stream C ###
YR_P_flow<- read.csv(file="Cleaned_data/YaharaRiver_P.csv")
```

Total P data loaded.

# Step 2. Reformat and transform the data:

```{r}
### define unit conversions ###
lb_kg= 0.453592/1 #conversion factor from lb to kg (1lb = 0.453592 kg)
cf_L= 1/0.0353147 #conversion factor from ft^3 to L (1 ft^2 = 28.3168 L)
sec_day= 86400/1 #conversion factor from sec to day (86400 sec = 1 day)
mg_kg= 1/1000000 #conversion factor from mg to kg (1,000,000 mg = 1 kg)

### reformat data of stream A ###
PB_P<- PB_P_flow %>%
  mutate(sampledate=as.Date(sampledate, origin="1899-12-30", format="%Y-%m-%d"), #format date field
         Yr= as.integer(format(sampledate, "%Y")), #add year field
         daynum=as.integer(format(sampledate, "%j")), #add day number field
         Q_Lday=(Qout_cfs*cf_L*sec_day)) %>% #rename column for consistency across data frames
  filter(!is.na(Q_Lday)) %>% #remove any NAs in flow rate
  filter(!is.na(TotP_mgL)) %>% #remove any NAs in total P
  filter(Q_Lday>=0) #remove any negative flow rates

### reformat data of stream B ###
SMC_P<- SMC_P_flow %>%
  mutate(sampledate=as.Date(sampledate, origin="1899-12-30", format="%Y-%m-%d"),
         Yr= as.integer(format(sampledate, "%Y")), #add year field
         daynum=as.integer(format(sampledate, "%j")), #add day number field
        Q_Lday=(Qout_cfs*cf_L*sec_day)) %>% #rename column for consistency across data frames
  filter(!is.na(Q_Lday)) %>% #remove any NAs in flow rate
  filter(!is.na(TotP_mgL)) %>% #remove any NAs in total P 
  filter(Q_Lday>=0) #removes negative flow rates

### reformat data of stream C ###
YR_P<- YR_P_flow %>%
  mutate(sampledate=as.Date(sampledate, origin="1899-12-30", format="%Y-%m-%d"),
         Yr= as.integer(format(sampledate, "%Y")), #add year field
         daynum=as.integer(format(sampledate, "%j")), #add day number field
         Q_Lday=(Qout_cfs*cf_L*sec_day)) %>% #rename column for consistency across data frames
  filter(!is.na(Q_Lday)) %>% #remove any NAs in flow rate
  filter(!is.na(TotP_mgL)) %>% #remove any NAs in total P
  filter(Q_Lday>=0) #remove negative flow rates
```

Data defined and transformed for all inflows.

# Step 3. Check correlation between stream flow and concentrations:

a) plotting regressions of concentration against flow...

```{r}
A<- ggplot(data=PB_P, aes(x=log10(Q_Lday),y=log10(TotP_mgL)))+
  geom_point() + 
  labs(x= "log10(Flow rate) (L/day)", y= "log10(Total P) (mg/L)", title = "Pheasant Branch") +
  geom_smooth()+
  theme_classic()

B<- ggplot(data=SMC_P, aes(x=log10(Q_Lday),y=log10(TotP_mgL)))+
  geom_point() + 
  labs(x= "log10(Flow rate) (L/day)", y= "log10(Total P) (mg/L)", title = "Six Mile Creek") +
  geom_smooth()+
  theme_classic()

C<- ggplot(data=YR_P, aes(x=log10(Q_Lday),y=log10(TotP_mgL)))+
  geom_point() + 
  labs(x= "log10(Flow rate) (L/day)", y= "log10(Total P) (mg/L)", title = "Yahara River") +
  geom_smooth()+
  theme_classic()

### display regression plots together ###
SIfig5<-ggarrange(A,B,C, ncol=3, nrow=1) #prints plots together
SIfig5 #Prints the figure
ggsave(SIfig5, file="Products/SI_figures/Figure5_color.png") #save figure
```

Regressions plotted for each inflow.

b) regressing concentrations against flow, given day of year...

```{r} 
### print regression model results ###
PB_flowP<- lm(log10(PB_P$TotP_mgL) ~ log10(PB_P$Q_Lday)*daynum, data=PB_P) #write the model using simple linear regression
summary(PB_flowP) #print summary table for stream A

SMC_flowP<- lm(log10(SMC_P$TotP_mgL) ~ log10(SMC_P$Q_Lday)* daynum, data=SMC_P) #write the model
summary(SMC_flowP) #print summary table for stream B

YR_flowP<- lm(log10(YR_P$TotP_mgL) ~ log10(YR_P$Q_Lday)*daynum, data=YR_P) #write the model
summary(YR_flowP) #print summary table for stream C
```

Regression results printed.

# Step 4. Use LOADEST to estimate daily stream P loads:

a) estimating loads for stream A...

```{r}
### use rLOADEST package to model P load by flow ###
PB_flowP2<- loadReg(formula= TotP_mgL ~ model(3), #total P concentration ~ ln(flow rate) - ln(centroid flow rate)
                    data=PB_P, #data file
                    flow="Qout_cfs", #name the field with flow rates
                    dates="sampledate", #name the field with sample date
                    conc.units = "mg/L", #define the units of the response measurements
                    station="Pheasant Branch") #site name
print(PB_flowP2, brief=FALSE, load.only=FALSE)
 #print the output

### use rLOADEST package to estimate daily P load ###
PB_TotP_loads<- predLoad(PB_flowP2, newdata = PB_P, load.units="kg", by="day")

### use rLOADEST package to print diagnostic plots ###
png(filename="Products/SI_figures/Figure6.png", width=900, height= 600)
layout(matrix(c(1,2,3,4,5,6), nrow=2, ncol=3))
plot(PB_flowP2, which=1,set.up=FALSE)
plot(PB_flowP2, which=2,set.up=FALSE)
plot(PB_flowP2, which=3,set.up=FALSE)
#plot(PB_flowP2, which=4,set.up=FALSE) #takes too long to plot...issue somewhere with package
plot(PB_flowP2, which=5,set.up=FALSE)
plot(PB_flowP2, which=6,set.up=FALSE)
dev.off()

### save as new data file ###
write.csv(PB_TotP_loads, file="Cleaned_data/PB_TotP_loads.csv")
```

b) estimating loads for stream B...

```{r}
### use rLOADEST package to model P load by flow ###
SMC_flowP2<- loadReg(formula= TotP_mgL ~ model(4), #total P concentration ~ ln(flow rate) - ln(centroid flow rate)
                    data=SMC_P, #data file
                    flow="Qout_cfs", #name the field with flow rates
                    dates="sampledate", #name the field with sample date
                    conc.units = "mg/L", #define the units of the response measurements
                    station="Six Mile Creek") #site name
print(SMC_flowP2, brief=FALSE, load.only=FALSE)
 #print the output

### use rLOADEST package to estimate daily P load ###
SMC_TotP_loads<- predLoad(SMC_flowP2, newdata = SMC_P, load.units="kg", by="day")

### use rLOADEST package to print diagnostic plots ###
png(filename="Products/SI_figures/Figure7.png", width=900, height= 600)
layout(matrix(c(1,2,3,4,5,6), nrow=2, ncol=3))
plot(SMC_flowP2, which=1,set.up=FALSE)
plot(SMC_flowP2, which=2,set.up=FALSE)
plot(SMC_flowP2, which=3,set.up=FALSE)
#plot(SMC_flowP2, which=4,set.up=FALSE) #takes too long to plot...issue somewhere with package
plot(SMC_flowP2, which=5,set.up=FALSE)
plot(SMC_flowP2, which=6,set.up=FALSE)
dev.off()

### save as new data file ###
write.csv(SMC_TotP_loads, file="Cleaned_data/SMC_TotP_loads.csv")
```

c) estimating loads for stream C...

```{r}
### use rLOADEST package to model P load by flow ###
YR_flowP2<- loadReg(formula= TotP_mgL ~ model(4), #total P concentration ~ ln(flow rate) - ln(centroid flow rate)
                    data=YR_P, #data file
                    flow="Qout_cfs", #name the field with flow rates
                    dates="sampledate", #name the field with sample date
                    conc.units = "mg/L", #define the units of the response measurements
                    station="Yahara River") #site name
print(YR_flowP2, brief=FALSE, load.only=FALSE)
 #print the output

### use rLOADEST package to estimate daily P load ###
YR_TotP_loads<- predLoad(YR_flowP2, newdata = YR_P, load.units="kg", by="day")

### use rLOADEST package to print diagnostic plots ###
png(filename="Products/SI_figures/Figure8.png", width=900, height= 600)
layout(matrix(c(1,2,3,4,5,6), nrow=2, ncol=3))
plot(YR_flowP2, which=1,set.up=FALSE)
plot(YR_flowP2, which=2,set.up=FALSE)
plot(YR_flowP2, which=3,set.up=FALSE)
#plot(YR_flowP2, which=4,set.up=FALSE) #takes too long to plot...issue somewhere with package
plot(YR_flowP2, which=5,set.up=FALSE)
plot(YR_flowP2, which=6,set.up=FALSE)
dev.off()

### save as new data file ###
write.csv(YR_TotP_loads, file="Cleaned_data/YR_TotP_loads.csv")
```

Stream C daily loads estimated.

# Step 5. Merge the data:
Note, R only merges two data frames at a time.

```{r}
### add column with site name ###
PB_TotP_loads$Site<- as.character("Pheasant Branch")
SMC_TotP_loads$Site<- as.character("Six Mile Creek")
YR_TotP_loads$Site<- as.character("Yahara River")

### add data frames as new rows in new data frame ###
Inflows<- rbind(PB_TotP_loads, SMC_TotP_loads, YR_TotP_loads) #keep records with same sample date

### rename fields and reformat ###
Inflows$Site<- as.factor(Inflows$Site) #assign levels
names(Inflows)[names(Inflows)=="Date"] <- "sampledate" #rename column
Inflows<- Inflows %>%
  mutate(Yr= as.integer(format(sampledate, "%Y")), #add year field
         daynum=as.integer(format(sampledate, "%j"))) #add day number field

### add total load from all inflows ###
Inflows_all<- Inflows %>%
  select(sampledate, Flow, Flux, Site) %>%
  group_by(sampledate) %>%
  summarize(Flow=sum(Flow),
            Flux=sum(Flux)) %>%
  mutate(Yr= as.integer(format(sampledate, "%Y")), #add year field
         daynum=as.integer(format(sampledate, "%j"))) #add day number field
Inflows_all$Site<- as.factor("All") #assign levels
Inflows_all$Std.Err<-"" #leave field empty
Inflows_all$SEP<- ""
Inflows_all$L95<- ""
Inflows_all$U95<- ""
Inflows<- rbind(Inflows, Inflows_all) #add "All" records to other dataframe
```

Total P data files merged into one file, "Inflows."

# Step 6. Save new data file.

```{R}
write.csv(Inflows, file="Cleaned_data/TotalP_inflows.csv") #save as new data file
```

Data file saved as "TotalP_inflows.csv" in the Cleaned_data folder.

# Step 7. View and summarize the data:

a) viewing daily total P loading for each inflow...

```{r}
### plot daily load for each inflow and total load ###
SIfig16<- ggplot(data=Inflows, 
       aes(x=daynum, 
           y=Flux)) +
  geom_line()+
  facet_wrap(~Site) +
  theme_classic(base_size=12)+
  labs(y="Daily Stream Total P Load (kg/day)",
       x="Day of Year") +
  scale_y_continuous(trans="sqrt") + #transform y-axis scale to see small values
  annotate('rect', fill='gray60', alpha=.2, #creates a shadow box 
           xmin=-Inf, xmax=196, ymin=0, ymax=Inf) +#shadow box highlights CyanoHAB season
  annotate('rect', fill='gray60', alpha=.2, #creates a shadow box 
           xmin=252, xmax=Inf, ymin=0, ymax=Inf)  #shadow box highlights CyanoHAB season
SIfig16 #prints out plot
ggsave(SIfig16, file="Products/SI_figures/Figure16.png")
```

Plotted Total P for day of year across the three inflows, for all years. 

b) calculating average daily total P loading during the CyanoHAB season...

```{r}
### daily average ###
Inflows_avg<- Inflows %>%
  filter(daynum>= 196 & daynum <=252,
         Yr>=2013 & Yr<=2018) %>% #filter for study period
  group_by(Site) %>% #compute for each inflow
  summarise(mean=mean(Flux, na.rm=TRUE),#average daily total P loading
            se=(sd(Flux, na.rm=TRUE)/sqrt(length(Flux))),
            min=min(Flux, na.rm=TRUE),
            max=max(Flux, na.rm=TRUE))
Inflows_avg
```

Estimated average daily total P loading during CyanoHAB season for each inflow.

c) calculating average annual total P load during CyanoHAB season...

```{r}
Inflows_sumavg<- Inflows %>%
  filter(daynum>= 196 & daynum <=252,
         Yr>=2013 & Yr <= 2018,
         Site=="All") %>%#filter for CyanoHAB season
  group_by(Yr) %>%
  summarize(sum=sum(Flux)) #sum dailyP loads 
Inflows_sumavg
mean(Inflows_sumavg$sum)
sd(Inflows_sumavg$sum, na.rm=TRUE)/sqrt(length(Inflows_sumavg$sum))
```

Average annual stream total P load calculated.
