---
title: "TotalP_external_atmdep"
author: "Lauren A. Knose, ORISE-EPA"
date: '2023-05-24'
output: html_document
editor_options: 
  chunk_output_type: console
---

The purpose of this program is to calculate P load from atmospheric deposition. 
P load from atm = P_wet + P_dry.
P_wet = P concentration * daily Precipitation volume
P_dry = P deposition * Lake surface area


# Step 1. Load dependent packages and data needed:

a) Loading dependent packages...
```{r}
library(dplyr) #needed for reshaping/reformaing data
library(lubridate) #needed for date functions
library(ggplot2) #needed for plots
```

Packages loaded.

b) Loading the data needed...
Note, ensure date data is formatted in R as Date.

```{r}
precip<- read.csv(file="Original_data/NWS_data.csv")
```

Precipitation data loaded.

# Step 2. Transform and filter the data:

a) defining conversion coefficients...

```{r}
km2_m2= 1000000/1 #conversion factor from km^2 to m^2
A_lake_m2 = 39.40*km2_m2 #surface area of lake in m^2 
mg_kg = 1/1000000 #conversion factor from mg to kg
in_m = 1/39.3701 #conversion factor from in to m (1 m = 39.701 inches)
m3_L = 1000/1 #conversion factor from m^3 to L (1000 L = 1 m^3)
P_atm_dry = 0.17 #mg P/m2-day rate of P dry deposition during summer (Soranno et 
#al 1997)
P_atm_wet = 0.032 #mg/L rate of P wet deposition (Soranno et al 1997)
```

Conversion coefficients defined.

b) transforming and filtering the data...
```{r}
precip<- precip %>%
  mutate(sampledate=as.Date(sampledate, format="%m/%d/%Y", #reformat as date
                            origin="1899-12-30")) %>%
  filter(!Precipitation_in=="T") %>% #remove flagged records with value "T"
  mutate(Precipitation_in=as.numeric(Precipitation_in)) %>% #convert to numeric
  filter(!is.na(Precipitation_in)) %>% #remove NAs
  mutate(dryP_atm_kgday=((P_atm_dry*A_lake_m2)*mg_kg), #calculate dry P
         wetP_atm_kgday=(P_atm_wet*m3_L*(Precipitation_in*in_m*A_lake_m2)*mg_kg),
         #calculate wet P deposited
         totP_atm_kgday=(dryP_atm_kgday+wetP_atm_kgday), #calculate total P_atm
         daynum=format(sampledate, format = "%j")
        )
```

Total P from atmospheric deposition transformed and filtered.

# Step 3. Interpolate data between observations to get daily loads: 

a) interpolating...
```{r}
atm_date_range<- seq( #creates a vector of sequence data
  min(precip$sampledate), #starting at the first sample date
  max(precip$sampledate), #ending at the last sample date
  by="1 day") #observations every 1 day 
atm_interpolated<- approx( #creates a vector of data that linearly interpolates 
  precip$sampledate, #across date
  precip$totP_atm_kgday, #the predicted values of total P
  xout=atm_date_range) #for every 1 day
precip_interp <- data.frame( #creates a new data frame that has 
  sampledate = as.Date(atm_interpolated$x), #measurements every 1 day
                             atm_TotP_kgday=atm_interpolated$y) #predicted total P

precip_interp <-precip_interp %>%
  mutate(daynum=as.integer(format(sampledate, format="%j")), #add column with 
         #daynum for filtering
         Yr=as.integer(format(sampledate, format="%Y")) #add column with Yr 
         )
```

Total P loading interpreted for atmospheric deposition.

# Step 4. Save estimated P atm deposition as separate data file:

```{r}
write.csv(precip_interp, file="Cleaned_data/TotalP_atm.csv")
```

New file saved as "TotalP_atm.csv" in the Cleaned_data folder.
# Step 4. View and summarize the data:

a) plotting atmospheric deposition total P load by day of year...
```{r}
ggplot(data=precip_interp, 
       aes(x=daynum, y=atm_TotP_kgday)) +
  geom_point() +
  geom_smooth(col="red", lty="dashed") +
  theme_classic(base_size=12) +
  labs(x="Day of Year", y="Total P deposited (wet and dry), in kg/day") 
```

Data plotted.

b) calculating average daily total P loading from atm deposit during CyanoHAB season...

```{r}
precip_dayavg<- precip_interp %>%  
  filter(daynum >=196 & daynum <= 258) %>%
  summarize(mean=mean(atm_TotP_kgday, #average daily total P loading
                      na.rm=TRUE),
            se=(sd(atm_TotP_kgday, na.rm=TRUE)/sqrt(length(atm_TotP_kgday))),
            min=min(atm_TotP_kgday, na.rm=TRUE),
            max=max(atm_TotP_kgday, na.rm=TRUE)
            )
precip_dayavg
```

c) calculating average annual total P loading from atm deposit during CyanoHAB season...

```{r}
precip_sumtot<- precip_interp %>%
  filter(daynum>= 196 & daynum <=258) %>% #filter for CyanoHAB season
  group_by(Yr) %>% #group all records by year
  summarise(sum_TP_kgyr=sum(atm_TotP_kgday))  #calculate the total P load
precip_sumtot
sum_table<-precip_sumtot %>%
  summarize(mean=mean(sum_TP_kgyr),
          se=(sd(sum_TP_kgyr)/sqrt(length(sum_TP_kgyr))),
            min=min(sum_TP_kgyr),
            max=max(sum_TP_kgyr)
            )
sum_table
```

d) plotting P_atm by year (cyanoHAB season only)...

```{r}
ggplot(precip_sumtot, aes(x=as.factor(Yr), y=sum_TP_kgyr)) +
  geom_col() +
  theme_classic(base_size=12) +
  labs(x="Year", y="Total P loading from atmospheric deposition (kg/yr)") 
```

Total P load from atm deposit plotted by year for CyanoHAB season.
