---
title: "TotalP_external_atmdep"
author: "Lauren A. Knose, ORISE-EPA"
date: '2024-05-30'
output: html_document
editor_options: 
  chunk_output_type: console
---

The purpose of this program is to calculate P load from atmospheric deposition. 
P load from atm = P_wet + P_dry.
P_wet = P concentration * daily Precipitation volume
P_dry = P deposition * Lake surface area


# Step 1. Load dependent packages and data needed:

a) Loading dependent packages...
```{r}
library(dplyr) #needed for reshaping/reformaing data
library(ggplot2) #needed for plots
```

Packages loaded.

b) Loading the data needed...
Note, ensure date data is formatted in R as Date.

```{r}
precip<- read.csv(file="Original_data/NWS_data.csv")
```

Precipitation data loaded.

# Step 2. Format and filter the data:

a) filtering data...
```{r}
precip<- precip %>%
  select(sampledate, Precipitation_in) %>%
  mutate(sampledate=as.Date(sampledate, format="%m/%d/%Y", #reformat as date
                            origin="1899-12-30"), #format as date
         Yr= as.integer(format(sampledate, "%Y")), #add year field
         daynum=as.integer(format(sampledate, "%j"))) %>% #add daynum
  filter(Yr>=2013 & Yr<=2018) #filter for study period
```

Data filtered.

b) formatting data...

```{r}
precip2<- precip %>%
  mutate(Precipitation_in=ifelse(Precipitation_in=="T", "", Precipitation_in),#remove flagged records with value "T"
         Precipitation_in=as.numeric(Precipitation_in)) #format as numeric
```

Data formated.

c) address missing values...

```{r}
precip2<- precip2 %>%
  mutate(Precipitation_in=na.approx(Precipitation_in, na.rm=F, rule=2)) #fill missing vlaues with na approx or closest value
summary(precip2$Precipitation_in) #check for NAs
```

Missing values interpolated

# Step 3. Calculate the daily P loads:

a) defining conversion coefficients...

```{r}
### define conversion coefficients ###
km2_m2= 1000000/1 #conversion factor from km^2 to m^2
A_lake_m2 = 39.40*km2_m2 #surface area of lake in m^2 
mg_kg = 1/1000000 #conversion factor from mg to kg
in_m = 1/39.3701 #conversion factor from in to m (1 m = 39.701 inches)
m3_L = 1000/1 #conversion factor from m^3 to L (1000 L = 1 m^3)

### define the deposition rates ###
P_atm_dry = 0.17 #mg P/m2-day rate of P dry deposition during summer (Soranno et 
#al 1997)
P_atm_wet = 0.032 #mg/L rate of P wet deposition (Soranno et al 1997)
```

Conversion coefficients defined.

b) transforming and filtering the data...

```{r}
atmP<- precip2 %>%
  mutate(dryP_atm_kgday=((P_atm_dry*A_lake_m2)*mg_kg), #calculate dry P
         wetP_atm_kgday=(P_atm_wet*m3_L*(Precipitation_in*in_m*A_lake_m2)*mg_kg),
         #calculate wet P deposited
         totP_atm_kgday=(dryP_atm_kgday+wetP_atm_kgday)) #calculate total P_atm
summary(atmP) #check for NAs
```

Total P from atmospheric deposition transformed and filtered.

# Step 3. Save estimated P atm deposition as separate data file:

```{r}
write.csv(atmP, file="Cleaned_data/TotalP_atm.csv")
```

New file saved as "TotalP_atm.csv" in the Cleaned_data folder.
# Step 4. View and summarize the data:

a) calculating average daily total P loading from atm deposit during CyanoHAB season...

```{r}
precip_dayavg<- atmP %>%  
  filter(daynum>= 196 & daynum<= 252) %>% #cyanoHAB season
  summarize(mean=mean(totP_atm_kgday, #average daily total P loading
                      na.rm=TRUE),
            se=(sd(totP_atm_kgday, na.rm=TRUE)/sqrt(length(totP_atm_kgday))),
            min=min(totP_atm_kgday, na.rm=TRUE),
            max=max(totP_atm_kgday, na.rm=TRUE)
            )
precip_dayavg
```

b) calculating average annual total P loading from atm deposit during CyanoHAB season...

```{r}
precip_sumtot<- atmP %>%
  filter(daynum>= 196 & daynum<= 252) %>% #cyanoHAB season
  group_by(Yr) %>% #group all records by year
  summarise(sum_TP_kgyr=sum(totP_atm_kgday))  #calculate the total P load
precip_sumtot
sum_table<-precip_sumtot %>%
  summarize(mean=mean(sum_TP_kgyr),
          se=(sd(sum_TP_kgyr)/sqrt(length(sum_TP_kgyr))),
            min=min(sum_TP_kgyr),
            max=max(sum_TP_kgyr)
            )
sum_table
```

c) plotting P_atm by year (cyanoHAB season only)...

```{r}
SI_fig11<- ggplot(precip_sumtot, aes(x=as.factor(Yr), y=sum_TP_kgyr)) +
  geom_col() +
  theme_classic(base_size=12) +
  labs(x="Year", y="Total P loading from atmospheric deposition (kg/yr)")
SI_fig11
ggsave("Products/SI_figures/Figure11.png")
```

Total P load from atm deposit plotted by year for CyanoHAB season.
