---
title: "TotalP_external_all.Rmd"
author: "Lauren A. Knose, ORISE-EPA"
date: '20234-05-31'
output: html_document
editor_options: 
  chunk_output_type: console
---

The purpose of this program is to estimate total external P load from atmospheric deposition, stream inflows, and other sources.

# Step 1. Load dependent packages and data needed:

a) Loading dependent packages...

```{r}
library(ggplot2) #needed for plots
library(dplyr) #needed for reformaing data
library(reshape2) #needed for reshaping data frames from wide to long
```

Packages loaded.

b) Loading dependent data....

```{r}
streams<- read.csv(file="Cleaned_data/TotalP_inflows.csv") 
atm<- read.csv(file="Cleaned_data/TotalP_atm.csv")
```

Data loaded.

# Step 2. Reformat and merge data:

a) reformatting stream data...

```{r}
### identify keys to filter ###
colnames(streams) #print column names
unique(streams$Site) #print the different site names

### filter stream data ###
streams<- streams %>%
  mutate(X=NULL, #remove index column
         sampledate=as.Date(sampledate,
                            origin="1899-12-30",
                            format="%Y-%m-%d")) %>%
  mutate(Pstreams_kgday=Flux) %>%#reformat date
  filter(Yr>=2013 & Yr<=2018) %>% #filter for the study period
  select(sampledate, Yr, daynum, Site, Pstreams_kgday) #select key fields

### transform data frame long to wide ###
streams2<- dcast(streams, sampledate + Yr + daynum ~ Site, value.var="Pstreams_kgday")

### create field with alternative value for target inflow nutrient reduction scenario ###
streams3<- streams2 %>%
  rename(PB_kgday="Pheasant Branch",
         SMC_kgday="Six Mile Creek",
         YR_kgday="Yahara River",
         Pstreams_kgday="All") %>%
  mutate(Pstreams_kgday_CF1=(Pstreams_kgday-(PB_kgday*0.1))) %>% #counterfactual scenario 1
  select(sampledate, Yr, daynum, Pstreams_kgday, Pstreams_kgday_CF1)
```

Stream data reformatted.

b) reformatting atm data...
```{r}
### identify keys to filter ###
colnames(atm) #print column names

### filter atmospheric deposition data ###
atm2<- atm %>%
  mutate(X=NULL, #remove index column
         sampledate=as.Date(sampledate, origin="1899-12-30"),
         Patm_kgday=totP_atm_kgday)%>% #reformat date
  filter(Yr>=2013 & Yr<=2018) #filter for the study period
```

Atm data reformatted.

b) merging data...

```{r}
### merge the two data files ###
extP_load<- merge(atm2, streams3, by=c("Yr", "daynum", "sampledate")) #adds two data files together
```

Data merged.

# Step 3. Calculate Pother and total P loads:

```{r}
### add fields for total P with and without other P sources ###
extP_load <- extP_load %>%
  mutate(Pother_kgday=((Patm_kgday + Pstreams_kgday)*.47), #assign new column for P_other with 47% of total
         Pext_kgday=((Patm_kgday + Pstreams_kgday)*1.47),
         Pext_kgday_CF1=((Patm_kgday + Pstreams_kgday_CF1)+Pother_kgday))  #increase P_ext 47% to account for P_other
```

External P loading merged.

# Step 3. Save data as new data file:

```{r}
write.csv(extP_load, file="Cleaned_data/TotalP_allext.csv")
```

New data file created with total external P loading.

# Step 4. Summarize and view data:

a) calculating average daily and average annual external P loads...

```{r}
### average daily external P load ###
Pext_day<- extP_load %>%
  filter(daynum>=196 & daynum<=252) %>% #filter for the CyanoHAB season)
  summarize(mean=mean(Pext_kgday, na.rm=TRUE), #calculate mean
            se=(sd(Pext_kgday, na.rm=TRUE)/sqrt(length(Pext_kgday))),
            min=min(Pext_kgday, na.rm=TRUE),
            max=max(Pext_kgday, na.rm=TRUE))
Pext_day # the estimated daily average  P_ext with P_other

### calculate average Pext per m2 ###
A_lake_km2 = 39.40 #surface area of lake in km^2
Pext_day/A_lake_km2 

### sum annual external P load ###
Pext_yr<- extP_load %>%
  filter(daynum>=196 & daynum<=252) %>% #filter for the CyanoHAB season
  mutate(Yr=format(sampledate, "%Y")) %>% #add field with year
  group_by(Yr) %>%
  summarize(sum=sum(Pext_kgday, na.rm = TRUE)) #calculate total P_ext per year
Pext_yr

### average annual external P load ###
Pext_yr_avg<- Pext_yr %>%
  summarize(mean=mean(sum, na.rm=TRUE), #print the average annual
            se=sd(sum, na.rm=TRUE)/sqrt(length(sum)),
            min=min(sum, na.rm=TRUE),
            max=max(sum, na.rm=TRUE)) #print sd for avg
Pext_yr_avg #the estimated annual average P_ext with the P_other
```

P_ext summarized.

b) plotting annual loads by source...

```{r}
### reshape data frame to long format ###
extP_load_long<- extP_load %>%
  select(sampledate, Yr, daynum, Patm_kgday, Pstreams_kgday, Pother_kgday, Pext_kgday) 
extP_load_long<-melt(extP_load_long, #the table you want to reshape wide to long
       id.vars=c("sampledate", "Yr", "daynum")) #the column to use as unique id

### reformat fields ###
extP_load_long<- extP_load_long %>%
  rename(Source=variable,
         Pext_kgday=value)
extP_load_long$Source<-as.character(extP_load_long$Source) #tell R character data

### select the three sources and rename values ###
extP_load_source<- extP_load_long %>% 
  filter(daynum>=196 & daynum<=252) %>% #filter for cyanoHAB season
  filter(Source=="Patm_kgday"|  Source=="Pstreams_kgday"|Source=="Pother_kgday") #filter for the three sources

extP_load_source$Source[extP_load_source$Source=="Patm_kgday"] <- "atmosphere"
extP_load_source$Source[extP_load_source$Source=="Pstreams_kgday"] <- "streams"
extP_load_source$Source[extP_load_source$Source=="Pother_kgday"] <- "other"

### summarize each external source by year ###
extP_load_source_sum<- extP_load_source %>%
  group_by(Yr, Source) %>%
  summarize(TotP_kgyr=sum(Pext_kgday, na.rm=TRUE)) %>%
  ungroup()
write.csv(extP_load_source_sum, file="Products/SI_figures/Figure15a_data.csv")

### plot total Pext by source and year ###  
SIfig15a<- ggplot(extP_load_source_sum, aes(x=as.character(Yr), y=TotP_kgyr, 
                             fill=Source)) + #separate sources by fill color
  geom_bar(position="stack", stat="identity", color="black") + #stacked bar graph
  scale_fill_manual(values=c("grey", "white", "black")) + #atm = grey, streams=black, other=white
  theme_classic(base_size=12) +
  labs(x="Year", y="External total P load (kg/yr)", title="a)") +
  theme(legend.position=c(0.2, 0.75))
SIfig15a #print plot
ggsave(SIfig15a, file="Products/SI_figures/Figure15a.png")
```

Total P load by source for each year plotted.

c) summarizing daily and annual P_ext by source...

```{r}
### average daily external P load by source ###
Pext_day_source<- extP_load_source %>%
  group_by(Source) %>%
  summarize(mean=mean(Pext_kgday, #average daily total P loading
                      na.rm=TRUE),
            se=(sd(Pext_kgday, na.rm=TRUE)/sqrt(length(Pext_kgday))),
            min=min(Pext_kgday, na.rm=TRUE),
            max=max(Pext_kgday, na.rm=TRUE))
Pext_day_source

### sum annual external P load by source ###
extP_load_long_yr<- extP_load_source %>%
  group_by(Yr, Source) %>%
  summarize(sum=sum(Pext_kgday)) #add total P load
extP_load_long_yr #print table

### average annual external P load by source ###
extP_load_long_yr_avg<- extP_load_long_yr %>%
  group_by(Source) %>%
  summarize(mean=mean(sum, na.rm=TRUE), #print the average annual
            se=sd(sum, na.rm=TRUE)/sqrt(length(sum)),
            min=min(sum, na.rm=TRUE),
            max=max(sum, na.rm=TRUE)) #print sd for avg
extP_load_long_yr_avg  
```

P_ext summarized by source.
