---
title: "TotalP_water.Rmd"
author: "Lauren A. Knose, ORISE-EPA"
date: '2024-05-20'
output: html_document
editor_options: 
  chunk_output_type: console
---

The purpose of this program is to process measurements taken in the surface 
water (epilimnion) and bottom water (hypolimnion) of the lake. 


# Step 1. load dependent packages and data needed:

a) Loading the dependent packages...

```{r}
library(dplyr) #needed for transforming data
library(ggplot2) #needed for plotting
library(rLakeAnalyzer) #needed to find thermocline depth from temperature profile
library(reshape2) #needed to reshape data frame from wide to long or long to wide
library(zoo) #needed to linearly interpolate missing values
library(tidyverse) #needed for crossing function
```

Packages loaded.

b) Loading the data with total P concentrations, water temperature, and external P loads...

```{r}
### load data with total P concentration measurements ###
water_P<- read.csv(file="Original_data/ntl1_v9_1_chem.csv", header=TRUE) #has nutrient data measured in lake

### load data with water temperature measurements ###
watertemp<-read.csv(file="Original_data/ntl29_v12_temp.csv") #data from temp string
```

# Step 2. Reformat and filter data:

a) reformatting and filtering water P data...

```{r}
### reformat water P data ###
water_P2<- water_P %>%
  mutate(sampledate=as.Date(sampledate, origin="1899-12-30")) %>% #format as Date
  rename(Yr=year4, #rename year field
         TotP_mgL=totpuf_sloh, #rename total P field with units (mg/L)
         DisP_mgL=drp_sloh, #rename dissolved P field with units (mg/L)
         Depth_m = depth)  #rename depth field with units (meters)

### filter data to study site and study period ###
water_P2<- water_P2 %>%
  filter(lakeid=="ME", #filter for Lake Mendota
         rep=="1") %>% #remove analytical reps
  filter(Yr>=2013 & Yr <=2018) %>%  #filter for the study years 
  select(sampledate, TotP_mgL, DisP_mgL, Depth_m, daynum, Yr) #%>%
#  filter(!is.na(TotP_mgL)) #remove records with NAs
```

Water column P data formatted and filtered.

b) reformatting and filtering water temperature data...

```{r}
### reformat water temperature data ###
watertemp2<- watertemp %>% #make new data frame
  mutate(sampledate=as.Date(sampledate, format="%m/%d/%Y",#specify format of date
                            origin="1899-12-30"),
         daynum=as.integer(format(sampledate, format="%j"))) %>% #add in day number
  rename(Yr=year4, #rename year field
         Depth_m=depth, #rename column with units
         watertemp_C=wtemp) #rename column with units 
### filter water temperature data ###
watertemp2<- watertemp2 %>%
  filter(lakeid=="ME") %>% #filter for Lake Mendota data only
  filter(rep==1) %>% #filter for rep 1 (no analytical reps)
  filter(Yr>=2013 & Yr <=2018) %>% #filter for the study years 
  select(sampledate, Depth_m, watertemp_C, daynum, Yr) #%>% #select vars needed
#  filter(!is.na(watertemp_C)) #remove NAs
```

Water temperature data reformatted and filtered.

# Step 3. View and summarize data:

a) plotting P concentrations across depth and time...

```{r}
### plot P concentration across both depth and time ###
SI_fig13<- ggplot(data=subset(water_P2, daynum>= 189 & daynum <=259), #create limno plot with depth inverse on y-axis
       aes(x=daynum, #sampled date as x-axis
           y=Depth_m, 
           color=TotP_mgL)) + #and data points colored by temperature
  geom_point() + #plot points
  facet_wrap(~Yr) + #make separate plot for each year
  scale_y_reverse() + #inverses the y-axis from zero to max(depth) for limno plot
  scale_color_gradient(high=scales::muted("red"), #red = high values
                        low=scales::muted("blue"), #blue = low values
                        name="Total P (mg/L)") + #name of legend
  geom_hline(yintercept=8, lty="dashed") + #add line for bottom of epi
  theme_classic() +
  labs(x="Day of Year", y="Depth (m)") 
SI_fig13
ggsave(SI_fig13, file="Products/SI_figures/Figure13_color.png")
```

Total P concentrations plotted across depth and date.

b) plotting water temperature across depth and date...

```{r}
### plot water temp across depth (y-axis) and date (x-axis) ###
SI_fig15<- ggplot(data=subset(watertemp2, daynum>= 189 & daynum <=259), #create limno plot with depth inverse on y-axis
       aes(x=daynum, #sampled date as x-axis
           y=Depth_m, 
           color=watertemp_C)) + #and data points colored by temperature
  geom_point() + #plot points
  facet_wrap(~Yr) + #make separate plot for each year
  scale_y_reverse() + #inverses the y-axis from zero to max(depth) for limno plot
  scale_color_gradient(high=scales::muted("red"), #red = hot temp
                        low=scales::muted("blue"), #blue = cold temp
                        name="Water Temperature (C)") + #name of legend
  geom_hline(yintercept=8, lty="dashed") + #add line for bottom of epi
  theme_classic() +
  labs(x="Day of Year", y="Depth (m)") 
SI_fig15
ggsave(SI_fig15, file="Products/SI_figures/Figure15_color.png")
```

Water temperature plotted across depth and date.

# Step 4. Calculate P_hypo:epi:

a) calculating average P concentration by layer...

```{r}
### define the lower boundary of the epilimnion and highest boundary of the hypolimnion from temperature profiles ###
epi_max_m<- 8
hypo_min_m<- 20

### filter P data to epilimnion ###
Epi_P<- water_P2 %>%
  filter(Depth_m <= epi_max_m) %>% #for observations taken at or above 8 m
  group_by(sampledate) %>% #group all observations in same sample date
  summarize(TotP_mgL=mean(TotP_mgL, na.rm=TRUE),#calculates mean for epi 
            DisP_mgL=mean(DisP_mgL, na.rm=TRUE)) %>%
  mutate(layer="epilimnion") #adds field with layer name

### filter P data to hypolimnion ###
Hypo_P<- water_P2 %>%
  filter(Depth_m >= hypo_min_m) %>% #for observations taken at or below 20m
  group_by(sampledate) %>% #group all observations in same sample date
  summarize(TotP_mgL=mean(TotP_mgL, na.rm=TRUE), #calculates mean for hypo
            DisP_mgL=mean(DisP_mgL, na.rm=TRUE)) %>%
  mutate(layer="hypolimnion") #adds field with layer name
```

Mean P concentrations calculated by layer. 

b) merging and viewing the data...

```{r}
### reshape and reformat data ###
P_ratio_long<- rbind(Epi_P, Hypo_P) %>%
  mutate(Yr=as.integer(format(sampledate, format="%Y")), #add year
         daynum=as.integer(format(sampledate, format="%j"))) #add day number

### plot data ###
SI_fig16<- ggplot(data=subset(P_ratio_long, daynum>= 189 & daynum <=259), aes(x=daynum, color=layer))+ #filter for CyanoHAB season +- 7 days),  #plot thermocline depth across date
   geom_line(aes(y=TotP_mgL), lty="solid") + #
   geom_point(aes(y=TotP_mgL), shape=1) + #add lines between points
   geom_line(aes(y=DisP_mgL), lty="dashed") + #
   geom_point(aes(y=DisP_mgL), shape=2) + #add lines between points
   facet_wrap(~Yr) + #plot individual years
   geom_hline(yintercept=0.03, lty="dashed") +
   scale_linetype_manual(values=c("TotP_mgL"=1, "DisP_mgL"=2)) +
   scale_color_manual(values=c("blue","darkred")) +
  scale_shape_manual(name="P form", labels=c("Total P", "Dissolved P")) +
   theme_classic() +
   labs(x="Date", y="P concentration (mg/L)", color="Layer")#add axis labels
SI_fig16
ggsave(SI_fig16, file="Products/SI_figures/Figure16_color.png")
```

Data plotted.

c) calculating P_hypo:epi...

```{r}
### remove unwanted fields ###
P_ratio_long<- P_ratio_long %>%
  select(!c(DisP_mgL)) #remove unwanted fields

### reshape data frame from long to wide ###
P_ratio<- dcast(P_ratio_long, sampledate+Yr+daynum ~ layer, value.var = "TotP_mgL")

### calculate P_hypo:epi ###
P_ratio<- P_ratio %>%
  rename(epi_totP_mgL=epilimnion, #rename fields with untis
         hypo_totP_mgL=hypolimnion) %>% #rename fields with units
  mutate(TotP_HEratio=ifelse((hypo_totP_mgL/epi_totP_mgL)!=Inf, 
                             (hypo_totP_mgL/epi_totP_mgL), ""))

### fill missing P_hypo:epi with nearest value ###
P_ratio<- P_ratio %>% 
  group_by(Yr) %>% 
  mutate(TotP_HEratio=na.approx(TotP_HEratio, na.rm=FALSE, rule=2)) %>%
  ungroup()
```

P_hypo:epi calculated.

# Step 5. Interpolate missing P values across depth:

a) interpolating missing P concentrations across depth...

```{r}
### create a function to interpolate across depth for each sample date ###
estimate_TotP_by_date <- function(target_date, target_depth) { #write function that
  data_for_date <- water_P2 %>% #creates a new data frame for a single date
    filter(sampledate == target_date) %>% #filters for specific date
    arrange(Depth_m) #and arranges depths min to max
  approx(data_for_date$Depth_m, #linear regression interpolation of
         data_for_date$TotP_mgL, #y values
         xout = target_depth)$y #from depth min to max
}

### filter out any observations that have < 2 values (>2 values needed for interpolation) ###
water_interp<- water_P2 %>%
  filter(daynum>= 100 & daynum <=259) %>% #filter for CyanoHAB season  +- 7 days
  group_by(sampledate) %>% #for each sampledate
  filter(!n()<2) %>% #remove dates with less than 2 observations
  ungroup() #ungroup table

### interpolate P concentrations across depth within each sample date ###
TotP_interp_depth <- crossing( #creates a tibble that 
  tibble(date = unique(water_interp$sampledate)), #has same dates as those in frame
  # depths can now be any value
  tibble(Depth_m = seq(0, 25, #define depth min and max
                       length.out = 26)) #set the number of interpolated values
) %>%
  group_by(date) %>%
  mutate(TotP_mgL = estimate_TotP_by_date(date[1], Depth_m),
         daynum=as.integer(format(date, format="%j"))) %>%
  ungroup()

### save as data frame ###
TotP_interp_depth<- as.data.frame(TotP_interp_depth)
```

Total P interpolated across depth

b) plotting interpolated P across depth..

```{r}
### plot interpolated water P data ###
SI_fig14<- ggplot(data=subset(TotP_interp_depth, daynum>=196 & daynum<=252), #create limno plot with depth inverse on y-axis
       aes(x=as.integer(format(date, "%j")), #sample date as x-axis
           y=Depth_m, #depth as y-axis
           color=TotP_mgL)) + #and data points colored by temperature
  geom_point()+
  facet_wrap(~as.factor(format(date, "%Y"))) + #makes separate plot for each year
  scale_y_reverse() + #inverses the y-axis from zero to max(depth)
  scale_color_gradient(high=scales::muted("red"), #red = high P concentration
                        low=scales::muted("blue"), #blue = low P concentration
                        name="Total P (ug/L)") + 
  geom_hline(yintercept=8, lty="dashed") + #add in horizontal line for epi max
  theme_classic() +
  labs(x="Date", y="Depth (m)")  #add axis labels
SI_fig14 #print plot
ggsave(SI_fig14, file="Products/SI_figures/Figure14_color.png") #save plot
```

P concentration interpolated across depth.

# Step 6. Calculate the change in epilimnion volume:
Note, data frame needs to be in wide format with each depth a separate column (add prefix= "wtr_") to match package function dependencies.

a) calculating change in thermocline depth...

```{r}
### reformat data for input into rLakeAnalyzer ###
wtr<- watertemp2 %>%
  select(sampledate, Depth_m, watertemp_C) #only include needed columns
wtr<- dcast(wtr, sampledate ~ Depth_m) # reshape depth from long to wide
colnames(wtr)<- paste("wtr", colnames(wtr), sep="_") #add wtr_ before column name
# fit package format dependencies
wtr<- wtr %>%
  rename(datetime=wtr_sampledate) #rename column to fit package format dependencies

### determine the thermocline depth ###
t.d<- ts.thermo.depth(wtr, na.rm=TRUE) #calculates thermocline depth as time series

### plot the thermocline depth by date ###
SI_fig17<- ggplot(data=t.d, aes(x=datetime, 
                     y=thermo.depth))+ #plot thermocline depth across date
  geom_point(shape=1) + #points as open circles
  geom_line() + #add lines between points
  scale_y_reverse() + #reverse the y scale for limno plot
  theme_classic() +
  labs(x="Date", y="Thermocline depth (m)")#add axis labels
SI_fig17
ggsave(SI_fig17, file="Products/SI_figures/Figure17.png")

### seperate data by year then assign thermocline depth for each observation (t) ###
t.d2<- t.d %>% #make new table
  mutate(Yr=as.numeric(format(datetime, "%Y"))) %>% #add in year
  group_by(Yr) %>% #for each year of day
  rename(Thermo_depth_m=thermo.depth) %>% #add units and remove duplicate column
  mutate(Thermo_depth_m_k1=lag(Thermo_depth_m, n=1)) %>% #thermocline depth at t-1 
  ungroup()

### reassign first thermocline depth at t-1 each year (Nas) as thermocline depth at t ###
t.d2$Thermo_depth_m_k1[is.na(t.d2$Thermo_depth_m_k1)] <- t.d2$Thermo_depth_m[is.na(t.d2$Thermo_depth_m_k1)]#replace NAs in t-1 with the t value

### calculate change in thermocline depth ###
t.d2<- data.frame(t.d2) %>%
  mutate(Chg_Thermodepth_m= Thermo_depth_m-Thermo_depth_m_k1)
```

Thermocline depth calculated. 

b) calculating volume lost or gained to epi...

```{r}
### calculate the change in thermocline depth between t and t-1 to get the volume ###
A_strat= 25850000 # surface area of stratified region in m^2 
Vol_epi<- t.d2 %>%
  rename(sampledate=datetime) %>%
  mutate(Volchg_epi_m3 = Chg_Thermodepth_m*A_strat,#calculate daily change in volume
         daynum=as.numeric(format(sampledate, "%j"))) #add daynum
```

Thermocline and change in epi volume calculated.

# Step 7. Calculate P added (removed) from change in epi volume:

a) merging P concentrations with epi volume change...
```{r}
### add key fields to data frame A to match with data frame B ###
Vol_epi2<- Vol_epi %>%
  group_by(Yr) %>%
  mutate(sampledate_k=lag(sampledate, n=1),
         Depth_m_k=as.integer(round(Thermo_depth_m_k1, 0))) %>%
  ungroup() 

### add field with matching P value from data frame B ###
Vol_chg<- merge(Vol_epi2, TotP_interp_depth, 
                by.x=c("sampledate_k", "Depth_m_k"), 
                by.y=c("date", "Depth_m"),
                all.x=TRUE)
```

Data merged.

# Step 9: Save the new data files:

```{r}
### save the P concentration and epi volume change ###
write.csv(Vol_chg, file="Cleaned_data/TotalP_water.csv")

### save the P_hypo:epi ###
write.csv(P_ratio, file="Cleaned_data/TotalP_ratio.csv")
```

Data saved.