---
title: "TotalP_water.Rmd"
author: "Lauren A. Knose, ORISE-EPA"
date: '2024-05-20'
output: html_document
editor_options: 
  chunk_output_type: console
---

The purpose of this program is to process measurements taken in the surface 
water (epilimnion) and bottom water (hypolimnion) of the lake. 


# Step 1. load dependent packages and data needed:

a) Loading the dependent packages...

```{r}
library(dplyr) #needed for transforming data
library(ggplot2) #needed for plotting
library(rLakeAnalyzer) #needed to find thermocline depth from temperature profile
library(reshape2) #needed to reshape data frame from wide to long or long to wide
library(zoo) #needed to linearly interpolate missing values
library(tidyverse) #needed for crossing function
```

Packages loaded.

b) Loading the data with total P concentrations, water temperature, and external P loads...

```{r}
### load data with total P concentration measurements ###
water_P<- read.csv(file="Original_data/ntl1_v9_1_chem.csv", header=TRUE) #has nutrient data measured in lake

### load data with water temperature measurements ###
watertemp<-read.csv(file="Original_data/ntl29_v12_temp.csv") #data from temp string
```

# Step 2. Reformat and filter data:

a) reformatting and filtering water P data...

```{r}
### reformat water P data ###
water_P2<- water_P %>%
  mutate(sampledate=as.Date(sampledate, origin="1899-12-30")) %>% #format as Date
  rename(Yr=year4, #rename year field
         TotP_mgL=totpuf_sloh, #rename total P field with units (mg/L)
         DisP_mgL=drp_sloh, #rename dissolved P field with units (mg/L)
         Depth_m = depth)  #rename depth field with units (meters)

### filter data to study site and study period ###
water_P2<- water_P2 %>%
  filter(lakeid=="ME", #filter for Lake Mendota
         rep=="1") %>% #remove analytical reps
  filter(Yr>=2013 & Yr <=2018) %>%  #filter for the study years 
  filter(daynum>= 189 & daynum <=259) %>% #filter for CyanoHAB season +- 7 days
  select(sampledate, TotP_mgL, DisP_mgL, Depth_m, daynum, Yr) #%>%
#  filter(!is.na(TotP_mgL)) #remove records with NAs
```

Water column P data formatted and filtered.

b) reformatting and filtering water temperature data...

```{r}
### reformat water temperature data ###
watertemp2<- watertemp %>% #make new data frame
  mutate(sampledate=as.Date(sampledate, format="%m/%d/%Y",#specify format of date
                            origin="1899-12-30"),
         daynum=as.integer(format(sampledate, format="%j"))) %>% #add in day number
  rename(Yr=year4, #rename year field
         Depth_m=depth, #rename column with units
         watertemp_C=wtemp) #rename column with units 
### filter water temperature data ###
watertemp2<- watertemp2 %>%
  filter(lakeid=="ME") %>% #filter for Lake Mendota data only
  filter(rep==1) %>% #filter for rep 1 (no analytical reps)
  filter(Yr>=2013 & Yr <=2018) %>% #filter for the study years 
  filter(daynum>= 189 & daynum <=259) %>% #filter for CyanoHAB season  +- 7 days
  select(sampledate, Depth_m, watertemp_C, daynum, Yr) #%>% #select vars needed
#  filter(!is.na(watertemp_C)) #remove NAs
```

Water temperature data reformatted and filtered.

# Step 3. View and summarize data:

a) plotting P concentrations across depth and time...

```{r}
### plot P concentration across both depth and time ###
SI_fig14<- ggplot(data=water_P2, #create limno plot with depth inverse on y-axis
       aes(x=daynum, #sampled date as x-axis
           y=Depth_m, 
           color=TotP_mgL)) + #and data points colored by temperature
  geom_point() + #plot points
  facet_wrap(~Yr) + #make separate plot for each year
  scale_y_reverse() + #inverses the y-axis from zero to max(depth) for limno plot
  scale_color_gradient(high=scales::muted("red"), #red = high values
                        low=scales::muted("blue"), #blue = low values
                        name="Total P (mg/L)") + #name of legend
  geom_hline(yintercept=8, lty="dashed") + #add line for bottom of epi
  theme_classic() +
  labs(x="Day of Year", y="Depth (m)") 
SI_fig14
ggsave(SI_fig14, file="Products/SI_figures/Figure14_color.png")
```

Total P concentrations plotted across depth and date.

b) plotting water temperature across depth and date...

```{r}
### plot water temp across depth (y-axis) and date (x-axis) ###
SI_fig15<- ggplot(data=watertemp2, #create limno plot with depth inverse on y-axis
       aes(x=daynum, #sampled date as x-axis
           y=Depth_m, 
           color=watertemp_C)) + #and data points colored by temperature
  geom_point() + #plot points
  facet_wrap(~Yr) + #make separate plot for each year
  scale_y_reverse() + #inverses the y-axis from zero to max(depth) for limno plot
  scale_color_gradient(high=scales::muted("red"), #red = hot temp
                        low=scales::muted("blue"), #blue = cold temp
                        name="Water Temperature (C)") + #name of legend
  geom_hline(yintercept=8, lty="dashed") + #add line for bottom of epi
  theme_classic() +
  labs(x="Day of Year", y="Depth (m)") 
SI_fig15
ggsave(SI_fig15, file="Products/SI_figures/Figure15_color.png")

### define the lower boundary of the epilimnion and highest boundary of the hypolimnion from temperature profiles ###
epi_max_m<- 8
hypo_min_m<- 20
```

Water temperature plotted across depth and date.

# Step 4. Calculate P_hypo:epi:

a) calculating mean epi and hypo P concentrations...

```{r}
### filter P data to epilimnion ###
Epi_P<- water_P2 %>%
  filter(Depth_m <= epi_max_m) %>% #for observations taken at or above 8 m
  group_by(sampledate) %>% #group all observations in same sample date
  summarize(epi_TotP_mgL=mean(TotP_mgL, na.rm=TRUE),
            epi_DisP_mgL=mean(DisP_mgL, na.rm=TRUE)) #calculates mean for all depths in epi 

### filter P data to hypolimnion ###
Hypo_P<- water_P2 %>%
  filter(Depth_m >= hypo_min_m) %>% #for observations taken at or below 20m
  group_by(sampledate) %>% #group all observations in same sample date
  summarize(hypo_TotP_mgL=mean(TotP_mgL, na.rm=TRUE),
            hypo_DisP_mgL=mean(DisP_mgL, na.rm=TRUE)) #calculates mean for all depths in hypo, adds negligible number to avoid zero

### merge the two data frames, add day and year ###
P_ratio<- merge(Epi_P, Hypo_P, by="sampledate")
P_ratio<- P_ratio %>%
  mutate(Yr=as.integer(format(sampledate, format="%Y")),
         daynum=as.integer(format(sampledate, format="%j"))) #add in day number)

### rename columns and add P_hypo:epi ratio ###
P_ratio <- P_ratio %>%
  mutate(TotP_HEratio=ifelse((hypo_TotP_mgL/epi_TotP_mgL)!=Inf, (hypo_TotP_mgL/epi_TotP_mgL), NA), #if value is Inf, replace value with NA
         DisP_HEratio=ifelse((hypo_DisP_mgL/epi_DisP_mgL)!=Inf, (hypo_DisP_mgL/epi_DisP_mgL), NA)) #if value is Inf, replace value with NA

### check for NAs ###
summary(P_ratio$TotP_HEratio)
summary(P_ratio$DisP_HEratio)
```

P_hypo:epi calculated.

b) viewing P_hypo:epi across depth and date...

```{r}
SI_fig16<- ggplot(data=P_ratio, aes(x=daynum))+ #plot thermocline depth across date
  geom_line(aes(y=epi_TotP_mgL), color="blue") + #
  geom_point(aes(y=epi_TotP_mgL),color="blue") + #add lines between points
  geom_line(aes(y=epi_DisP_mgL), color="blue", lty="dashed") + #
  geom_point(aes(y=epi_DisP_mgL),color="blue", shape=2) + #add lines between points
  geom_line(aes(y=hypo_TotP_mgL), color="red") + #
  geom_point(aes(y=hypo_TotP_mgL), color="red") + #add lines between points
  geom_line(aes(y=hypo_DisP_mgL), color="red", lty="dashed") + #
  geom_point(aes(y=hypo_DisP_mgL),color="red", shape=2) + #add lines between points
  facet_wrap(~Yr) + #plot individual years
  geom_hline(yintercept=0.03, lty="dashed") +
  theme_classic() +
  labs(x="Date", y="P concentration (mg/L)")#add axis labels
SI_fig16
ggsave(SI_fig16, file="Products/SI_figures/Figure16_color.png")
```

Average P concentrations graphed by depth and date.

# Step 5. Calculate the change in epilimnion volume between sampling events:
Note, rLakeAnalyzer package is required here. Data frame needs to be in wide
format with each depth a separate column (add prefix= "wtr_") to match package
function dependencies.

a) calculating thermocline depth and epilimnion volume...

```{r}
### reformat data for input into rLakeAnalyzer ###
wtr<- watertemp2 %>%
  select(sampledate, Depth_m, watertemp_C) #only include needed columns
wtr<- dcast(wtr, sampledate ~ Depth_m) # reshape depth from long to wide
colnames(wtr)<- paste("wtr", colnames(wtr), sep="_") #add wtr_ before column name
# fit package format dependencies
wtr<- wtr %>%
  rename(datetime=wtr_sampledate) #rename column to fit package format dependencies

### calculate the thermocline depth ###
t.d<- ts.thermo.depth(wtr, na.rm=TRUE) #calculates thermocline depth as time series

### plot the thermocline depth by date ###
SI_fig17<- ggplot(data=t.d, aes(x=datetime, 
                     y=thermo.depth))+ #plot thermocline depth across date
  geom_point(shape=1) + #points as open circles
  geom_line() + #add lines between points
  scale_y_reverse() + #reverse the y scale for limno plot
  theme_classic() +
  labs(x="Date", y="Thermocline depth (m)")#add axis labels
SI_fig17
ggsave(SI_fig17, file="Products/SI_figures/Figure17.png")

### seperate data by year then assign thermocline depth for each observation (t) ###
t.d2<- t.d %>% #make new table
  mutate(Yr=as.numeric(format(datetime, "%Y"))) %>% #add in year
  group_by(Yr) %>% #for each year of day
  mutate(Thermo_depth_m=thermo.depth, thermo.depth=NULL, #add units and remove duplicate column
         Thermo_depth_m_k1=lag(Thermo_depth_m, n=1)) #thermocline depth at t-1 

### reassign first thermocline depth at t-1 each year (Nas) as thermocline depth at t ###
t.d2$Thermo_depth_m_k1[is.na(t.d2$Thermo_depth_m_k1)] <- t.d2$Thermo_depth_m[is.na(t.d2$Thermo_depth_m_k1)]#replace NAs in t-1 with the t value

### calculate the change in thermocline depth between t and t-1 to get the volume ###
A_strat= 25850000 # surface area of stratified region in m^2 
Vol_epi<- t.d2 %>%
  mutate(Chg_Thermodepth_m = (Thermo_depth_m - Thermo_depth_m_k1),#calculate daily change in depth 
         Vol_epi_m3 = Chg_Thermodepth_m*A_strat,#calculate daily change in volume
         sampledate=datetime, datetime=NULL, #rename column, remove duplicate
         daynum=as.numeric(format(sampledate, "%j")), #add in daynum
         Depth_m=as.integer(round(Thermo_depth_m, 0))) %>% #round the thermocline depth to nearest integer
  select(sampledate, Yr, daynum, Vol_epi_m3, Thermo_depth_m, Depth_m)
```

Thermocline and change in epi volume calculated.

b) plotting change in epi volume over time...

```{r}
SI_fig18<- ggplot(data=Vol_epi, #create plot
       aes(x=daynum, #sampled date as x-axis
           y=Vol_epi_m3)) + #depth as y-axis
  geom_point() + #plot point graph
  geom_line() + #add lines beween points
  facet_wrap(~Yr) + #make separate graph for each year
  geom_hline(yintercept=0, lty="dashed") + #add in horizontal line for zero change
  theme_classic() +
  labs(x="Date", y="Change in Epilimnion Volume (m^3)", #add axis labels 
       caption="Horizontal dashed line indicates zero net change.")
SI_fig18
ggsave(SI_fig18, file="Products/SI_figures/Figure18.png")
```

Change in epilimnion volume plotted. 

# Step 5. Interpolate missing data and bin observations to equal intervals:

a) interpolating missing P concentrations across depth...

```{r}
### create a function to interpolate across depth for each sample date ###
estimate_TotP_by_date <- function(target_date, target_depth) { #write function that
  data_for_date <- water_P2 %>% #creates a new data frame for a single date
    filter(sampledate == target_date) %>% #filters for specific date
    arrange(Depth_m) #and arranges depths min to max
  approx(data_for_date$Depth_m, #linear regression interpolation of
         data_for_date$TotP_mgL, #y values
         xout = target_depth)$y #from depth min to max
}

### filter out any observations that have < 2 values (>2 values needed for interpolation) ###
water_P2<- water_P2 %>%
  group_by(sampledate) %>% #for each sampledate
  filter(!n()<2) %>% #remove dates with less than 2 observations
  ungroup() #ungroup table

### interpolate P concentrations across depth within each sample date ###
TotP_interp_depth <- crossing( #creates a tibble that 
  tibble(date = unique(water_P2$sampledate)), #has same dates as those in frame
  # depths can now be any value
  tibble(Depth_m = seq(0, 25, #define depth min and max
                       length.out = 26)) #set the number of interpolated values
) %>%
  group_by(date) %>%
  mutate(TotP_mgL = estimate_TotP_by_date(date[1], Depth_m)) %>%
  ungroup()

### plot interpolated water P data ###
SI_fig19<- ggplot(data=TotP_interp_depth, #create limno plot with depth inverse on y-axis
       aes(x=as.integer(format(date, "%j")), #sample date as x-axis
           y=Depth_m, #depth as y-axis
           color=TotP_mgL)) + #and data points colored by temperature
  geom_point()+
  facet_wrap(~as.factor(format(date, "%Y"))) + #makes separate plot for each year
  scale_y_reverse() + #inverses the y-axis from zero to max(depth)
  scale_color_gradient(high=scales::muted("red"), #red = high P concentration
                        low=scales::muted("blue"), #blue = low P concentration
                        name="Total P (ug/L)") + 
  geom_hline(yintercept=8, lty="dashed") + #add in horizontal line for epi max
  theme_classic() +
  labs(x="Date", y="Depth (m)")  #add axis labels
SI_fig19 #print plot
ggsave(SI_fig19, file="Products/SI_figures/Figure19_color.png") #save plot
```

P concentration interpolated across depth.

b) binning P observations to expected frequency...

```{r}
### define expected observation frequency ###
Yr<- seq(from=2013, #define year start of study period
             to=2018, #define year end of study period
             by=1) #creates a sequence every 1 year
Yr #prints the expected years of data
daynum_bins<- seq(from=196, #define date start of CyanoHAB season
                  to=252, #define date end of CyanoHAB season
                  by=14) #creates a sequence every 14 days 
daynum_bins #prints the expected observation dates of data

### add fields with day and year numbers ###
TotP_interp_depth <- TotP_interp_depth %>%
  mutate(daynum=as.numeric(format(date, "%j")), #add field with day num
         Yr=as.numeric(format(date, "%Y"))) #add field with yr

### bin observations to expected frequency ###
TotP_interp_depth <- TotP_interp_depth  %>% 
mutate(daynum_bin=
         ifelse(daynum<=(daynum_bins[2]-7), #if observation is <= bin 2 min
                daynum_bins[1], #yes returns bin 1
                ifelse(daynum>(daynum_bins[2]-7)&daynum<=(daynum_bins[3]-7),
              #if observation is > bin 2 min and <= bin 3 min
              daynum_bins[2], #yes returns bin 2
              ifelse(daynum>(daynum_bins[3]-7)&daynum<=(daynum_bins[4]-7),
                     #if observation is > bin 3 min and <= bin 4 min
                     daynum_bins[3], #yes returns bin 3
                     ifelse(daynum>(daynum_bins[4]-7)&daynum<=(daynum_bins[5]-7),
                            #if observation is > bin 4 min and <= bin 5 min 
                            daynum_bins[4], #yes returns bin 4
                            daynum_bins[5])#no returns bin 5
                     )
              )
       ) 
)

### check observations rebinned ###
ggplot(data=TotP_interp_depth, #create limno plot with depth inverse on y-axis
       aes(x=daynum_bin, #daynum as x-axis
           y=Depth_m, #depth as y-axis
           color=TotP_mgL)) + #and data points colored by temperature
  geom_point()+
  facet_wrap(~as.factor(Yr)) + #makes separate plot for each year
  scale_y_reverse() + #inverses the y-axis from zero to max(depth)
  scale_color_gradient(high=scales::muted("red"), #red = high P concentration
                        low=scales::muted("blue"), #blue = low P concentration
                        name="Total P (ug/L)") + 
  geom_hline(yintercept=8, lty="dashed") + #add in horizontal line for epi max
  theme_classic() +
  labs(x="Day of Year", y="Depth (m)")  #add axis labels
```

 P measurements at equal intervals. 

c) binning change in epi volume to expected frequency... 

```{r}
### bin observations to expected frequency ###
Vol_epi <- Vol_epi  %>% 
mutate(daynum_bin=
         ifelse(daynum<=(daynum_bins[2]-7), #if observation is <= bin 2 min
                daynum_bins[1], #yes returns bin 1
                ifelse(daynum>(daynum_bins[2]-7)&daynum<=(daynum_bins[3]-7),
              #if observation is > bin 2 min and <= bin 3 min
              daynum_bins[2], #yes returns bin 2
              ifelse(daynum>(daynum_bins[3]-7)&daynum<=(daynum_bins[4]-7),
                     #if observation is > bin 3 min and <= bin 4 min
                     daynum_bins[3], #yes returns bin 3
                     ifelse(daynum>(daynum_bins[4]-7)&daynum<=(daynum_bins[5]-7),
                            #if observation is > bin 4 min and <= bin 5 min 
                            daynum_bins[4], #yes returns bin 4
                            daynum_bins[5])#no returns bin 5
                     )
              )
       ) 
)
Vol_epi<- Vol_epi %>% 
  filter(!duplicated(cbind(Yr, daynum_bin), fromLast = TRUE)) #remove duplicate
```

Epi volume change binned to expected frequency. 

d) binning P_hypo:epi to expected frequencies...

```{r}
### add Yr and daynum to data frame ###
P_ratio<- P_ratio %>%
  mutate(Yr=as.integer(format(sampledate, "%Y")),
         daynum=as.integer(format(sampledate, "%j")))

### bin observations to expected frequency ###
P_ratio <- P_ratio  %>% 
mutate(daynum_bin=
         ifelse(daynum<=(daynum_bins[2]-7), #if observation is <= bin 2 min
                daynum_bins[1], #yes returns bin 1
                ifelse(daynum>(daynum_bins[2]-7)&daynum<=(daynum_bins[3]-7),
              #if observation is > bin 2 min and <= bin 3 min
              daynum_bins[2], #yes returns bin 2
              ifelse(daynum>(daynum_bins[3]-7)&daynum<=(daynum_bins[4]-7),
                     #if observation is > bin 3 min and <= bin 4 min
                     daynum_bins[3], #yes returns bin 3
                     ifelse(daynum>(daynum_bins[4]-7)&daynum<=(daynum_bins[5]-7),
                            #if observation is > bin 4 min and <= bin 5 min 
                            daynum_bins[4], #yes returns bin 4
                            daynum_bins[5])#no returns bin 5
                     )
              )
       ) 
)
```

P_hypo:epi binned to expected frequencies.

# Step 6. Merge the data frames: 

```{r}
### merging TotP_interpolated by depth and Volume change ###
P_vol<- merge(TotP_interp_depth, #joins x data frame
             Vol_epi, #joins y data frame
             by.x=c('Yr', 'daynum_bin', 'Depth_m'), #using these key fields in x
             by.y=c('Yr', 'daynum_bin', 'Depth_m'),#using these key fields in y
             all=TRUE) 
```

# Step 7. Save the new data files:

```{r}
### save the P concentration and epi volume change ###
write.csv(P_vol, file="Cleaned_data/TotalP_water.csv")

### save the P_hypo:epi ###
write.csv(P_ratio, file="Cleaned_data/TotalP_ratio.csv")
```

Data saved.