---
title: "TotalP_internal.Rmd"
author: "Lauren A. Knose, ORISE-EPA"
date: '2024-06-01'
output: html_document
editor_options: 
  chunk_output_type: console
---

The purpose of this program is to calculate the internal legacy P loading for Lake Mendota, WI.

# Step 1. Load dependent packages and data needed:

a) loading dependent packages...

```{r}
library(ggplot2) #needed for plots
library(dplyr) #needed for reformating data
library(zoo) #needed to linearly interpolate missing values
library(reshape2) #needed to reshape data from wide to long
```

Packages loaded.

b) loading dependent data...

```{r}
P_vol<- read.csv(file="Cleaned_data/TotalP_water.csv") 
P_ratio<- read.csv(file="Cleaned_data/TotalP_ratio.csv")
```

# Step 2. Calculate daily internal P loads:

a) calculating Pstrat...

```{r}
### define unit conversions ###
L_m3<- 1000 #unit conversion from L to m^3
mg_kg<- 1/1000000 #unit conversion from mg to kg
t<- 14 #units of time between observations in days

## fill missing values ###
P_vol2<- P_vol %>%
  group_by(Yr) %>% #group by year
  mutate(TotP_mgL= na.approx(TotP_mgL, na.rm=FALSE)) %>% #interpolate between obs
  fill(TotP_mgL, .direction="up") %>% #fill remaining NAs with next value
  ungroup()

### calculate change in P mass between observations ###
Pint<- as.data.frame(P_vol2) %>%
  mutate(Vol_add_epi_m3=ifelse(Volchg_epi_m3>0, Volchg_epi_m3, 0),
         Pstrat_kg=(TotP_mgL*L_m3*mg_kg*Vol_add_epi_m3), #Pmass change
         Strat_TotP_kgday=Pstrat_kg/t) #Pint in kg/day
```

Pstrat calculated.

b) calculating Punstr...

```{r}
### define units and conversion coefficients ###
mg_kg<- 1/1000000 #conversion factor from mg to kg
A_unstrat<- 13980000 #unstratified lake surface area, where sediments are 
# directly adjacent to epilimnion in m^2
Sed_P<- 2.4 # mg P/m2-day (average based on previous study by Stauffer 1987 who 
# determined the average sediment release of P from sediments in contact with 
#the epilimnion in Lake Mendota (original range was 1.2 to 3.5 mg P /m2-day), measured by sediment cores.

### calculate P loads ###
Pint<- Pint %>%
  mutate(Unstr_TotP_kgday=(Sed_P*A_unstrat*mg_kg), #calculates internal P loading
        #from unstratified area
        Int_TotP_kgday= Unstr_TotP_kgday+Strat_TotP_kgday) #calculates total internal
#legacy P loading)
```

Punstr calculated.

# Step 3. Merge data files with P_int and P_hypo:epi: 

```{r}
### remove unnecessary fields ###
Pint2<- Pint %>% select(c(Yr, daynum.x, sampledate, Strat_TotP_kgday, Unstr_TotP_kgday, Int_TotP_kgday))

P_ratio2<- P_ratio %>% select(c(Yr, daynum, sampledate, TotP_HEratio, epi_totP_mgL))

### merge observations with data frame of frequencies ###
Int_P_all<- merge(Pint2, P_ratio2, 
                  by.x=c("Yr","daynum.x"), 
                  by.y=c("Yr", "daynum"), all.x=TRUE) 
```

Data merged.

# Step 4. Interpolate daily Pint loads:

a) determining  daily expected records...

```{r}
### determine expected observation frequencies ###
Yr<- seq(from=2013, #define year start of study period
             to=2018, #define year end of study period
             by=1) #creates a sequence every 1 year
daynum_bins<- seq(from=1, #define date start of sequence
                  to=365, #define date end of CyanoHAB season
                  by=1) #creates a sequence every 1 day 

### create data frame with the frequencies wanted ###
df<- data.frame(x=rep(Yr, each=365),y= rep(daynum_bins, times=6))
df<- df %>%
  rename(Yr=x, daynum_bin=y)
```

Daily records added.

b) merging data frames...

```{r}
### merge observations with data frame of frequencies ###
Pint_all<- merge(Int_P_all, df, 
               by.x=c("Yr","daynum.x"), 
               by.y=c("Yr", "daynum_bin"), all.y=TRUE) 
```

c) interpolating missing daily values...

```{r}
## interpolate between observations ###
Pint_all<- Pint_all %>%
  group_by(Yr) %>% #group by year
  mutate(Strat_TotP_kgday= na.approx(Strat_TotP_kgday, na.rm=FALSE, rule=2),
         Unstr_TotP_kgday=na.approx(Unstr_TotP_kgday, na.rm=FALSE, rule=2),
         Int_TotP_kgday=na.approx(Int_TotP_kgday, na.rm=FALSE, rule=2),
         TotP_HEratio=na.approx(TotP_HEratio, na.rm=FALSE, rule=2),
         epi_totP_mgL=na.approx(epi_totP_mgL, na.rm=FALSE, rule=2)) %>% 
  ungroup() %>%
  rename(daynum=daynum.x)
```

Daily observations interpolated.

# Step 4. Save the data as a new data file:

```{r}
write.csv(Pint_all, file="Cleaned_data/TotalP_allint.csv")
```

# Step 4. Summarize and view the data:

a) calculating average daily values during the CyanoHAB season...

```{r}
### calculating average daily internal P load ###
IntP_dayavg<- Pint_all %>%
  filter(daynum>=196 & daynum<=252) %>%
  summarize(mean=mean(Int_TotP_kgday, na.rm=TRUE), #calculate mean
            se=(sd(Int_TotP_kgday, na.rm=TRUE)/sqrt(length(Int_TotP_kgday))),
            min=min(Int_TotP_kgday, na.rm=TRUE),
            max=max(Int_TotP_kgday, na.rm=TRUE)
            )
IntP_dayavg

### calculating average daily P_hypo:epi ###
Pratio_dayavg<- Pint_all %>%
  filter(daynum>=196 & daynum<=252) %>%
  summarize(mean=mean(TotP_HEratio, na.rm=TRUE), #calculate mean
            se=(sd(TotP_HEratio, na.rm=TRUE)/sqrt(length(TotP_HEratio))),
            min=min(TotP_HEratio, na.rm=TRUE),
            max=max(TotP_HEratio, na.rm=TRUE)
            )
Pratio_dayavg
```

Average daily int P load calculated.

b) calculating average annual values during the CyanoHAB season...

```{r}
### calculating average annual internal P load ###
IntP_yr<- Pint_all %>%
  filter(daynum>=196 & daynum<=252) %>%
  group_by(Yr) %>%
  summarize(sum=sum(Int_TotP_kgday)) #add total P load
IntP_yr
IntP_yravg<- IntP_yr %>%
  summarize(mean=mean(sum, na.rm=TRUE), #print the average annual
            se=sd(sum, na.rm=TRUE)/sqrt(length(sum)),
            min=min(sum, na.rm=TRUE),
            max=max(sum, na.rm=TRUE)) #print sd for avg
IntP_yravg #the estimated annual average P_ext with the P_other

### calculating average P_hypo:epi for each year ###
Pratio_dayavg<- Pint_all %>%
  filter(daynum>=196 & daynum<=252) %>%
  group_by(Yr) %>%
  summarize(mean=mean(TotP_HEratio, na.rm=TRUE), #calculate mean
            se=(sd(TotP_HEratio, na.rm=TRUE)/sqrt(length(TotP_HEratio))),
            min=min(TotP_HEratio, na.rm=TRUE),
            max=max(TotP_HEratio, na.rm=TRUE)
            )
Pratio_dayavg
```

Average annual int P load calculated.

c) calculating the average daily values by source (unst vs strat)...

```{r}
IntP_dayavg_source<- Pint_all %>%
  filter(daynum>=196 & daynum<=252) %>%
  summarize(mean_unstr=mean(Unstr_TotP_kgday, na.rm=TRUE), #calculate mean
            se_unstr=(sd(Unstr_TotP_kgday, na.rm=TRUE)/sqrt(length(Unstr_TotP_kgday))),
            min_unstr=min(Unstr_TotP_kgday, na.rm=TRUE),
            max_unstr=max(Unstr_TotP_kgday, na.rm=TRUE),
            mean_strat=mean(Strat_TotP_kgday, na.rm=TRUE), #calculate mean
            se_strat=(sd(Strat_TotP_kgday, na.rm=TRUE)/sqrt(length(Strat_TotP_kgday))),
            min_strat=min(Strat_TotP_kgday, na.rm=TRUE),
            max_strat=max(Strat_TotP_kgday, na.rm=TRUE)
            )
IntP_dayavg_source
```

Average daily P_int summarized.
