---
title: "Impact_model.Rmd"
author: "Lauren A. Knose, ORISE-EPA"
date: '2024-05-08'
output: html_document
editor_options: 
  chunk_output_type: console
---

The purpose of this program is to perform an impact analysis comparing the internal vs external P loading on changes in Chl-a and CyanoDensity.


# Step 1. Load dependent packages and data needed:

a) Loading dependent packages...
```{r}
library(dplyr) #needed for reshaping/reformating data
library(corrplot) #needed for correlation matrix plot
library(ggplot2) #needed for plotting and saving plots
library(car) #needed to get model VIFs
```

b) Loading data...

```{r}
internalloads<- read.csv(file="Cleaned_data/TotalP_allint.csv")
externalloads<- read.csv(file="Cleaned_data/TotalP_allext.csv")
cyanoHABseverity<- read.csv(file="CleaneD_data/cyanoHAB_severity.csv")


cyanoHABseverity2<- cyanoHABseverity %>%
  select(Yr, daynum_bin, Chla_ugL)



HL_all<- merge(internalloads, externalloads,
                by="sampledate")
HL_all<-merge(HL_all, cyanoHABseverity2, 
             by.x=c("Yr.x","daynum_bin"), #19 observations
             by.y=c("Yr", "daynum_bin") #36 observations
             )
```


# Step 3. Perform regression:

```{r}
regka_CD<- glm(formula=log10(CD_cellsmL) ~ #Chla_t
              log10(lag(CD_cellsmL, n=7)) + #Chl-a_t-k
                log10(lag(HE_ratio_TotP_mgL, n=7)), #P_ext,t-k
              data=EFL_all) #data frame
summary(regka_CD)#prints the AIC 
summary(lm(regka_CD))#prints the R2 and adjusted R2
aov(regka_CD) #prints the sum of squares for the regression
vif(regka_CD) #prints the VIFs 
```

b) Running second model for Chl-a...
Model: log10(Chla_t) ~ log10(Chla_t-k) + log10(P_ext,t-k) + P_int,t-k + error

```{r}
regkb_Chla<- glm(formula=log10(Chla_ugL) ~ #Chla_t
              log10(lag(Chla_ugL, n=1)) + #Chla_t-k
                log10(lag(TOTP_kg_All, n=1)) + #P_ext,t-k
                log10(lag(intload_P_kgday, n=1)), #P_int,t-k
              data=HL_all) #data frame
summary(regkb_Chla)#prints the glm output
summary(lm(regkb_Chla))#prints the R2 and adjusted R2
aov(regkb_Chla)#prints the sum of squares for the regression and residuals
vif(regkb_Chla) #prints the VIFs 
png(file="Products/SI_figures/Figure11.png", width=750, height=500) #open png file
layout(matrix(c(1,2,3,4),2,2)) #layout 4 graphs/page
fig11<-plot(regkb_Chla) #creates plot
dev.off() #closes the file
par(mfrow=c(1,1))
```

c) Running third model for Chl-a...
Model: log10(Chla_t) ~ log10(Chla_t-7) + log10(P_ext,t-k) + log10(P_hypo:P_epiP,t-k) + error

```{r}
regkc_Chla<- glm(formula=log10(Chla_ugL) ~ #Chla_t
              log10(Chla_ugL_k7) + #Chla_t-k
               log10(Pext_kgday_k28) + #P_ext,t-k
                log10(HE_ratio_k28), #P_hypo:P_epiP,t-k
              data=Me_complete) #data frame
summary(regkc_Chla)#prints the glm output
summary(lm(regkc_Chla))#prints the R2 and adjusted R2
aov(regkc_Chla)#prints the sum of squares for the regression and residuals
vif(regkc_Chla) #prints the VIFs 
png(file="Products/SI_figures/Figure12.png", width=750, height=500) #open png file
layout(matrix(c(1,2,3,4),2,2)) #layout 4 graphs/page
fig12<-plot(regkc_Chla) #creates plot
dev.off() #closes the file
par(mfrow=c(1,1))
```

d) Running first model for CD...
Model: log10(CD_t) ~ log10(CD_t-k) + log10(P_ext,t-k) + error

```{r}
regka_CD<- glm(formula=log10(CD_cellsmL) ~ #CD_t
              log10(CD_cellsmL_k7) + #CD_t-k
                log10(Pext_kgday_k28), #P_ext,t-k
              data=Me_complete) #data frame
summary(regka_CD)#prints the glm output
summary(lm(regka_CD))#prints the R2 and adjusted R2
aov(regka_CD)#prints the sum of squares for the regression and residuals
vif(regka_CD) #prints the VIFs 
png(file="Products/SI_figures/Figure13.png", width=750, height=500) #open png file
layout(matrix(c(1,2,3,4),2,2)) #layout 4 graphs/page
fig13<-plot(regka_CD) #creates plot
dev.off() #closes the file
par(mfrow=c(1,1))
```

e) Running second model for CD...
Model: log10(CD_t) ~ log10(CD_t-k) + log10(P_ext,t-k) + P_int,t-k + error

```{r}
regkb_CD<- glm(formula=log10(CD_cellsmL) ~ #CD_t
              log10(CD_cellsmL_k7) + #CD_t-k
                log10(Pext_kgday_k28) + #P_ext,t-k
                Int_TotP_kgday_k28, #P_int,t-k
              data=Me_complete) #data frame
summary(regkb_CD)#prints the glm output
summary(lm(regkb_CD))#prints the R2 and adjusted R2
aov(regkb_CD)#prints the sum of squares for the regression and residuals
vif(regkb_CD) #prints the VIFs 
png(file="Products/SI_figures/Figure14.png", width=750, height=500) #open png file
layout(matrix(c(1,2,3,4),2,2)) #layout 4 graphs/page
fig14<-plot(regkb_CD) #creates plot
dev.off() #closes the file
par(mfrow=c(1,1))
```

f) Running third model for CD...
Model: CD_t ~ CD_t-7 + P_ext,t-k + P_hypo:P_epiP,t-k + error

```{r}
regkc_CD<- glm(formula=log10(CD_cellsmL) ~ #CD_t
              log10(CD_cellsmL_k7) + #CD_t-k
                log10(Pext_kgday_k28) + #P_ext,t-k
                log10(HE_ratio_k28), #P_hypo:P_epiP,t-k
              data=Me_complete) #data frame
summary(regkc_CD)#prints the glm output
summary(lm(regkc_CD))#prints the R2 and adjusted R2
aov(regkc_CD)#prints the sum of squares for the regression and residuals
vif(regkc_CD) #prints the VIFs 
png(file="Products/SI_figures/Figure15.png", width=750, height=500) #open png file
layout(matrix(c(1,2,3,4),2,2)) #layout 4 graphs/page
fig15<-plot(regkc_CD) #creates plot
dev.off() #closes the file
par(mfrow=c(1,1))
```

Pick the model with the lowest AIC, Residual standard error, and each term is significant (t-score p-value < 0.05) as the best model.

# Step 4. Using the best model, calculate the variance for each term for each 
year:

a) calculating source-attributed impact for Chla...
```{r}
YRregk_Chla_2012<- aov(formula=log10(Chla_ugL) ~ #Chla_t
              log10(Chla_ugL_k7) + #Chla_t-k
                log10(Pext_kgday_k28) + #P_ext,t-k
                Int_TotP_kgday_k28, #P_int,t-k
              data=subset(Me_complete, Yr==2012)) #subset data by year
Chla_SS_2012<- summary(YRregk_Chla_2012)#prints the sum of squares for the regression 
Chla_impact2012<- data.frame(Chla_SS_2012[[1]])[,2,drop=FALSE] #pulls out the SS as table

YRregk_Chla_2013<- aov(formula=log10(Chla_ugL) ~ #Chla_t
              log10(Chla_ugL_k7) + #Chla_t-k
                log10(Pext_kgday_k28) + #P_ext,t-k
                Int_TotP_kgday_k28, #P_int,t-k
              data=subset(Me_complete, Yr==2013)) #subset data by year
Chla_SS_2013<- summary(YRregk_Chla_2013)#prints the sum of squares for the regression 
Chla_impact2013<- data.frame(Chla_SS_2013[[1]])[,2,drop=FALSE] #pulls out the SS as table

YRregk_Chla_2014<- aov(formula=log10(Chla_ugL) ~ #Chla_t
              log10(Chla_ugL_k7) + #Chla_t-k
                log10(Pext_kgday_k28) + #P_ext,t-k
                Int_TotP_kgday_k28, #P_int,t-k
              data=subset(Me_complete, Yr==2014)) #subset data by year
Chla_SS_2014<- summary(YRregk_Chla_2014)#prints the sum of squares for the regression 
Chla_impact2014<- data.frame(Chla_SS_2014[[1]])[,2,drop=FALSE] #pulls out the SS as table

YRregk_Chla_2015<- aov(formula=log10(Chla_ugL) ~ #Chla_t
              log10(Chla_ugL_k7) + #Chla_t-k
                log10(Pext_kgday_k28) + #P_ext,t-k
                Int_TotP_kgday_k28, #P_int,t-k
              data=subset(Me_complete, Yr==2015)) #subset data by year
Chla_SS_2015<- summary(YRregk_Chla_2015)#prints the sum of squares for the regression 
Chla_impact2015<- data.frame(Chla_SS_2015[[1]])[,2,drop=FALSE] #pulls out the SS as table

YRregk_Chla_2016<- aov(formula=log10(Chla_ugL) ~ #Chla_t
              log10(Chla_ugL_k7) + #Chla_t-k
                log10(Pext_kgday_k28) + #P_ext,t-k
                Int_TotP_kgday_k28, #P_int,t-k
              data=subset(Me_complete, Yr==2016)) #subset data by year
Chla_SS_2016<- summary(YRregk_Chla_2016)#prints the sum of squares for the regression 
Chla_impact2016<- data.frame(Chla_SS_2016[[1]])[,2,drop=FALSE] #pulls out the SS as table

YRregk_Chla_2017<- aov(formula=log10(Chla_ugL) ~ #Chla_t
              log10(Chla_ugL_k7) + #Chla_t-k
                log10(Pext_kgday_k28) + #P_ext,t-k
                Int_TotP_kgday_k28, #P_int,t-k
              data=subset(Me_complete, Yr==2017)) #subset data by year
Chla_SS_2017<- summary(YRregk_Chla_2017)#prints the sum of squares for the regression 
Chla_impact2017<- data.frame(Chla_SS_2017[[1]])[,2,drop=FALSE] #pulls out the SS as table

YRregk_Chla_2018<- aov(formula=log10(Chla_ugL) ~ #Chla_t
              log10(Chla_ugL_k7) + #Chla_t-k
                log10(Pext_kgday_k28) + #P_ext,t-k
                Int_TotP_kgday_k28, #P_int,t-k
              data=subset(Me_complete, Yr==2018)) #subset data by year
Chla_SS_2018<- summary(YRregk_Chla_2018)#prints the sum of squares for the regression 
Chla_impact2018<- data.frame(Chla_SS_2018[[1]])[,2,drop=FALSE] #pulls out the SS as table

Model_output_Chla<- cbind(Chla_impact2012, Chla_impact2013, Chla_impact2014, Chla_impact2015, Chla_impact2016, Chla_impact2017, Chla_impact2018) #combines all tables together

Model_output_Chla$Sum_Sq_part<- c("Seed stock", "External load", "Internal load", "Residual") #add column with predictor name
colnames(Model_output_Chla)<- c("2012", "2013", "2014", "2015", "2016", "2017", "2018", "Sum_Sq_part")
# renames columns with year for sum squares
write.csv(Model_output_Chla, file="Products/Figure7_data.csv")
```

Yearly source-attributed impact outputted.

b) calculating source-attributed impact for CD...

```{r}
regka_CD<- glm(formula=log10(CD_cellsmL) ~ #Chla_t
              log10(lag(CD_cellsmL, n=7)) + #Chl-a_t-k
                log10(lag(HE_ratio_TotP_mgL, n=7)), #P_ext,t-k
              data=EFL_all) #data frame
unique(EFL_all$Yr)


YRregk_CD_2015<- aov(formula=log10(CD_cellsmL) ~ #CD_t
              log10(lag(CD_cellsmL, n=7)) + #CD_t-k
                log10(lag(HE_ratio_TotP_mgL, n=7)), #P_int,t-k
              data=subset(EFL_all, Yr==2015) #subset data by year
            )
SS_2015<- summary(YRregk_CD_2015)#prints the sum of squares for the regression 
impact2015<- data.frame(SS_2015[[1]])[,2,drop=FALSE] #pulls out the SS as table

YRregk_CD_2016<- aov(formula=log10(CD_cellsmL) ~ #CD_t
              log10(lag(CD_cellsmL, n=7)) + #CD_t-k
                log10(lag(HE_ratio_TotP_mgL, n=7)), #P_int,t-k
              data=subset(EFL_all, Yr==2016) #subset data by year
            )
SS_2016<- summary(YRregk_CD_2016)#prints the sum of squares for the regression 
impact2016<- data.frame(SS_2016[[1]])[,2,drop=FALSE] #pulls out the SS as table

YRregk_CD_2017<- aov(formula=log10(CD_cellsmL) ~ #CD_t
              log10(lag(CD_cellsmL, n=7)) + #CD_t-k
                log10(lag(HE_ratio_TotP_mgL, n=7)), #P_int,t-k
              data=subset(EFL_all, Yr==2017) #subset data by year
            )
SS_2017<- summary(YRregk_CD_2017)#prints the sum of squares for the regression 
impact2017<- data.frame(SS_2017[[1]])[,2,drop=FALSE] #pulls out the SS as table
Model_output_CD<- cbind(impact2015, impact2016, impact2017) #combines all tables together
Model_output_CD$Sum_Sq_part<- c("Seed stock", "Internal load potential", "Residual") #add column with predictor name
colnames(Model_output_CD)<- c("2015", "2016", "2017","Sum_Sq_part")
# renames columns with year for sum squares
write.csv(Model_output_CD, file="Products/Figure_EFL_data.csv")
```

```{r}
YRregk_CD_2013<- aov(formula=log10(CD_cellsmL) ~ #CD_t
              log10(CD_cellsmL_k7) + #CD_t-k
                log10(Pext_kgday_k28) + #P_ext,t-k
                Int_TotP_kgday_k28, #P_int,t-k
              data=subset(Me_complete, Yr==2013) #subset data by year
            )
SS_2013<- summary(YRregk_CD_2013)#prints the sum of squares for the regression 
impact2013<- data.frame(SS_2013[[1]])[,2,drop=FALSE] #pulls out the SS as table

YRregk_CD_2014<- aov(formula=log10(CD_cellsmL) ~ #CD_t
              log10(CD_cellsmL_k7) + #CD_t-k
                log10(Pext_kgday_k28) + #P_ext,t-k
                Int_TotP_kgday_k28, #P_int,t-k
              data=subset(Me_complete, Yr==2014) #subset data by year
            )
SS_2014<- summary(YRregk_CD_2014)#prints the sum of squares for the regression 
impact2014<- data.frame(SS_2014[[1]])[,2,drop=FALSE] #pulls out the SS as table

YRregk_CD_2015<- aov(formula=log10(CD_cellsmL) ~ #CD_t
              log10(CD_cellsmL_k7) + #CD_t-k
                log10(Pext_kgday_k28) + #P_ext,t-k
                Int_TotP_kgday_k28, #P_int,t-k
              data=subset(Me_complete, Yr==2015) #subset data by year
            )
SS_2015<- summary(YRregk_CD_2015)#prints the sum of squares for the regression 
impact2015<- data.frame(SS_2015[[1]])[,2,drop=FALSE] #pulls out the SS as table

YRregk_CD_2016<- aov(formula=log10(CD_cellsmL) ~ #CD_t
              log10(CD_cellsmL_k7) + #CD_t-k
                log10(Pext_kgday_k28) + #P_ext,t-k
                Int_TotP_kgday_k28, #P_int,t-k
              data=subset(Me_complete, Yr==2016) #subset data by year
            )
SS_2016<- summary(YRregk_CD_2016)#prints the sum of squares for the regression 
impact2016<- data.frame(SS_2016[[1]])[,2,drop=FALSE] #pulls out the SS as table

YRregk_CD_2017<- aov(formula=log10(CD_cellsmL) ~ #CD_t
              log10(CD_cellsmL_k7) + #CD_t-k
                log10(Pext_kgday_k28) + #P_ext,t-k
                Int_TotP_kgday_k28, #P_int,t-k
              data=subset(Me_complete, Yr==2017) #subset data by year
            )
SS_2017<- summary(YRregk_CD_2017)#prints the sum of squares for the regression 
impact2017<- data.frame(SS_2017[[1]])[,2,drop=FALSE] #pulls out the SS as table

YRregk_CD_2018<- aov(formula=log10(CD_cellsmL) ~ #CD_t
              log10(CD_cellsmL_k7) + #CD_t-k
                log10(Pext_kgday_k28) + #P_ext,t-k
                Int_TotP_kgday_k28, #P_int,t-k
              data=subset(Me_complete, Yr==2018) #subset data by year
            )
SS_2018<- summary(YRregk_CD_2018)#prints the sum of squares for the regression 
impact2018<- data.frame(SS_2018[[1]])[,2,drop=FALSE] #pulls out the SS as table

Model_output_CD<- cbind(impact2012, impact2013, impact2014, impact2015, impact2016, impact2017, impact2018) #combines all tables together
Model_output_CD$Sum_Sq_part<- c("Seed stock", "External load", "Internal load", "Residual") #add column with predictor name
colnames(Model_output_CD)<- c("2012", "2013", "2014", "2015", "2016", "2017", "2018", "Sum_Sq_part")
# renames columns with year for sum squares
write.csv(Model_output_CD, file="Products/Figure8_data.csv")
```

Yearly source-attributed impact outputted.
