---
title: "TotalP_external_all.Rmd"
author: "Lauren A. Knose, ORISE-EPA"
date: '2023-06-20'
output: html_document
editor_options: 
  chunk_output_type: console
---

The purpose of this program is to calculate total external P load from both atmospheric deposition and stream inflows.

# Step 1. Load dependent packages and data needed:

a) Loading dependent packages...

```{r}
library(ggplot2) #needed for plots
library(dplyr) #needed for reformaing data
library(reshape2) #needed for reshaping data frames from wide to long
```

Packages loaded.

b) Loading dependent data....

```{r}
streams<- read.csv(file="Cleaned_data/TotalP_inflows.csv") 
atm<- read.csv(file="Cleaned_data/TotalP_atm.csv")
```

Data loaded.

# Step 2. Define, filter, and merge data:

a) defining and filtering data...

```{r}
streams2<- streams %>%
  mutate(X=NULL, #remove index column
         sampledate=as.Date(sampledate,
                            origin="1899-12-30",
                            format="%Y-%m-%d")) %>%
  filter(Site=="All") %>%
  mutate(Pstreams_kgday=Flux) %>%#reformat date
  filter(daynum>=196 & daynum<=258) %>% #filter for the CyanoHAB season
  filter(Yr>=2012 & Yr<=2018) %>% #filter for the study period
  select(c("sampledate", "Pstreams_kgday")) #remove the extra columns
atm2<- atm %>%
  mutate(X=NULL, #remove index column
         sampledate=as.Date(sampledate, origin="1899-12-30"),
         Patm_kgday=atm_TotP_kgday)%>% #reformat date
  filter(daynum>=196 & daynum<=258) %>% #filter for the CyanoHAB season
  filter(Yr>=2012 & Yr<=2018) %>% #filter for the study period
  select(c("sampledate", "Patm_kgday")) #remove the extra columns
```

Data defined.

b) merging data...

```{r}
extP_load<- merge(atm2, streams2, by=c("sampledate")) #adds two data files together
extP_load <- extP_load %>%
  mutate(Pext_kgday_noother=(Patm_kgday + Pstreams_kgday), #combine atm P and stream P
         Pother_kgday=(Pext_kgday_noother*.47), #assign new column for P_other with 47% of total
         Pext_kgday=Pext_kgday_noother*1.47) #increase P_ext 47% to account for P_other
```

External P loading merged.

# Step 3. Save data as new data file:

```{r}
write.csv(extP_load, file="Cleaned_data/TotalP_allext.csv")
```

New data file created with total external P loading.

# Step 4. Summarize and view data:

a) plot total P load for CyanoHAB season by source for each year...

```{r}
extP_load_long<- melt(extP_load, #the table you want to reshape wide to long
               id.vars=c("sampledate"), #the column to use as unique id
               measure.vars=c("Patm_kgday","Pstreams_kgday", "Pother_kgday", "Pext_kgday_noother","Pext_kgday"), 
               #the columns where values will be taken
               variable.name="Source", #label the source of P loading
               value.name="Pext_kgday" #label the value with units
               )
extP_load_long$Source<-as.character(extP_load_long$Source) #tell R character data held
extP_load_long<- extP_load_long %>%
  filter(Source==c("Patm_kgday", "Pstreams_kgday", "Pother_kgday")) #remove totals here first)
unique(extP_load_long$Source) #check totals were removed
extP_load_long$Source[extP_load_long$Source=="Patm_kgday"] <- "atmosphere"
extP_load_long$Source[extP_load_long$Source=="Pstreams_kgday"] <- "streams"
extP_load_long$Source[extP_load_long$Source=="Pother_kgday"] <- "other"

extP_load_long<- extP_load_long %>%
  mutate(Yr=as.character(format(sampledate, "%Y")))%>%  #add year back in data frame
  group_by(Yr, Source) %>% #for each year and each source
  summarize(TotPext_kgday=sum(Pext_kgday)) %>% #sum up all source Pext_kgday 
  ungroup()
  
write.csv(extP_load_long, file="Products/Figure9b_data.csv") #saves data table
fig9b<- ggplot(extP_load_long, aes(x=Yr, y=TotPext_kgday, 
                             fill=Source)) + #separate sources by fill color
  geom_bar(position="stack", stat="identity", color="black") + #stacked bar graph
  scale_fill_manual(values=c("grey", "white", "black")) + #atm = grey, streams=black, other=white
  theme_classic(base_size=12) +
  labs(x="Year", y="External total P loading (kg/yr)")
fig9b #print plot
ggsave(fig9b, file="Products/Figure9b.png")
```

Total P load by source for each year plotted.

b) calculating average daily and average annual Other total P loading during the CyanoHAB season...

```{r}
Other_avg<- extP_load %>%
  summarise(mean=mean(Pother_kgday, #average daily total P loading
                      na.rm=TRUE),
            se=(sd(Pother_kgday, na.rm=TRUE)/sqrt(length(Pother_kgday))),
            min=min(Pother_kgday, na.rm=TRUE),
            max=max(Pother_kgday, na.rm=TRUE))
Other_avg
Other_sumavg<- extP_load %>%
  mutate(Yr=format(sampledate, "%Y")) %>% #add back in Yr to group data
  select(Yr, Pother_kgday) %>% #select the fields needed
  group_by(Yr) %>% #for every year
  summarize(sum=sum(Pother_kgday)) #add total P load
Other_sumavg
mean(Other_sumavg$sum) # print average annual total P loading from other sources
sd(Other_sumavg$sum)/sqrt(length(Other_sumavg$sum)) #print sd for avg
```

c) calculate average daily total P load for CyanoHAB season...

```{r}
Pext_noother_day<- extP_load %>%
  summarize(mean=mean(Pext_kgday_noother, na.rm=TRUE), #calculate mean
            se=(sd(Pext_kgday_noother, na.rm=TRUE)/sqrt(length(Pext_kgday_noother))),
            min=min(Pext_kgday_noother, na.rm=TRUE),
            max=max(Pext_kgday_noother, na.rm=TRUE)
            )
Pext_noother_day # the estimated daily average P_ext without the P_other

Pext_day<- extP_load %>%
  summarize(mean=mean(Pext_kgday, na.rm=TRUE), #calculate mean
            se=(sd(Pext_kgday, na.rm=TRUE)/sqrt(length(Pext_kgday))),
            min=min(Pext_kgday, na.rm=TRUE),
            max=max(Pext_kgday, na.rm=TRUE)
            )
Pext_day # the estimated daily average  P_ext with P_other
```

Daily average P_ext calculated.

d) calculating average annual total P load for CyanoHAB season...

```{r}
Pext_noother_yr<- extP_load %>%
  mutate(Yr=format(sampledate, "%Y")) %>% #add back in Yr to group data
  group_by(Yr) %>%
  summarize(sum=sum(Pext_kgday_noother))
Pext_noother_yr<- Pext_noother_yr %>%
  summarize(mean=mean(Pext_noother_yr$sum), #print the average annual
            se=sd(Pext_noother_yr$sum)/sqrt(length(Pext_noother_yr$sum))) #print sd for avg
Pext_noother_yr #the estimated annual average P_ext without the P_other

Pext_yr<- extP_load %>%
  mutate(Yr=format(sampledate, "%Y")) %>% #add back in Yr to group data
  group_by(Yr) %>%
  summarize(sum=sum(Pext_kgday))
Pext_yr<- Pext_yr %>%
  summarize(mean=mean(Pext_yr$sum), #print the average annual
            se=sd(Pext_yr$sum)/sqrt(length(Pext_yr$sum))) #print sd for avg
Pext_yr #the estimated annual average P_ext with the P_other
```

Annual average P_ext calculated.