---
title: "TotalP_internal.Rmd"
author: "Lauren A. Knose, ORISE-EPA"
date: '2023-08-25'
output: html_document
editor_options: 
  chunk_output_type: console
---

The purpose of this program is to calculate the internal legacy P loading for Lake
Mendota, WI, from entrainment. Internal Legacy Total P Load (kg-day-1) = [P_t]* (Vol_epi_t - Vol_epi,t-1)/t.

# Step 1. Load dependent packages and data needed:

a) loading dependent packages...

```{r}
library(ggplot2) #needed for plots
library(dplyr) #needed for reformating data
library(zoo) #needed to linearly interpolate missing values
library(tidyverse) #needed for crossing function
library(reshape2) #needed to reshape data from wide to long
```

Packages loaded.

b) loading dependent data...

```{r}
waterP<- read.csv(file="Original_data/ntl1_v9_1_chem.csv", header=TRUE) 
# original water chemistry data file which has P concentrations
EpiVol<- read.csv(file="Cleaned_data/EpiVolume.csv", header=TRUE)
# output data file with change in epi volume calculations
```

# Step 2. Define and filter the data:

a) defining and filtering the P concentration data...

```{r}
waterP2<- waterP %>%
  mutate(sampledate=as.Date(sampledate, origin="1899-12-30"), #format as Date
         TotP_mgL=totpuf_sloh, totpuf_sloh=NULL, #rename field with units and remove duplicate column
         DisP_mgL=drp_sloh, drp_sloh=NULL, #rename field with units and remove duplicate column
         Depth_m = as.integer(round(depth, 0)), depth=NULL, #round depth to nearest integer, rename field with units, remove duplicate column
         daynum=as.integer(format(sampledate, "%j")), #add field with day 
         Yr=as.integer(format(sampledate, "%Y"))) %>% #add field with year
  filter(lakeid=="ME", #filter for Lake Mendota
         rep=="1") %>% #filter for rep 1 data only (exclude analytical reps)
  filter(!is.na(TotP_mgL)) %>% #remove NA values
  filter(Yr>=2012 & Yr<=2018) %>% #filter for study years
  filter(daynum >=189 & daynum <=259) %>% #filter for CyanoHAB season +- 7 days
  select(sampledate, TotP_mgL, Depth_m, daynum, Yr) #select fields needed
```

P concentration data defined and filtered. 

b) defining and filtering the change in epi volume data...

```{r}
EpiVol2<- EpiVol %>%
  mutate(sampledate=as.Date(date, origin="1899-12-30"),#reformat date
         daynum=as.integer(format(sampledate, "%j")),#add field with day
         Yr=as.integer(format(sampledate, "%Y")), #add field with year
         Depth_m=as.integer(round(Thermo_depth_m, 0)), Thermo_depth_m=NULL) %>% #round the depth to nearest integer, rename field with units, remove duplicate column
  filter(Yr>=2012 & Yr<=2018) %>% #filter for study period
  filter(daynum>=189 & daynum <=259) %>% #filter for CyanoHAB season +- 7 days
  select(sampledate, Vol_epi_m3, Depth_m, Yr, daynum) #select fields needed  
```

Thermocline data filtered.

# Step 3. View the data:

a) plotting data available on total P concentration...

```{r}
ggplot(data=waterP2, #creates plot
       aes(x=daynum, #sampled date as x-axis
           y=Depth_m, #depth as y-axis
           color=TotP_mgL)) + #data points colored by temperature
  geom_point() + #plot point graph
  facet_wrap(~Yr) + #make separate graph for each year
  scale_y_reverse() + #inverses the y-axis from zero to max(depth) for limno plot
  scale_color_gradient(high=scales::muted("red"), #red = hot temp
                        low=scales::muted("blue"), #blue = cold temp
                        name="Total P (ug/L)") + #name legend
  geom_hline(yintercept=8, lty="dashed") + #add in horizontal line for epi max
  theme_classic() +
  labs(x="Date", y="Depth (m)", #add axis labels
  caption="Horizontal dashed line indicates epilimnion maximum used for sampling at 8 m.") 
```

Total P concentration data plotted. Note, P concentration at depth only available
monthly, but P concentration in epilimnion available biweekly. Missing observations
will need to be interpolated.

b) plotting data available on thermocline depth...

```{r}
ggplot(data=EpiVol2, #create plot
       aes(x=daynum, #sampled date as x-axis
           y=Vol_epi_m3)) + #depth as y-axis
  geom_point() + #plot point graph
  geom_line() + #add lines beween points
  facet_wrap(~Yr) + #make separate graph for each year
  geom_hline(yintercept=0, lty="dashed") + #add in horizontal line for zero change
  theme_classic() +
  labs(x="Date", y="Change in Epilimnion Volume (m^3)", #add axis labels 
       caption="Horizontal dashed line indicates zero net change.")
```

Change in epilimnion volume plotted. 

# Step 4. Interpolate missing P concentrations across depth:

a) creating a function that will interpolate across depth for each sample date...

```{r}
estimate_TotP_by_date <- function(target_date, target_depth) { #write function that
  data_for_date <- waterP2 %>% #creates a new data frame for a single date
    filter(sampledate == target_date) %>% #filters for specific date
    arrange(Depth_m) #and arranges depths min to max
  approx(data_for_date$Depth_m, #linear regression interpolation of
         data_for_date$TotP_mgL, #y values
         xout = target_depth)$y #from depth min to max
}
```

Function created.

b) filtering out any dates that have < 2 values (at least 2 values needed for 
interpolation)...

```{r}
waterP2<- waterP2 %>%
  group_by(sampledate) %>% #for each sampledate
  filter(!n()<2) %>% #remove dates with less than 2 observations
  ungroup() #ungroup table
```

Dates with not enough data removed. 

c) interpolating P concentrations across depth for each sample date...
```{r}
TotP_interp_depth <- crossing( #creates a tibble that 
  tibble(date = unique(waterP2$sampledate)), #has same dates as those in frame
  # depths can now be any value
  tibble(Depth_m = seq(0, 25, #define depth min and max
                       length.out = 26)) #set the number of interpolated values
) %>%
  group_by(date) %>%
  mutate(TotP_mgL = estimate_TotP_by_date(date[1], Depth_m)) %>%
  ungroup()
```

Data interpolated.

d) viewing interpolated data...
```{r}
ggplot(data=TotP_interp_depth, #create limno plot with depth inverse on y-axis
       aes(x=as.integer(format(date, "%j")), #sample date as x-axis
           y=Depth_m, #depth as y-axis
           color=TotP_mgL)) + #and data points colored by temperature
  geom_point()+
  facet_wrap(~as.factor(format(date, "%Y"))) + #makes separate plot for each year
  scale_y_reverse() + #inverses the y-axis from zero to max(depth)
  scale_color_gradient(high=scales::muted("red"), #red = high P concentration
                        low=scales::muted("blue"), #blue = low P concentration
                        name="Total P (ug/L)") + 
  geom_hline(yintercept=8, lty="dashed") + #add in horizontal line for epi max
  theme_classic() +
  labs(x="Date", y="Depth (m)", #add axis labels
  caption="Horizontal dashed line indicates epilimnion maximum used for sampling at 8 m.") 
```

Total P interpolated across depth for each sample date.

# Step 5. Bin P observations into same day number each year as Epi Vol:

a) defining expected observation frequencies...
```{r}
Yr<- seq(from=2012, #define year start of study period
             to=2018, #define year end of study period
             by=1) #creates a sequence every 1 year
Yr #prints the expected years of data
daynum_bins<- seq(from=196, #define date start of CyanoHAB season
                  to=252, #define date end of CyanoHAB season
                  by=14) #creates a sequence every 14 days 
daynum_bins #prints the expected observation dates of data
```

Defined expected observation frequency.

b) binning observed dates into expected dates...

```{r}
TotP_interp_depth <- TotP_interp_depth %>%
  mutate(daynum=as.numeric(format(date, "%j")), #add field with day num
         Yr=as.numeric(format(date, "%Y"))) #add field with yr

TotP_interp_depth <- TotP_interp_depth  %>% 
mutate(daynum_bin=
         ifelse(daynum<(daynum_bins[2]-7), #if observation is <= bin 2 min
                daynum_bins[1], #yes returns bin 1
                ifelse(daynum>=(daynum_bins[2]-7)&daynum<(daynum_bins[3]-7),
              #if observation is > bin 2 min and <= bin 3 min
              daynum_bins[2], #yes returns bin 2
              ifelse(daynum>=(daynum_bins[3]-7)&daynum<(daynum_bins[4]-7),
                     #if observation is > bin 3 min and <= bin 4 min
                     daynum_bins[3], #yes returns bin 3
                     ifelse(daynum>=(daynum_bins[4]-7)&daynum<(daynum_bins[5]-7),
                            #if observation is > bin 4 min and <= bin 5 min 
                            daynum_bins[4], #yes returns bin 4
                            daynum_bins[5])#no returns bin 5
                     )
              )
       ) 
)
```

Observed dates assigned into closest expected dates.  

# Step 6. Assign P concentration to thermocline depth for each day:

a) joining the two data frames by key fields...

```{r}
IntP<- merge(EpiVol2, #joins x data frame
             TotP_interp_depth, #joins y data frame
             by.x=c('Yr', 'daynum', 'Depth_m'), #using these key fields in x
             by.y=c('Yr', 'daynum_bin', 'Depth_m'), #using these key fields in y
             all.x=TRUE) #keep all records in x
```

Two data frames joined.

b) interpolating missing P concentration data...

```{r}
IntP$TotP_mgL<- na.approx(IntP$TotP_mgL) #interpolates NAs
```

P concentration data interpolated.

c) calculating the internal legacy P loading by source (entrainment vs sediments
adjacent to epilimnion), and total internal P loading.

```{r}
t<- 14 #length between observations in days
mgL_mgm3<- 1000/1 #conversion factor from mg/L to mg/m^3
mg_kg<- 1/1000000 #conversion factor from mg to kg
A_unstrat<- 13980000 #unstratified lake surface area, where sediments are 
# directly adjacent to epilimnion in m^2
Sed_P<- 2.4 # mg P/m2-day (average based on previous study by Stauffer 1987 who 
# determined the average sediment release of P from sediments in contact with 
#the epilimnion in Lake Mendota (original range was 1.2 to 3.5 mg P /m2-day), measured by sediment cores.
IntP<- IntP %>%
  mutate(Ent_TotP_kgday= (TotP_mgL*mgL_mgm3*mg_kg)*Vol_epi_m3/t, #calculates 
        #internal P load from entrainment (stratified area)
        Sed_TotP_kgday=(Sed_P*A_unstrat*mg_kg), #calculates internal P loading
        #from sediments adjacent to epilimnion (unstratified area)
        Int_TotP_kgday= Sed_TotP_kgday+Ent_TotP_kgday) #calculates total internal
#legacy P loading
```

Internal legacy P load by source and total calculated.

# Step 7. Summarize and view the data:

a) summarizing average Ent P loading during the CyanoHAB season...
```{r}
Int_yr<- IntP %>%
  group_by(Yr) %>%
  summarize(mean_entrain=mean(Ent_TotP_kgday, na.rm=TRUE),
            se_entrain=(sd(Ent_TotP_kgday, na.rm=TRUE)/sqrt(length(Ent_TotP_kgday))),
            min_entrain=min(Ent_TotP_kgday, na.rm=TRUE),
            max_entrain=max(Ent_TotP_kgday, na.rm=TRUE),
            mean_total=mean(Int_TotP_kgday, na.rm=TRUE),
            se_total=(sd(Int_TotP_kgday, na.rm=TRUE)/sqrt(length(Int_TotP_kgday))),
            min_total=min(Int_TotP_kgday, na.rm=TRUE),
            max_total=max(Int_TotP_kgday, na.rm=TRUE)
            ) #calculate annual total P load
Int_yr
```

The average annual internal legacy total P load was `r mean(Int_yr$mean_total)` +- 
`r (sd(Int_yr$se_total)/sqrt(length(Int_yr$se_total)))` kg/yr. 

b) plot total P load for CyanoHAB season by source for each year...
```{r}
IntP_long<- melt(IntP, #the table you want to reshape wide to long
               id.vars=c("sampledate"), #the column to use as unique id
               measure.vars=c("Ent_TotP_kgday","Sed_TotP_kgday"), 
               #the columns where values will be taken
               variable.name="Source", #label the source of P loading
               value.name="TotP_kgday" #label the value with units
               )
IntP_long$Source<-as.character(IntP_long$Source) #tell R character data held
IntP_long$Source[IntP_long$Source=="Ent_TotP_kgday"] <- "entrainment"
IntP_long$Source[IntP_long$Source=="Sed_TotP_kgday"] <- "adjacent sediments"

IntP_long<- IntP_long %>%
  mutate(Yr=format(sampledate, "%Y") #add year back in data frame
         ) %>%
  group_by(Yr, Source) %>% #for each year and each source
  summarize(TotP_kgday=sum(TotP_kgday)) #sum up all the TotP_kgday

ggplot(subset(IntP_long, Yr<=2018), aes(x=Yr, y=TotP_kgday, 
                             fill=Source)) + #separate sources by fill color
  geom_bar(position="stack", stat="identity", color="black") + #stacked bar graph
  scale_fill_manual(values=c("grey", "black")) + #entrain = grey, sed=black
  theme_classic(base_size=12) +
  labs(x="Year", y="Total P loading (kg/yr)") 
```

Total P load by source for each year plotted.

b) calculate average daily total P load from entrainment for CyanoHAB season...

```{r}
IntP_dayavg<- IntP %>%
  summarize(mean_ent=mean(Ent_TotP_kgday, na.rm=TRUE), #calculate mean
            se_ent=(sd(Ent_TotP_kgday, na.rm=TRUE)/sqrt(length(Ent_TotP_kgday))),
            min_ent=min(Ent_TotP_kgday, na.rm=TRUE),
            max_ent=max(Ent_TotP_kgday, na.rm=TRUE)
            )
IntP_dayavg
```

Average daily P load from entrainment calculated.

# Step 9. Save the data as a new data file:

```{r}
write.csv(IntP, file="Cleaned_data/TotalP_allint.csv")
```