---
title: "Impact_model.Rmd"
author: "Lauren A. Knose, ORISE-EPA"
date: '2023-08-30'
output: html_document
editor_options: 
  chunk_output_type: console
---

The purpose of this program is to perform an impact analysis comparing the internal vs external P loading on changes in Chl-a and CyanoDensity.


# Step 1. Load dependent packages and data needed:

a) Loading dependent packages...
```{r}
library(dplyr) #needed for reshaping/reformating data
library(corrplot) #needed for correlation matrix plot
library(ggplot2) #needed for plotting and saving plots
```

b) Loading data...

```{r}
Me_all_interp<- read.csv(file="Cleaned_data/Impact_model_input_data.csv")
```

# Step 2. Check whether variables are correlated:

```{r}
cor<- Me_all_interp %>% #select fields with responses, include columns with predictors
  select(Chla_ugL_interp, CD_cellsmL_interp, #Chla and CD
         Chla_ugL_interp_k7,CD_cellsmL_interp_k7, #Chla_t-k and CD_t-k
         ExtP_kgday_interp_k28, #P_ext,t-k
         IntP_kgday_interp_k28, #P_int,t-k
         HE_ratio_interp_k28 #P_hypo:P_epi, t-k
         ) %>%
  mutate(#shorten names and remove duplicate fields
    CD=CD_cellsmL_interp, CD_cellsmL_interp=NULL,
    Chla=Chla_ugL_interp,Chla_ugL_interp=NULL,
    Chla_k7=Chla_ugL_interp_k7, Chla_ugL_interp_k7=NULL,
    CD_k7=CD_cellsmL_interp_k7,CD_cellsmL_interp_k7=NULL,
    Pext=ExtP_kgday_interp_k28, ExtP_kgday_interp_k28=NULL,
    Pint=IntP_kgday_interp_k28, IntP_kgday_interp_k28=NULL,
    P_ratio=HE_ratio_interp_k28,HE_ratio_interp_k28=NULL) %>%
  na.omit() #remove NAs

M=cor(cor, method="pearson") #perform correlation using Kendall tau method
write.csv(M, file="Products/Figure16_data.csv") #save figure data as table
png(height=500, width=500, file="Products/Figure16_color.png", type="cairo")
corrplot(M, #plot cor matrix
                  diag=FALSE, #remove diagonal that shows corr of vars with themselves
                  outline=TRUE, #add outline around shapes to see
                  sig.level=c(0.001, 0.01, 0.05), insig='blank', #add signif stars
                  addCoef.col='black', #add corr values in black
                  type='upper')#only show upper half of matrix
dev.off()
```

# Step 3. Perform regression:

a) Running first model for Chl-a...
Model: log10(Chla_t) ~ log10(Chla_t-k) + log10(P_ext,t-k) + error

```{r}
regka_Chla<- glm(formula=log10(Chla_ugL_interp) ~ #Chla_t
              log10(Chla_ugL_interp_k7) + #Chl-a_t-k
                log10(ExtP_kgday_interp_k28), #P_ext,t-k
              data=Me_all_interp) #data frame
summary(regka_Chla)#prints the glm output
aov(regka_Chla)#prints the sum of squares for the regression and residuals
png(file="Products/Figure17.png", width=750, height=500) #open png file
layout(matrix(c(1,2,3,4),2,2)) #layout 4 graphs/page
fig17<-plot(regka_Chla) #creates plot
dev.off() #closes the file
par(mfrow=c(1,1))
```

b) Running second model for Chl-a...
Model: log10(Chla_t) ~ log10(Chla_t-k) + log10(P_ext,t-k) + P_int,t-k + error

```{r}
regkb_Chla<- glm(formula=log10(Chla_ugL_interp) ~ #Chla_t
              log10(Chla_ugL_interp_k7) + #Chla_t-k
                log10(ExtP_kgday_interp_k28) + #P_ext,t-k
                IntP_kgday_interp_k28, #P_int,t-k
              data=Me_all_interp) #data frame
summary(regkb_Chla)#prints the glm output
aov(regkb_Chla)#prints the sum of squares for the regression and residuals
png(file="Products/Figure18.png", width=750, height=500) #open png file
layout(matrix(c(1,2,3,4),2,2)) #layout 4 graphs/page
fig18<-plot(regkb_Chla) #creates plot
dev.off() #closes the file
par(mfrow=c(1,1))
```

c) Running third model for Chl-a...
Model: log10(Chla_t) ~ log10(Chla_t-7) + log10(P_ext,t-k) + log10(P_hypo:P_epiP,t-k) + error

```{r}
regkc_Chla<- glm(formula=log10(Chla_ugL_interp) ~ #Chla_t
              log10(Chla_ugL_interp_k7) + #Chla_t-k
               log10(ExtP_kgday_interp_k28) + #P_ext,t-k
                log10(HE_ratio_interp_k28), #P_hypo:P_epiP,t-k
              data=Me_all_interp) #data frame
summary(regkc_Chla)#prints the glm output
aov(regkc_Chla)#prints the sum of squares for the regression and residuals
png(file="Products/Figure19.png", width=750, height=500) #open png file
layout(matrix(c(1,2,3,4),2,2)) #layout 4 graphs/page
fig19<-plot(regkc_Chla) #creates plot
dev.off() #closes the file
par(mfrow=c(1,1))
```

d) Running first model for CD...
Model: log10(CD_t) ~ log10(CD_t-k) + log10(P_ext,t-k) + error

```{r}
regka_CD<- glm(formula=log10(CD_cellsmL_interp) ~ #CD_t
              log10(CD_cellsmL_interp_k7) + #CD_t-k
                log10(ExtP_kgday_interp_k28), #P_ext,t-k
              data=Me_all_interp) #data frame
summary(regka_CD)#prints the glm output
aov(regka_CD)#prints the sum of squares for the regression and residuals
png(file="Products/Figure20.png", width=750, height=500) #open png file
layout(matrix(c(1,2,3,4),2,2)) #layout 4 graphs/page
fig20<-plot(regka_CD) #creates plot
dev.off() #closes the file
par(mfrow=c(1,1))
```

e) Running second model for CD...
Model: log10(CD_t) ~ log10(CD_t-k) + log10(P_ext,t-k) + P_int,t-k + error

```{r}
regkb_CD<- glm(formula=log10(CD_cellsmL_interp) ~ #CD_t
              log10(CD_cellsmL_interp_k7) + #CD_t-k
                log10(ExtP_kgday_interp_k28) + #P_ext,t-k
                IntP_kgday_interp_k28, #P_int,t-k
              data=Me_all_interp) #data frame
summary(regkb_CD)#prints the glm output
aov(regkb_CD)#prints the sum of squares for the regression and residuals
png(file="Products/Figure21.png", width=750, height=500) #open png file
layout(matrix(c(1,2,3,4),2,2)) #layout 4 graphs/page
fig21<-plot(regkb_CD) #creates plot
dev.off() #closes the file
par(mfrow=c(1,1))
```

f) Running third model for CD...
Model: CD_t ~ CD_t-7 + P_ext,t-k + P_hypo:P_epiP,t-k + error

```{r}
regkc_CD<- glm(formula=log10(CD_cellsmL_interp) ~ #CD_t
              log10(CD_cellsmL_interp_k7) + #CD_t-k
                log10(ExtP_kgday_interp_k28) + #P_ext,t-k
                log10(HE_ratio_interp_k28), #P_hypo:P_epiP,t-k
              data=Me_all_interp) #data frame
summary(regkc_CD)#prints the glm output
aov(regkc_CD)#prints the sum of squares for the regression and residuals
png(file="Products/Figure22.png", width=750, height=500) #open png file
layout(matrix(c(1,2,3,4),2,2)) #layout 4 graphs/page
fig22<-plot(regkc_CD) #creates plot
dev.off() #closes the file
par(mfrow=c(1,1))
```

Pick the model with the lowest AIC, Residual standard error, and each term is significant (t-score p-value < 0.05) is the best model.

# Step 4. Using the best model, calculate the variance for each term for each 
year:

a) calculating source-attributed impact for Chla...
```{r}
YRregk_Chla_2013<- aov(formula=log10(Chla_ugL_interp) ~ #Chla_t
              log10(Chla_ugL_interp_k7) + #Chla_t-k
                log10(ExtP_kgday_interp_k28) + #P_ext,t-k
                IntP_kgday_interp_k28, #P_int,t-k
              data=subset(Me_all_interp, Yr==2013)) #subset data by year
Chla_SS_2013<- summary(YRregk_Chla_2013)#prints the sum of squares for the regression 
Chla_impact2013<- data.frame(Chla_SS_2013[[1]])[,2,drop=FALSE] #pulls out the SS as table

YRregk_Chla_2014<- aov(formula=log10(Chla_ugL_interp) ~ #Chla_t
              log10(Chla_ugL_interp_k7) + #Chla_t-k
                log10(ExtP_kgday_interp_k28) + #P_ext,t-k
                IntP_kgday_interp_k28, #P_int,t-k
              data=subset(Me_all_interp, Yr==2014)) #subset data by year
Chla_SS_2014<- summary(YRregk_Chla_2014)#prints the sum of squares for the regression 
Chla_impact2014<- data.frame(Chla_SS_2014[[1]])[,2,drop=FALSE] #pulls out the SS as table

YRregk_Chla_2015<- aov(formula=log10(Chla_ugL_interp) ~ #Chla_t
              log10(Chla_ugL_interp_k7) + #Chla_t-k
                log10(ExtP_kgday_interp_k28) + #P_ext,t-k
                IntP_kgday_interp_k28, #P_int,t-k
              data=subset(Me_all_interp, Yr==2015)) #subset data by year
Chla_SS_2015<- summary(YRregk_Chla_2015)#prints the sum of squares for the regression 
Chla_impact2015<- data.frame(Chla_SS_2015[[1]])[,2,drop=FALSE] #pulls out the SS as table

YRregk_Chla_2016<- aov(formula=log10(Chla_ugL_interp) ~ #Chla_t
              log10(Chla_ugL_interp_k7) + #Chla_t-k
                log10(ExtP_kgday_interp_k28) + #P_ext,t-k
                IntP_kgday_interp_k28, #P_int,t-k
              data=subset(Me_all_interp, Yr==2016)) #subset data by year
Chla_SS_2016<- summary(YRregk_Chla_2016)#prints the sum of squares for the regression 
Chla_impact2016<- data.frame(Chla_SS_2016[[1]])[,2,drop=FALSE] #pulls out the SS as table

YRregk_Chla_2017<- aov(formula=log10(Chla_ugL_interp) ~ #Chla_t
              log10(Chla_ugL_interp_k7) + #Chla_t-k
                log10(ExtP_kgday_interp_k28) + #P_ext,t-k
                IntP_kgday_interp_k28, #P_int,t-k
              data=subset(Me_all_interp, Yr==2017)) #subset data by year
Chla_SS_2017<- summary(YRregk_Chla_2017)#prints the sum of squares for the regression 
Chla_impact2017<- data.frame(Chla_SS_2017[[1]])[,2,drop=FALSE] #pulls out the SS as table

YRregk_Chla_2018<- aov(formula=log10(Chla_ugL_interp) ~ #Chla_t
              log10(Chla_ugL_interp_k7) + #Chla_t-k
                log10(ExtP_kgday_interp_k28) + #P_ext,t-k
                IntP_kgday_interp_k28, #P_int,t-k
              data=subset(Me_all_interp, Yr==2018)) #subset data by year
Chla_SS_2018<- summary(YRregk_Chla_2018)#prints the sum of squares for the regression 
Chla_impact2018<- data.frame(Chla_SS_2018[[1]])[,2,drop=FALSE] #pulls out the SS as table

Model_output_Chla<- cbind(Chla_impact2013, Chla_impact2014, Chla_impact2015, 
                          Chla_impact2016, Chla_impact2017, Chla_impact2018) #combines all tables together
Model_output_Chla$Sum_Sq_part<- c("Seed stock", "External load", "Internal load", "Residual") #add column with predictor name
colnames(Model_output_Chla)<- c("2013", "2014", "2015", "2016", "2017", "2018", "Sum_Sq_part")
# renames columns with year for sum squares
write.csv(Model_output_Chla, file="Cleaned_data/Impact_model_output_Chla.csv")
write.csv(Model_output_Chla, file="Products/Figure5_data.csv")
```

Yearly source-attributed impact outputted.

b) calculating source-attributed impact for CD...

```{r}
YRregk_CD_2013<- aov(formula=log10(CD_cellsmL_interp) ~ #CD_t
              log10(CD_cellsmL_interp_k7) + #CD_t-k
                log10(ExtP_kgday_interp_k28) + #P_ext,t-k
                IntP_kgday_interp_k28, #P_int,t-k
              data=subset(Me_all_interp, Yr==2013) #subset data by year
            )
SS_2013<- summary(YRregk_CD_2013)#prints the sum of squares for the regression 
impact2013<- data.frame(SS_2013[[1]])[,2,drop=FALSE] #pulls out the SS as table

YRregk_CD_2014<- aov(formula=log10(CD_cellsmL_interp) ~ #CD_t
              log10(CD_cellsmL_interp_k7) + #CD_t-k
                log10(ExtP_kgday_interp_k28) + #P_ext,t-k
                IntP_kgday_interp_k28, #P_int,t-k
              data=subset(Me_all_interp, Yr==2014) #subset data by year
            )
SS_2014<- summary(YRregk_CD_2014)#prints the sum of squares for the regression 
impact2014<- data.frame(SS_2014[[1]])[,2,drop=FALSE] #pulls out the SS as table

YRregk_CD_2015<- aov(formula=log10(CD_cellsmL_interp) ~ #CD_t
              log10(CD_cellsmL_interp_k7) + #CD_t-k
                log10(ExtP_kgday_interp_k28) + #P_ext,t-k
                IntP_kgday_interp_k28, #P_int,t-k
              data=subset(Me_all_interp, Yr==2015) #subset data by year
            )
SS_2015<- summary(YRregk_CD_2015)#prints the sum of squares for the regression 
impact2015<- data.frame(SS_2015[[1]])[,2,drop=FALSE] #pulls out the SS as table

YRregk_CD_2016<- aov(formula=log10(CD_cellsmL_interp) ~ #CD_t
              log10(CD_cellsmL_interp_k7) + #CD_t-k
                log10(ExtP_kgday_interp_k28) + #P_ext,t-k
                IntP_kgday_interp_k28, #P_int,t-k
              data=subset(Me_all_interp, Yr==2016) #subset data by year
            )
SS_2016<- summary(YRregk_CD_2016)#prints the sum of squares for the regression 
impact2016<- data.frame(SS_2016[[1]])[,2,drop=FALSE] #pulls out the SS as table

YRregk_CD_2017<- aov(formula=log10(CD_cellsmL_interp) ~ #CD_t
              log10(CD_cellsmL_interp_k7) + #CD_t-k
                log10(ExtP_kgday_interp_k28) + #P_ext,t-k
                IntP_kgday_interp_k28, #P_int,t-k
              data=subset(Me_all_interp, Yr==2017) #subset data by year
            )
SS_2017<- summary(YRregk_CD_2017)#prints the sum of squares for the regression 
impact2017<- data.frame(SS_2017[[1]])[,2,drop=FALSE] #pulls out the SS as table

YRregk_CD_2018<- aov(formula=log10(CD_cellsmL_interp) ~ #CD_t
              log10(CD_cellsmL_interp_k7) + #CD_t-k
                log10(ExtP_kgday_interp_k28) + #P_ext,t-k
                IntP_kgday_interp_k28, #P_int,t-k
              data=subset(Me_all_interp, Yr==2018) #subset data by year
            )
SS_2018<- summary(YRregk_CD_2018)#prints the sum of squares for the regression 
impact2018<- data.frame(SS_2018[[1]])[,2,drop=FALSE] #pulls out the SS as table

Model_output_CD<- cbind(impact2013, impact2014, impact2015, impact2016, 
                          impact2017, impact2018) #combines all tables together
Model_output_CD$Sum_Sq_part<- c("Seed stock", "External load", "Internal load", "Residual") #add column with predictor name
colnames(Model_output_CD)<- c("2013", "2014", "2015", "2016", "2017", "2018", "Sum_Sq_part")
# renames columns with year for sum squares
write.csv(Model_output_CD, file="Cleaned_data/Impact_model_output_CD.csv")
write.csv(Model_output_CD, file="Products/Figure6_data.csv")
```

Yearly source-attributed impact outputted.
