---
title: "TotalP_external_inflows.Rmd"
author: "Lauren A. Knose, ORISE-EPA"
date: '2023-03-06'
output:
  html_document: default
  pdf_document: default
editor_options: 
  chunk_output_type: console
---

The purpose of this program is to calculate the external total P loading
to Lake Mendota, from the three main inflows of Pheasant Branch (PB), Yahara 
River (YR), and Six-Mile Creek (SMC). 
P_inflow = P concentration * flow rate


# Step 1. Load the data and packages needed:

a) Loading the packages needed...

```{r}
library(dplyr) #needed for transforming data
library(reshape2) #needed for transforming data wide to long
library(lubridate) #needed for date conversion functions
library(ggplot2) #needed for plotting
library(ggbreak) #needed for changing scales in graphs
```

Packages loaded.


b) Loading the data needed...

```{r}
PB_TotP_flow<- read.csv(file="Cleaned_data/PheasantBranch_TotP.csv")
YR_TotP_flow<- read.csv(file="Cleaned_data/YaharaRiver_TotP.csv")
SMC_TotP_flow<- read.csv(file="Cleaned_data/SixMileCreek_TotP.csv")
```

Total P data loaded.

The site ID and source for Pheasant Branch is `r unique(PB_TotP_flow$Site_ID)`.
The site ID and source for Yahara River is `r unique(YR_TotP_flow$Site_ID)`.
The site ID and source for Six-mile Creek is `r unique(SMC_TotP_flow$Site_ID)`.
Note, there are two sites, one with historic data and one with current data, for
Six-mile Creek.

# Step 2. Transform and filter the data:

a) defining conversion coefficients...

```{r}
lb_kg= 0.453592/1 #conversion factor from lb to kg (1lb = 0.453592 kg)
cf_L= 1/0.0353147 #conversion factor from ft^3 to L (1 ft^2 = 28.3168 L)
sec_day= 86400/1 #conversion factor from sec to day (86400 sec = 1 day)
mg_kg= 1/1000000 #conversion factor from mg to kg (1,000,000 mg = 1 kg)
```

Conversion coefficients defined.

b) transforming and filtering the data...

```{r}
PB_TotP_flow<- PB_TotP_flow %>%
  mutate(sampledate=as.Date(sampledate, origin="1899-12-30", format="%m/%d/%Y"),
         Q_cfs=Qout_cfs, #rename column for consistency across data frames
         Qout_cfs=NULL, #remove excess column
         TotP_kgday=(Q_cfs*TotP_mgL*cf_L*sec_day*mg_kg)
         ) %>%
  filter(Q_cfs>=0) #remove any records where there is a negative flow rate
YR_TotP_flow<- YR_TotP_flow %>%
  mutate(sampledate=as.Date(sampledate, origin="1899-12-30", format="%m/%d/%Y"),
         TotP_kgday=(Q_cfs*TotP_mgL*cf_L*sec_day*mg_kg)
         )%>%
  filter(Q_cfs>=0) #remove any records where there is a negative flow rate
SMC_TotP_flow<- SMC_TotP_flow %>%
  mutate(sampledate=as.Date(sampledate, origin="1899-12-30", format="%m/%d/%Y"),
         TotP_kgday=(Q_cfs*TotP_mgL*cf_L*sec_day*mg_kg)
         )%>%
  filter(Q_cfs>=0) #remove any records where there is a negative flow rate
```

Total P data transformed and filtered for all inflows.

c) viewing data available...

```{r}
layout(matrix(c(1,2,3),3,1)) # optional 3 graphs/page
plot(PB_TotP_flow$sampledate, PB_TotP_flow$TotP_kgday, 
     xlab = "Sample Date", ylab = "Total P Loading (kg/day)", 
     main = "Pheasant Branch", cex.lab=1.5, cex.main=2, cex.axis=2)
plot(YR_TotP_flow$sampledate, YR_TotP_flow$TotP_kgday, 
     xlab = "Sample Date", ylab = "Total P Loading (kg/day)", 
     main = "Yahara River", cex.lab=1.5, cex.main=2, cex.axis=2)
plot(SMC_TotP_flow$sampledate, SMC_TotP_flow$TotP_kgday, 
     xlab = "Sample Date", ylab = "Total P Loading (kg/day)", 
     main = "Six-Mile Creek", cex.lab=1.5, cex.main=2, cex.axis=2)
par(mfrow=c(1,1))#returns graphing viewer to 1 plot per page
```

Total P data plotted. Total P available for all three inflows from 2012 to 2022.

# Step 3. Interpolate data between observations to get daily loads: 

a) interpolating data for PB inflow using linear regression...

```{r}
PB_date_range<- seq( #creates a vector of sequence data
  min(PB_TotP_flow$sampledate), #starting at the first sample date
  max(PB_TotP_flow$sampledate), #ending at the last sample date
  by="1 day") #observations every 1 day
PB_interpolated<- approx( #creates a vector of data that linearly interpolates 
  PB_TotP_flow$sampledate, #across date
  PB_TotP_flow$TotP_kgday, #the predicted values of total P
  xout=PB_date_range) #for every 1 day
PB_TotP_flow_interp <- data.frame( #creates a new data frame that has 
  sampledate = as.Date(PB_interpolated$x), #measurements every 1 day
                             PB_TotP_kgday=PB_interpolated$y) #predicted total P 
```

Total P loading interpreted for Pheasant Branch.

b) interpolating data for YR inflow using linear regression...

```{r}
YR_date_range<- seq( #creates a vector of sequence data
  min(YR_TotP_flow$sampledate), #starting at the first sample date
  max(YR_TotP_flow$sampledate), #ending at the last sample date
  by="1 day") #observations every 14 days (bimonthly) 
YR_interpolated<- approx( #creates a vector of data that linearly interpolates 
  YR_TotP_flow$sampledate, #across date
  YR_TotP_flow$TotP_kgday, #the predicted values of Hypo P: Epi P
  xout=YR_date_range) #for every 1 day
YR_TotP_flow_interp <- data.frame( #creates a new data frame that has 
  sampledate = as.Date(YR_interpolated$x), #measurements every 14 days
                             YR_TotP_kgday=YR_interpolated$y) #predicted hypo: epi P 
```

Total P loading interpolated for Yahara River.

c) interpolating data for SMC inflow using linear regression...

```{r}
SMC_date_range<- seq( #creates a vector of sequence data
  min(SMC_TotP_flow$sampledate), #starting at the first sample date
  max(SMC_TotP_flow$sampledate), #ending at the last sample date
  by="1 day") #observations every 14 days (bimonthly) 
SMC_interpolated<- approx( #creates a vector of data that linearly interpolates 
  SMC_TotP_flow$sampledate, #across date
  SMC_TotP_flow$TotP_kgday, #the predicted values of Hypo P: Epi P
  xout=SMC_date_range) #for every 1 day
SMC_TotP_flow_interp <- data.frame( #creates a new data frame that has 
  sampledate = as.Date(SMC_interpolated$x), #measurements every 14 days
                             SMC_TotP_kgday=SMC_interpolated$y) #predicted hypo: epi P 
```

Total P loading interpolated for Six-mile Creek.

# Step 4. Merge the data:
Note, R only merges two data frames at a time.

```{r}
Inflows<- merge(PB_TotP_flow_interp, YR_TotP_flow_interp, #merge the two files
                by="sampledate") #keep records with same sample date
Inflows<- merge(Inflows, SMC_TotP_flow_interp, #merge two data frames
                by="sampledate") #keep records with same sample date
Inflows<- Inflows %>%
  mutate(allInflows_TotP_kgday=(PB_TotP_kgday+YR_TotP_kgday+SMC_TotP_kgday), #sum
         Yr= as.integer(format(sampledate, "%Y")), #add year field
         daynum=as.integer(format(sampledate, "%j")) #add day number field
         )
```

Total P data files merged into one file, "Inflows."

# Step 5. View and summarize the data:

a) viewing daily total P loading for each inflow...

```{r}
Inflows_long<- melt(Inflows, #reshape data frame from wide to long
               id.vars=c("sampledate"), #the column to use as unique id
               measure.vars=c("PB_TotP_kgday","YR_TotP_kgday","SMC_TotP_kgday",
                              "allInflows_TotP_kgday"), 
               #the columns where values will be taken
               variable.name="Inflow", #assigns variable field name
               value.name="TotP_kgday" #assigns value field name
               )
Inflows_long$Inflow<-as.character(Inflows_long$Inflow) #tell R character data held
Inflows_long$Inflow[Inflows_long$Inflow=="PB_TotP_kgday"] <- "Pheasant Branch"
Inflows_long$Inflow[Inflows_long$Inflow=="YR_TotP_kgday"] <- "Yahara River"
Inflows_long$Inflow[Inflows_long$Inflow=="SMC_TotP_kgday"] <- "Six-mile Creek"
Inflows_long$Inflow[Inflows_long$Inflow=="allInflows_TotP_kgday"] <- "All inflows"

Inflows_long<- Inflows_long %>%
  filter(!is.na(TotP_kgday)) %>% #remove records that have NA
  mutate(daynum=as.integer(format(sampledate, format="%j")), #add column with 
         #daynum for filtering
         Yr=as.integer(format(sampledate, format="%Y"))) %>% #add column with Yr
  filter(Yr>= 2012 & Yr <=2018)

write.csv(Inflows_long, file="Products/Figure9_data.csv")
fig9<- ggplot(data=Inflows_long, 
       aes(x=daynum, 
           y=TotP_kgday)) +
  geom_line()+
  facet_wrap(~Inflow) +
  theme_classic(base_size=12)+
  labs(y="Daily Stream Total P loading (kg/day)",
       x="Day of Year",
      caption="White highlight indicates CyanoHAB season, mid-July to mid-Sept.") +
  scale_y_continuous(trans="sqrt") + #transform y-axis scale to see small values
  annotate('rect', fill='gray60', alpha=.2, #creates a shadow box 
           xmin=-Inf, xmax=195, ymin=0, ymax=Inf) +#shadow box highlights CyanoHAB season
  annotate('rect', fill='gray60', alpha=.2, #creates a shadow box 
           xmin=255, xmax=Inf, ymin=0, ymax=Inf)  #shadow box highlights CyanoHAB season
ggsave(fig9, file="Products/Figure9.png")
```

Plotted Total P for day of year across the three inflows, for all years. 

b) calculating average daily total P loading for each inflow during the CyanoHAB 
season...

```{r}
Inflows_avg<- Inflows_long %>%
  filter(daynum>= 195 & daynum <=255) %>% #filter for CyanoHAB season
  group_by(Inflow) %>% #compute for each inflow
  summarise(mean=mean(TotP_kgday, #average daily total P loading
                      na.rm=TRUE),
            se=(sd(TotP_kgday, na.rm=TRUE)/sqrt(length(TotP_kgday))),
            min=min(TotP_kgday, na.rm=TRUE),
            max=max(TotP_kgday, na.rm=TRUE))
Inflows_avg
```

Estimated average daily total P loading during CyanoHAB season for each inflow.

c) calculating annual total P load during CyanoHAB season for each year...

```{r}
Inflows_sumavg<- Inflows %>%
  filter(daynum>= 195 & daynum <=255) %>% #filter for CyanoHAB season
  group_by(Yr) %>% #group all records by year
  summarise(sum_TP_kgyr=sum(allInflows_TotP_kgday))  #calculate the total P load
Inflows_sumavg
```

The average annual total P load, during the CyanoHAB season, was
`r mean(Inflows_sumavg$sum_TP_kgyr)` +- `r sd(Inflows_sumavg$sum_TP_kgyr)/sqrt(length(Inflows_sumavg$sum_TP_kgyr))` kg/yr. 

d) calculating annual total P load for year...

```{r}
Inflows_yravg<- Inflows %>%
  group_by(Yr) %>% #group all records by year
  summarise(sum_TP_kgyr=sum(allInflows_TotP_kgday))  #calculate the total P load
Inflows_yravg
```

The average annual total P load, for the year, was
`r mean(Inflows_yravg$sum_TP_kgyr)` +- `r sd(Inflows_yravg$sum_TP_kgyr)/sqrt(length(Inflows_yravg$sum_TP_kgyr))` kg/yr. 

# Step 6. Save new data file.

```{R}
write.csv(Inflows, file="Cleaned_data/TotalP_inflows.csv") #save as new data file
```

Data file saved as "TotalP_inflows.csv" in the Cleaned_data folder.